---
title: "Lineage Area Review"
output:
  html_document:
    toc: true
    toc_float: true
params:
  lineage_id: null
  analysis_dir: null
  all_results_dir: "data/lineage_area"
  manifest_csv: "manifests/image_manifest.csv"
  show_images: true
  max_images: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5
)
```

## Overview

This notebook reviews the saved area analyses across all lineages and highlights the most promising long-term distribution changes.

The intended use is:

1. Rank lineages by how strongly their adjusted area distributions change across passages.
2. Pick one promising lineage and transition.
3. Inspect the raw image panel around that transition to check whether the statistical signal looks real.

## Packages

```{r packages}
required_packages <- c(
  "dplyr",
  "ggplot2",
  "purrr",
  "readr",
  "tibble",
  "tidyr",
  "stringr",
  "scales"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    sprintf(
      "Install required packages before rendering: %s",
      paste(missing_packages, collapse = ", ")
    ),
    call. = FALSE
  )
}

library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(tibble)
library(tidyr)
library(stringr)
library(scales)
```

## Helpers

```{r helpers}
resolve_project_root <- function() {
  candidates <- c(
    normalizePath(".", winslash = "/", mustWork = FALSE),
    normalizePath("..", winslash = "/", mustWork = FALSE),
    normalizePath("../..", winslash = "/", mustWork = FALSE)
  )

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "README.md")) &&
        dir.exists(file.path(candidate, "workflows"))) {
      return(candidate)
    }
  }

  stop("Could not locate project root from current working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

project_path <- function(path) {
  is_absolute <- grepl("^[A-Za-z]:[/\\\\]|^/", path)
  out <- path
  out[!is_absolute] <- file.path(project_root, path[!is_absolute])
  out
}

analysis_dir_from_id <- function(lineage_id) {
  file.path(project_path(params$all_results_dir), paste0("lineage_", lineage_id))
}

parse_datetime_safe <- function(x) {
  if (inherits(x, "POSIXct")) return(x)
  x <- as.character(x)
  out <- as.POSIXct(rep(NA_character_, length(x)), tz = "UTC")
  formats <- c(
    "%Y-%m-%d %H:%M:%S",
    "%Y-%m-%dT%H:%M:%SZ",
    "%Y-%m-%d %H:%M:%OS",
    "%m/%d/%Y %H:%M",
    "%m/%d/%Y %H:%M:%S"
  )
  for (fmt in formats) {
    missing <- is.na(out) & !is.na(x) & nzchar(x)
    if (!any(missing)) next
    out[missing] <- as.POSIXct(x[missing], format = fmt, tz = "UTC")
  }
  out
}

safe_quantile <- function(x, prob) {
  if (!length(x) || all(is.na(x))) return(NA_real_)
  as.numeric(stats::quantile(x, probs = prob, na.rm = TRUE, names = FALSE))
}

rescale_01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || diff(rng) <= 0) return(rep(0, length(x)))
  (x - rng[[1]]) / diff(rng)
}

normalize_for_display <- function(img) {
  img <- as.array(img)
  if (length(dim(img)) == 3) {
    img <- apply(img[, , seq_len(min(3, dim(img)[3])), drop = FALSE], c(1, 2), mean)
  }
  q <- stats::quantile(img, probs = c(0.01, 0.99), na.rm = TRUE, names = FALSE)
  img <- pmin(pmax(img, q[[1]]), q[[2]])
  img <- img - min(img, na.rm = TRUE)
  mx <- max(img, na.rm = TRUE)
  if (is.finite(mx) && mx > 0) {
    img <- img / mx
  }
  img
}

plot_grayscale_matrix <- function(img, main = "") {
  raster_img <- as.raster(img[nrow(img):1, ], max = 1)
  plot.new()
  usr <- par("usr")
  rasterImage(raster_img, usr[1], usr[3], usr[2], usr[4], interpolate = FALSE)
  title(main = main, cex.main = 0.8)
}

read_analysis_bundle <- function(path) {
  obj <- readRDS(path)
  list(
    lineage_id = as.character(obj$lineage_id),
    cells = tibble::as_tibble(obj$cells),
    image_summary = tibble::as_tibble(obj$image_summary),
    passage_summary = tibble::as_tibble(obj$passage_summary)
  )
}

smooth_series <- function(x) {
  x <- as.numeric(x)
  n <- length(x)
  if (n <= 2) return(x)
  k <- min(5, if (n %% 2 == 1) n else n - 1)
  if (k < 3) k <- 3
  k <- min(k, if (n %% 2 == 1) n else n - 1)
  if (k < 3) return(x)
  stats::runmed(x, k = k, endrule = "median")
}

calc_monotonicity <- function(x) {
  net <- abs(x[[length(x)]] - x[[1]])
  total <- sum(abs(diff(x)))
  if (!is.finite(total) || total <= 0) return(0)
  net / total
}

lineage_trend_summary <- function(df) {
  df <- df %>% arrange(passage_position)
  n_passages <- nrow(df)
  start_passage_id <- df$passage_id[[1]]
  end_passage_id <- df$passage_id[[n_passages]]
  matched_passage_fraction <- if ("n_matched_images" %in% names(df)) {
    mean(df$n_matched_images > 0, na.rm = TRUE)
  } else {
    NA_real_
  }

  score_df <- df %>%
    filter(
      is.finite(median_z) |
      is.finite(iqr_z) |
      is.finite(q99_z) |
      is.finite(frac_z_gt_3)
    ) %>%
    arrange(passage_position)

  if (nrow(score_df) < 3) {
    return(tibble(
      n_passages = n_passages,
      n_scored_passages = nrow(score_df),
      matched_passage_fraction = matched_passage_fraction,
      start_passage_id = start_passage_id,
      end_passage_id = end_passage_id,
      smooth_delta_median_z = NA_real_,
      smooth_delta_iqr_z = NA_real_,
      smooth_delta_q99_z = NA_real_,
      smooth_delta_frac_z_gt_3 = NA_real_,
      max_adjacent_jump_q99_z = NA_real_,
      max_adjacent_jump_frac_z_gt_3 = NA_real_,
      monotonicity_median = NA_real_,
      monotonicity_q99 = NA_real_,
      monotonicity_frac_z_gt_3 = NA_real_,
      long_term_score = NA_real_,
      trend_class = "Insufficient data",
      trend_direction = "undetermined",
      trend_statement = "Fewer than three passages had enough analyzed objects to assess a longitudinal trend."
    ))
  }

  med_s <- smooth_series(score_df$median_z)
  iqr_s <- smooth_series(score_df$iqr_z)
  q99_s <- smooth_series(score_df$q99_z)
  frac_s <- smooth_series(score_df$frac_z_gt_3)

  delta_med <- med_s[[length(med_s)]] - med_s[[1]]
  delta_iqr <- iqr_s[[length(iqr_s)]] - iqr_s[[1]]
  delta_q99 <- q99_s[[length(q99_s)]] - q99_s[[1]]
  delta_frac <- frac_s[[length(frac_s)]] - frac_s[[1]]

  monotonicity_median <- calc_monotonicity(med_s)
  monotonicity_q99 <- calc_monotonicity(q99_s)
  monotonicity_frac <- calc_monotonicity(frac_s)

  long_term_score_raw <-
    abs(delta_med) +
    0.5 * abs(delta_iqr) +
    1.5 * abs(delta_q99) +
    8 * abs(delta_frac)

  consistency_weight <- mean(c(monotonicity_median, monotonicity_q99, monotonicity_frac), na.rm = TRUE)
  long_term_score <- long_term_score_raw * consistency_weight * log1p(n_passages)

  primary_signal <- c(
    center = abs(delta_med),
    spread = 0.75 * abs(delta_iqr),
    tail = max(abs(delta_q99), 4 * abs(delta_frac))
  )
  primary_axis <- names(primary_signal)[which.max(primary_signal)]
  signed_tail <- delta_q99 + 4 * delta_frac
  signed_center <- delta_med
  signed_spread <- delta_iqr

  if (!is.finite(long_term_score) || long_term_score < 0.75 || consistency_weight < 0.2) {
    trend_class <- "No clear long-term trend"
    trend_direction <- "none"
    trend_statement <- "No strong, persistent long-term drift is evident once passage-level noise is accounted for."
  } else if (primary_axis == "tail") {
    trend_direction <- if (signed_tail >= 0) "larger-tail" else "smaller-tail"
    trend_class <- if (signed_tail >= 0) "Upper-tail expansion" else "Upper-tail contraction"
    trend_statement <- sprintf(
      "%s: the upper tail changes more than the lineage center, suggesting a persistent %s of unusually large cells.",
      trend_class,
      if (signed_tail >= 0) "enrichment" else "loss"
    )
  } else if (primary_axis == "spread") {
    trend_direction <- if (signed_spread >= 0) "broader" else "tighter"
    trend_class <- if (signed_spread >= 0) "Distribution broadening" else "Distribution tightening"
    trend_statement <- sprintf(
      "%s: dispersion changes dominate over center shift, indicating the lineage is becoming %s over time.",
      trend_class,
      if (signed_spread >= 0) "more heterogeneous" else "more homogeneous"
    )
  } else {
    trend_direction <- if (signed_center >= 0) "larger" else "smaller"
    trend_class <- if (signed_center >= 0) "Bulk shift upward" else "Bulk shift downward"
    trend_statement <- sprintf(
      "%s: the lineage center drifts %s over time, consistent with a broad change in typical cell size rather than only a tail effect.",
      trend_class,
      if (signed_center >= 0) "toward larger residuals" else "toward smaller residuals"
    )
  }

  tibble(
    n_passages = n_passages,
    n_scored_passages = nrow(score_df),
    matched_passage_fraction = matched_passage_fraction,
    start_passage_id = start_passage_id,
    end_passage_id = end_passage_id,
    smooth_delta_median_z = delta_med,
    smooth_delta_iqr_z = delta_iqr,
    smooth_delta_q99_z = delta_q99,
    smooth_delta_frac_z_gt_3 = delta_frac,
    max_adjacent_jump_q99_z = max(abs(diff(score_df$q99_z)), na.rm = TRUE),
    max_adjacent_jump_frac_z_gt_3 = max(abs(diff(score_df$frac_z_gt_3)), na.rm = TRUE),
    monotonicity_median = monotonicity_median,
    monotonicity_q99 = monotonicity_q99,
    monotonicity_frac_z_gt_3 = monotonicity_frac,
    long_term_score = long_term_score,
    trend_class = trend_class,
    trend_direction = trend_direction,
    trend_statement = trend_statement
  )
}

choose_representative_passages <- function(df) {
  score_df <- df %>%
    filter(
      is.finite(median_z) |
      is.finite(iqr_z) |
      is.finite(q99_z) |
      is.finite(frac_z_gt_3)
    ) %>%
    arrange(passage_position)

  if (!nrow(score_df)) {
    return(character())
  }

  idx <- unique(c(1, ceiling(nrow(score_df) / 2), nrow(score_df)))
  score_df$passage_id[idx]
}
```

## Discover Analysis Outputs

```{r discover-results}
all_results_dir <- project_path(params$all_results_dir)
analysis_paths <- list.files(
  all_results_dir,
  pattern = "lineage_area_analysis\\.rds$",
  recursive = TRUE,
  full.names = TRUE
)

if (!length(analysis_paths)) {
  stop(sprintf("No lineage analysis outputs found under %s", all_results_dir), call. = FALSE)
}

analysis_index <- tibble(
  analysis_rds = analysis_paths,
  analysis_dir = dirname(analysis_paths),
  lineage_id = sub("^lineage_", "", basename(dirname(analysis_paths)))
)

knitr::kable(
  tibble(
    n_lineages_found = nrow(analysis_index),
    results_dir = all_results_dir
  )
)
```

## Load Summary Metrics Across All Lineages

```{r load-all-summaries}
summary_bundle <- purrr::map(analysis_index$analysis_rds, read_analysis_bundle)

all_passage_tbl <- purrr::map2_dfr(summary_bundle, analysis_index$analysis_dir, function(bundle, analysis_dir) {
  bundle$passage_summary %>%
    mutate(
      lineage_id = bundle$lineage_id,
      analysis_dir = analysis_dir,
      passage_position = as.integer(passage_position),
      first_image_time = parse_datetime_safe(first_image_time)
    )
})

all_image_tbl <- purrr::map2_dfr(summary_bundle, analysis_index$analysis_dir, function(bundle, analysis_dir) {
  bundle$image_summary %>%
    mutate(
      lineage_id = bundle$lineage_id,
      analysis_dir = analysis_dir,
      passage_position = as.integer(passage_position)
    )
})
```

## Per-Lineage Trend Summary

```{r lineage-summary}
lineage_rank_tbl <- all_passage_tbl %>%
  arrange(lineage_id, passage_position) %>%
  group_by(lineage_id) %>%
  group_modify(~ lineage_trend_summary(.x)) %>%
  ungroup()

top_lineage_tbl <- lineage_rank_tbl %>%
  select(
    lineage_id,
    n_passages,
    n_scored_passages,
    matched_passage_fraction,
    start_passage_id,
    end_passage_id,
    trend_class,
    long_term_score,
    smooth_delta_median_z,
    smooth_delta_iqr_z,
    smooth_delta_q99_z,
    smooth_delta_frac_z_gt_3,
    monotonicity_median,
    monotonicity_q99,
    monotonicity_frac_z_gt_3
  )

knitr::kable(top_lineage_tbl, digits = 3)
```

Interpretation:

- `trend_class` is the notebook's best high-level description of the long-term behavior in that lineage.
- `smooth_delta_*` values measure start-to-end change after median smoothing across scored passages.
- `matched_passage_fraction` shows how much of the manual lineage currently has embedding-derived area data.

## Cross-Lineage Overview

```{r lineage-overview-plot}
overview_long_tbl <- all_passage_tbl %>%
  select(lineage_id, passage_position, median_z, iqr_z, q99_z, frac_z_gt_3) %>%
  pivot_longer(
    cols = c(median_z, iqr_z, q99_z, frac_z_gt_3),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(lineage_id, metric) %>%
  arrange(passage_position, .by_group = TRUE) %>%
  mutate(smooth_value = smooth_series(value)) %>%
  ungroup()

ggplot(overview_long_tbl, aes(x = passage_position, y = value)) +
  geom_line(color = "grey75", linewidth = 0.5) +
  geom_point(color = "grey55", size = 1.3) +
  geom_line(aes(y = smooth_value, color = lineage_id), linewidth = 1) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  labs(
    title = "Passage-level trajectories across the three manual lineages",
    subtitle = "Grey points show raw passage summaries; colored lines show median-smoothed trajectories",
    x = "Passage position",
    y = "Summary value",
    color = "Lineage"
  ) +
  theme_minimal(base_size = 12)
```

```{r lineage-coverage-plot}
coverage_long_tbl <- all_passage_tbl %>%
  select(lineage_id, passage_position, n_manifest_images, n_matched_images, n_objects) %>%
  pivot_longer(
    cols = c(n_manifest_images, n_matched_images, n_objects),
    names_to = "metric",
    values_to = "value"
  )

ggplot(coverage_long_tbl, aes(x = passage_position, y = value, color = metric)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  facet_wrap(~ lineage_id, scales = "free_y", ncol = 1) +
  labs(
    title = "Data availability along each manual lineage",
    subtitle = "Use this to distinguish a real flattening of the signal from missing segmentation coverage",
    x = "Passage position",
    y = "Count",
    color = "Metric"
  ) +
  theme_minimal(base_size = 12)
```

## Detailed Lineage Review

```{r detailed-lineage-review, results='asis', fig.height=8, fig.width=10}
bundle_by_id <- stats::setNames(summary_bundle, analysis_index$lineage_id)

for (lineage_id in analysis_index$lineage_id) {
  current_lineage_id <- lineage_id
  cat("### ", current_lineage_id, "\n\n", sep = "")

  bundle <- bundle_by_id[[current_lineage_id]]
  passage_tbl <- bundle$passage_summary %>%
    mutate(
      passage_position = as.integer(passage_position),
      first_image_time = parse_datetime_safe(first_image_time)
    ) %>%
    arrange(passage_position)
  cell_tbl <- bundle$cells %>%
    mutate(
      passage_position = as.integer(passage_position),
      p_real = as.numeric(p_real),
      z = as.numeric(z)
    )
  summary_row <- lineage_rank_tbl %>% filter(lineage_id == .env$current_lineage_id) %>% slice_head(n = 1)

  cat(summary_row$trend_statement[[1]], "\n\n")

  detail_tbl <- summary_row %>%
    transmute(
      lineage_id,
      trend_class,
      n_passages,
      n_scored_passages,
      matched_passage_fraction,
      long_term_score,
      smooth_delta_median_z,
      smooth_delta_iqr_z,
      smooth_delta_q99_z,
      smooth_delta_frac_z_gt_3,
      monotonicity_median,
      monotonicity_q99,
      monotonicity_frac_z_gt_3
    )
  print(knitr::kable(detail_tbl, digits = 3))
  cat("\n")

  display_tbl <- passage_tbl %>%
    transmute(
      passage_position,
      passage_id,
      n_manifest_images,
      n_matched_images,
      n_objects,
      effective_n = round(effective_n, 1),
      nuisance_fraction = round(nuisance_fraction, 3),
      median_z = round(median_z, 3),
      iqr_z = round(iqr_z, 3),
      q99_z = round(q99_z, 3),
      frac_z_gt_3 = round(frac_z_gt_3, 3)
    )
  print(knitr::kable(display_tbl))
  cat("\n")

  plot_tbl <- passage_tbl %>%
    select(passage_position, median_z, iqr_z, q99_z, frac_z_gt_3) %>%
    pivot_longer(cols = -passage_position, names_to = "metric", values_to = "value") %>%
    group_by(metric) %>%
    arrange(passage_position, .by_group = TRUE) %>%
    mutate(smooth_value = smooth_series(value)) %>%
    ungroup()

  p_metrics <- ggplot(plot_tbl, aes(x = passage_position, y = value)) +
    geom_line(color = "grey75", linewidth = 0.5) +
    geom_point(color = "grey55", size = 1.5) +
    geom_line(aes(y = smooth_value), color = "#b2182b", linewidth = 1) +
    facet_wrap(~ metric, scales = "free_y", ncol = 2) +
    labs(
      title = paste("Lineage", current_lineage_id, ": passage-level trajectory"),
      subtitle = "Red line is the median-smoothed trajectory used for trend interpretation",
      x = "Passage position",
      y = "Summary value"
    ) +
    theme_minimal(base_size = 11)
  print(p_metrics)

  p_coverage <- passage_tbl %>%
    select(passage_position, n_manifest_images, n_matched_images, n_objects) %>%
    pivot_longer(cols = -passage_position, names_to = "metric", values_to = "value") %>%
    ggplot(aes(x = passage_position, y = value, color = metric)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 1.8) +
    labs(
      title = paste("Lineage", current_lineage_id, ": coverage along the lineage"),
      subtitle = "This distinguishes biological flattening from missing segmentation coverage",
      x = "Passage position",
      y = "Count",
      color = "Metric"
    ) +
    theme_minimal(base_size = 11)
  print(p_coverage)

  representative_passages <- choose_representative_passages(passage_tbl)
  density_tbl <- cell_tbl %>%
    filter(passage_id %in% representative_passages, is.finite(z), p_real >= 0.5) %>%
    group_by(passage_id) %>%
    slice_sample(n = min(dplyr::n(), 5000)) %>%
    ungroup() %>%
    mutate(
      passage_label = paste0("P", passage_position, ": ", passage_id),
      passage_label = factor(passage_label, levels = unique(passage_label))
    )

  if (nrow(density_tbl)) {
    p_density <- ggplot(density_tbl, aes(x = z, fill = passage_label)) +
      geom_density(alpha = 0.4) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
      labs(
        title = paste("Lineage", current_lineage_id, ": early / middle / late residual densities"),
        subtitle = "Use this to decide whether the trend reflects center shift, tail shift, or broadening",
        x = "Residual z",
        y = "Density",
        fill = "Passage"
      ) +
      theme_minimal(base_size = 11)
    print(p_density)
  }

  cat("\n")
}
```

## Reading This Notebook

Use the output in this order:

1. Start with the summary table to see whether each lineage looks like bulk shift, tail expansion/contraction, broadening/tightening, or no clear trend.
2. Check the overview plots to see whether the three lineages differ in center shift, upper-tail behavior, or dispersion.
3. For each lineage, inspect the detailed trajectory and coverage panels together before trusting any apparent plateau or jump.
4. Use the early/mid/late density comparison to decide whether the long-term change is mainly a center shift, a tail effect, or increased heterogeneity.

## Render

```{r render-help, eval=FALSE}
rmarkdown::render(
  "workflows/lineage_area_review.Rmd",
  params = list(
    lineage_id = NULL,
    all_results_dir = "data/lineage_area",
    manifest_csv = "manifests/image_manifest.csv",
    show_images = TRUE,
    max_images = 10
  )
)
```
