---
title: "Filtered Lineage Area Distributions"
output:
  html_document:
    toc: true
    toc_float: true
params:
  all_results_dir: "data/lineage_area"
  filtered_image_csv: "data/lineage_area/filtered_trajectory_images.csv"
  image_subset_mode: "filtered_progression"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5.8
)
```

## Overview

This notebook shows object-area distributions along passage trajectories for a configurable image subset.

The active subset mode is `r params$image_subset_mode`.

- `filtered_progression`: images from the streamlined trajectory filter
- `day0_only`: only images with `time_since_seeding_days == 0`

When `filtered_progression` is active, the filtered image list is read from `r params$filtered_image_csv`, so this notebook stays aligned with the trajectory notebook rather than reimplementing the filter.

```{r packages, include=FALSE}
required_packages <- c(
  "dplyr",
  "ggplot2",
  "purrr",
  "readr",
  "scales",
  "stringr",
  "tibble",
  "tidyr"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    sprintf(
      "Install required packages before rendering: %s",
      paste(missing_packages, collapse = ", ")
    ),
    call. = FALSE
  )
}

library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)
```

```{r helpers, include=FALSE}
resolve_project_root <- function() {
  candidates <- c(
    normalizePath(".", winslash = "/", mustWork = FALSE),
    normalizePath("..", winslash = "/", mustWork = FALSE),
    normalizePath("../..", winslash = "/", mustWork = FALSE)
  )

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "README.md")) &&
        dir.exists(file.path(candidate, "workflows"))) {
      return(candidate)
    }
  }

  stop("Could not locate project root from current working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

project_path <- function(path) {
  is_absolute <- grepl("^[A-Za-z]:[/\\\\]|^/", path)
  out <- path
  out[!is_absolute] <- file.path(project_root, path[!is_absolute])
  out
}

read_analysis_bundle <- function(path) {
  obj <- readRDS(path)
  list(
    lineage_id = as.character(obj$lineage_id),
    cells = tibble::as_tibble(obj$cells)
  )
}

load_png_cache_index <- function(manifest_csv = "manifests/image_manifest.csv", area_dir = "data/lineage_area", out_root = "data/lineage_png") {
  manifest_tbl <- readr::read_csv(project_path(manifest_csv), show_col_types = FALSE) %>%
    filter(artifact_type == "raw_image") %>%
    distinct(filename, filepath, .keep_all = TRUE)

  image_summary_files <- list.files(project_path(area_dir), pattern = "image_summary\\.csv$", recursive = TRUE, full.names = TRUE)

  if (!length(image_summary_files)) {
    return(tibble(
      filename = character(),
      raw_filepath = character(),
      png_relpath = character(),
      png_abspath = character(),
      png_exists = logical(),
      cache_lineage_id = character()
    ))
  }

  purrr::map_dfr(image_summary_files, function(path) {
    lineage_dir <- basename(dirname(path))

    readr::read_csv(path, show_col_types = FALSE) %>%
      distinct(filename, .keep_all = TRUE) %>%
      transmute(
        filename,
        cache_lineage_id = sub("^lineage_", "", lineage_dir)
      ) %>%
      left_join(manifest_tbl %>% transmute(filename, raw_filepath = filepath), by = "filename") %>%
      mutate(
        png_relpath = file.path(out_root, lineage_dir, sub("\\.[^.]+$", ".png", basename(filename))),
        png_abspath = project_path(png_relpath),
        png_exists = file.exists(png_abspath)
      )
  }) %>%
    distinct(filename, .keep_all = TRUE)
}

safe_quantile <- function(x, prob) {
  if (!length(x) || all(is.na(x))) return(NA_real_)
  as.numeric(stats::quantile(x, probs = prob, na.rm = TRUE, names = FALSE))
}

build_density_grid <- function(df, n_grid = 160) {
  df <- df %>%
    filter(is.finite(log10_area), is.finite(passage_position))

  if (!nrow(df)) {
    return(tibble())
  }

  x_grid <- seq(min(df$log10_area), max(df$log10_area), length.out = n_grid)

  df %>%
    group_by(lineage_id, passage_position) %>%
    group_modify(~ {
      x <- .x$log10_area
      if (length(x) < 2 || dplyr::n_distinct(x) < 2) {
        tibble(log10_area = x_grid, density = 0)
      } else {
        dens <- stats::density(x, from = min(x_grid), to = max(x_grid), n = n_grid, na.rm = TRUE)
        tibble(log10_area = dens$x, density = dens$y)
      }
    }) %>%
    ungroup()
}

trajectory_crop_card <- function(img_src, title_text, subtitle_text, crop_scale = sqrt(10)) {
  sprintf(
    paste0(
      "<div style=\"width:300px;\">",
      "<div style=\"width:300px;height:224px;overflow:hidden;border:1px solid #d0d0d0;background:#fafafa;\">",
      "<img src=\"%s\" style=\"width:100%%;height:100%%;object-fit:cover;transform:scale(%.4f);transform-origin:center center;display:block;\" />",
      "</div>",
      "<div style=\"font-size:13px;line-height:1.3;margin-top:6px;\">",
      "<div><strong>%s</strong></div>",
      "<div>%s</div>",
      "</div></div>"
    ),
    img_src,
    crop_scale,
    title_text,
    subtitle_text
  )
}

residualize_by_passage <- function(df, value_col) {
  value_sym <- rlang::ensym(value_col)

  df %>%
    group_by(lineage_id) %>%
    group_modify(~ {
      dat <- .x
      y <- dplyr::pull(dat, !!value_sym)
      ok <- is.finite(y) & is.finite(dat$passage_position)

      resid_vec <- rep(NA_real_, nrow(dat))
      if (sum(ok) >= 3) {
        unique_passages <- dplyr::n_distinct(dat$passage_position[ok])
        if (unique_passages >= 4) {
          fit <- stats::lm(y ~ splines::ns(dat$passage_position, df = min(3, unique_passages - 1)), subset = ok)
        } else if (unique_passages >= 2) {
          fit <- stats::lm(y ~ dat$passage_position, subset = ok)
        } else {
          fit <- stats::lm(y ~ 1, subset = ok)
        }
        resid_vec[ok] <- stats::resid(fit)
      }

      dat$residual <- resid_vec
      dat
    }) %>%
    ungroup() %>%
    rename(!!paste0(rlang::as_string(value_sym), "_resid") := residual)
}

adjust_for_confluence <- function(df, value_col = log10_area, covariate_col = confluence) {
  value_sym <- rlang::ensym(value_col)
  covariate_sym <- rlang::ensym(covariate_col)

  df %>%
    group_by(lineage_id) %>%
    group_modify(~ {
      dat <- .x
      y <- dplyr::pull(dat, !!value_sym)
      x <- dplyr::pull(dat, !!covariate_sym)
      ok <- is.finite(y) & is.finite(x)

      adjusted <- rep(NA_real_, nrow(dat))
      if (sum(ok) >= 10) {
        unique_x <- dplyr::n_distinct(round(x[ok], 4))
        if (unique_x >= 4) {
          fit <- stats::lm(y ~ splines::ns(x, df = min(3, unique_x - 1)), subset = ok)
        } else if (unique_x >= 2) {
          fit <- stats::lm(y ~ x, subset = ok)
        } else {
          fit <- stats::lm(y ~ 1, subset = ok)
        }

        pred <- rep(NA_real_, nrow(dat))
        pred[ok] <- stats::predict(fit, newdata = data.frame(x = x[ok]))
        adjusted[ok] <- y[ok] - pred[ok] + stats::median(y[ok], na.rm = TRUE)
      }

      dat$adjusted_value <- adjusted
      dat
    }) %>%
    ungroup() %>%
    rename(!!paste0(rlang::as_string(value_sym), "_adj_confluence") := adjusted_value)
}
```

```{r load-data, include=FALSE}
filtered_image_path <- project_path(params$filtered_image_csv)
png_cache_index <- load_png_cache_index(area_dir = params$all_results_dir)

analysis_paths <- list.files(
  project_path(params$all_results_dir),
  pattern = "lineage_area_analysis\\.rds$",
  recursive = TRUE,
  full.names = TRUE
)

if (!length(analysis_paths)) {
  stop(sprintf("No lineage analysis outputs found under %s", project_path(params$all_results_dir)), call. = FALSE)
}

summary_bundle <- purrr::map(analysis_paths, read_analysis_bundle)

all_cells_tbl <- purrr::map_dfr(summary_bundle, function(bundle) {
  bundle$cells %>%
    mutate(
      lineage_id = as.character(lineage_id),
      passage_id = as.character(passage_id),
      passage_position = suppressWarnings(as.integer(passage_position)),
      filename = as.character(filename),
      area_px = as.numeric(area_px),
      log_area = as.numeric(log_area),
      p_real = as.numeric(p_real)
    )
})

all_image_meta_tbl <- all_cells_tbl %>%
  transmute(
    lineage_id,
    passage_id,
    passage_position,
    filename,
    time_since_seeding_days = as.numeric(time_since_passage_start_hours) / 24,
    confluence = as.numeric(confluence_um2),
    cell_count = as.numeric(cell_count)
  ) %>%
  distinct(lineage_id, passage_id, passage_position, filename, .keep_all = TRUE)

filtered_image_tbl <- NULL
if (identical(params$image_subset_mode, "filtered_progression")) {
  if (!file.exists(filtered_image_path)) {
    stop(
      sprintf(
        "Filtered image CSV not found: %s. Render workflows/trajectory_alignment_exemplar_lineages_streamlined.Rmd first.",
        filtered_image_path
      ),
      call. = FALSE
    )
  }

  filtered_image_tbl <- readr::read_csv(filtered_image_path, show_col_types = FALSE) %>%
    transmute(
      lineage_id = as.character(lineage_id),
      passage_id = as.character(passage_id),
      passage_position = suppressWarnings(as.integer(passage_position)),
      filename = as.character(filename),
      time_since_seeding_days = as.numeric(time_since_seeding_days),
      confluence = as.numeric(confluence),
      cell_count = as.numeric(cell_count)
    ) %>%
    distinct(lineage_id, passage_id, passage_position, filename, .keep_all = TRUE)
} else if (identical(params$image_subset_mode, "day0_only")) {
  filtered_image_tbl <- all_image_meta_tbl %>%
    filter(is.finite(time_since_seeding_days), abs(time_since_seeding_days) < 1e-8)
} else {
  stop(
    sprintf(
      "Unknown image_subset_mode: %s. Supported modes are filtered_progression and day0_only.",
      params$image_subset_mode
    ),
    call. = FALSE
  )
}

if (!nrow(filtered_image_tbl)) {
  stop(sprintf("No images matched subset mode %s.", params$image_subset_mode), call. = FALSE)
}

subset_label <- dplyr::case_when(
  identical(params$image_subset_mode, "filtered_progression") ~ "Streamlined progression filter",
  identical(params$image_subset_mode, "day0_only") ~ "Day-0 images only",
  TRUE ~ params$image_subset_mode
)

filtered_cells_tbl <- all_cells_tbl %>%
  inner_join(
    filtered_image_tbl,
    by = c("lineage_id", "passage_id", "passage_position", "filename")
  ) %>%
  filter(is.finite(area_px), area_px > 0) %>%
  mutate(
    log10_area = log10(area_px),
    passage_f = factor(passage_position, levels = sort(unique(passage_position)))
  )

if (!nrow(filtered_cells_tbl)) {
  stop("No object rows remained after applying the filtered-image list.", call. = FALSE)
}

passage_summary_tbl <- filtered_cells_tbl %>%
  group_by(lineage_id, passage_id, passage_position) %>%
  summarise(
    n_images = n_distinct(filename),
    n_objects = n(),
    weighted_n = sum(p_real, na.rm = TRUE),
    q10 = safe_quantile(log10_area, 0.10),
    q25 = safe_quantile(log10_area, 0.25),
    median_log10_area = safe_quantile(log10_area, 0.50),
    q75 = safe_quantile(log10_area, 0.75),
    q90 = safe_quantile(log10_area, 0.90),
    .groups = "drop"
  )

overview_tbl <- filtered_cells_tbl %>%
  group_by(lineage_id) %>%
  summarise(
    surviving_images = n_distinct(filename),
    passages_with_images = n_distinct(passage_id),
    objects = n(),
    .groups = "drop"
  ) %>%
  arrange(lineage_id)

density_grid_tbl <- build_density_grid(filtered_cells_tbl)

adjusted_cells_tbl <- filtered_cells_tbl %>%
  adjust_for_confluence(log10_area, confluence)

adjusted_passage_summary_tbl <- adjusted_cells_tbl %>%
  group_by(lineage_id, passage_id, passage_position) %>%
  summarise(
    n_images = n_distinct(filename),
    q10 = safe_quantile(log10_area_adj_confluence, 0.10),
    q25 = safe_quantile(log10_area_adj_confluence, 0.25),
    median_log10_area_adj = safe_quantile(log10_area_adj_confluence, 0.50),
    q75 = safe_quantile(log10_area_adj_confluence, 0.75),
    q90 = safe_quantile(log10_area_adj_confluence, 0.90),
    .groups = "drop"
  )

adjusted_density_grid_tbl <- adjusted_cells_tbl %>%
  transmute(
    lineage_id,
    passage_position,
    log10_area = log10_area_adj_confluence
  ) %>%
  build_density_grid()

set.seed(1)
violin_tbl <- filtered_cells_tbl %>%
  group_by(lineage_id, passage_position) %>%
  group_modify(~ dplyr::slice_sample(.x, n = min(nrow(.x), 8000))) %>%
  ungroup()

image_area_tbl <- filtered_cells_tbl %>%
  group_by(lineage_id, passage_id, passage_position, filename) %>%
  summarise(
    median_log10_area = safe_quantile(log10_area, 0.50),
    q25_log10_area = safe_quantile(log10_area, 0.25),
    q75_log10_area = safe_quantile(log10_area, 0.75),
    n_objects = n(),
    .groups = "drop"
  ) %>%
  left_join(filtered_image_tbl, by = c("lineage_id", "passage_id", "passage_position", "filename")) %>%
  left_join(
    png_cache_index %>%
      select(filename, png_relpath, png_abspath, png_exists),
    by = "filename"
  ) %>%
  mutate(
    png_exists = dplyr::coalesce(png_exists, FALSE)
  )

passage_crop_tbl <- image_area_tbl %>%
  filter(png_exists) %>%
  group_by(lineage_id, passage_position) %>%
  group_modify(~ {
    dat <- .x %>%
      filter(is.finite(time_since_seeding_days)) %>%
      arrange(time_since_seeding_days, filename)

    if (!nrow(dat)) {
      return(.x %>% slice_head(n = 1))
    }

    target_time <- stats::median(dat$time_since_seeding_days, na.rm = TRUE)
    dat %>%
      mutate(time_distance = abs(time_since_seeding_days - target_time)) %>%
      slice_min(time_distance, n = 1, with_ties = FALSE) %>%
      select(-time_distance)
  }) %>%
  ungroup() %>%
  arrange(lineage_id, passage_position)

correlation_base_tbl <- image_area_tbl %>%
  select(lineage_id, passage_position, filename, median_log10_area, cell_count, time_since_seeding_days, confluence) %>%
  residualize_by_passage(median_log10_area) %>%
  residualize_by_passage(cell_count) %>%
  residualize_by_passage(time_since_seeding_days) %>%
  residualize_by_passage(confluence)

correlation_vars <- c(
  "median_log10_area_resid",
  "cell_count_resid",
  "time_since_seeding_days_resid",
  "confluence_resid"
)

correlation_labels <- c(
  median_log10_area_resid = "Median area",
  cell_count_resid = "Cell count",
  time_since_seeding_days_resid = "Time since seeding",
  confluence_resid = "Confluence"
)

cor_mat <- stats::cor(
  correlation_base_tbl[, correlation_vars],
  use = "pairwise.complete.obs",
  method = "spearman"
)

correlation_tbl <- as.data.frame(as.table(cor_mat), stringsAsFactors = FALSE) %>%
  as_tibble() %>%
  transmute(
    var_x = factor(correlation_labels[Var1], levels = unname(correlation_labels[correlation_vars])),
    var_y = factor(correlation_labels[Var2], levels = rev(unname(correlation_labels[correlation_vars]))),
    rho = Freq,
    label = sprintf("%.2f", rho)
  )
```

## Filtered Coverage

```{r filtered-coverage}
knitr::kable(overview_tbl, digits = 1)
```

```{r filtered-passages}
ggplot(
  passage_summary_tbl,
  aes(x = passage_position, y = n_images, color = lineage_id, group = lineage_id)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Filtered image coverage along each lineage",
    subtitle = paste("Subset:", subset_label),
    x = "Passage position",
    y = "Surviving images",
    color = "Lineage"
  ) +
  theme_minimal(base_size = 12)
```

## Quantile Trajectories

```{r quantile-ribbons}
ggplot(passage_summary_tbl, aes(x = passage_position, group = lineage_id)) +
  geom_ribbon(aes(ymin = q10, ymax = q90, fill = lineage_id), alpha = 0.18, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = lineage_id), alpha = 0.35, color = NA) +
  geom_line(aes(y = median_log10_area, color = lineage_id), linewidth = 1) +
  geom_point(aes(y = median_log10_area, size = n_images, color = lineage_id), alpha = 0.9) +
  facet_wrap(~ lineage_id, scales = "free_x", ncol = 1) +
  scale_size_continuous(range = c(1.8, 4.5), breaks = pretty_breaks()) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Passage-wise object area quantiles",
    subtitle = "Bands show the 10-90% and 25-75% ranges; the line marks the median of log10(area_px)",
    x = "Passage position",
    y = "log10(area_px)",
    color = "Lineage",
    fill = "Lineage",
    size = "Images"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

## Per-Passage Distributions

```{r violin-distributions}
ggplot(violin_tbl, aes(x = passage_f, y = log10_area, fill = lineage_id)) +
  geom_violin(scale = "width", trim = FALSE, color = NA, alpha = 0.7) +
  geom_boxplot(
    width = 0.14,
    outlier.shape = NA,
    fill = "white",
    color = "black",
    linewidth = 0.25
  ) +
  facet_wrap(~ lineage_id, scales = "free_x", ncol = 1) +
  labs(
    title = "Filtered object-area distributions by passage",
    subtitle = paste("Subset:", subset_label, "| violin width reflects density within each passage"),
    x = "Passage position",
    y = "log10(area_px)",
    fill = "Lineage"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

## Density Heatmaps

```{r density-heatmaps}
ggplot(density_grid_tbl, aes(x = passage_position, y = log10_area, fill = density)) +
  geom_raster(interpolate = TRUE) +
  facet_wrap(~ lineage_id, scales = "free_x", ncol = 1) +
  scale_fill_viridis_c(option = "C") +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Distribution shape along each lineage",
    subtitle = paste("Subset:", subset_label, "| darker regions indicate where object-area density concentrates"),
    x = "Passage position",
    y = "log10(area_px)",
    fill = "Density"
  ) +
  theme_minimal(base_size = 12)
```

## Representative Crops

```{r representative-crops, results='asis'}
if (!nrow(passage_crop_tbl)) {
  cat("No cached PNG crops were available for the filtered images.\n")
} else {
  for (lineage_name in unique(as.character(passage_crop_tbl$lineage_id))) {
    lineage_tbl <- passage_crop_tbl %>%
      filter(as.character(lineage_id) == lineage_name)

    if (!nrow(lineage_tbl)) next

    cards <- purrr::pmap_chr(
      list(
        img_src = lineage_tbl$png_relpath,
        passage = lineage_tbl$passage_position,
        time_days = lineage_tbl$time_since_seeding_days,
        median_area = lineage_tbl$median_log10_area
      ),
      function(img_src, passage, time_days, median_area) {
        trajectory_crop_card(
          img_src = file.path("..", img_src),
          title_text = sprintf("%s | P%s", lineage_name, passage),
          subtitle_text = sprintf("day %.2f | median %.2f", time_days, median_area)
        )
      }
    )

    cat(sprintf("<h3>%s</h3>", lineage_name))
    cat(
      "<div style=\"display:flex;flex-wrap:wrap;gap:10px;align-items:flex-start;\">",
      paste(cards, collapse = ""),
      "</div>"
    )
  }
}
```

## Residual Correlation Check

This view uses image-level median object area from the active subset and removes lineage-specific passage-position trends before computing Spearman correlations, so the heatmap highlights residual association beyond the main trajectory shape.

```{r residual-corrplot}
ggplot(correlation_tbl, aes(x = var_x, y = var_y, fill = rho)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = label), size = 4) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Residual correlation of image-level median object area",
    subtitle = "Residuals are taken after within-lineage adjustment for passage position",
    x = NULL,
    y = NULL,
    fill = "Spearman rho"
  ) +
  coord_fixed() +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

## Confluence-Adjusted Area

This section removes the within-lineage association between object area and confluence at the cell level, then shows the resulting adjusted area across passages on the same `log10(area_px)` scale.

```{r adjusted-quantile-ribbons}
ggplot(adjusted_passage_summary_tbl, aes(x = passage_position, group = lineage_id)) +
  geom_ribbon(aes(ymin = q10, ymax = q90, fill = lineage_id), alpha = 0.18, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = lineage_id), alpha = 0.35, color = NA) +
  geom_line(aes(y = median_log10_area_adj, color = lineage_id), linewidth = 1) +
  geom_point(aes(y = median_log10_area_adj, size = n_images, color = lineage_id), alpha = 0.9) +
  facet_wrap(~ lineage_id, scales = "free_x", ncol = 1) +
  scale_size_continuous(range = c(1.8, 4.5), breaks = pretty_breaks()) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Passage-wise object area after confluence adjustment",
    subtitle = paste("Subset:", subset_label, "| within each lineage, cell areas are residualized against confluence"),
    x = "Passage position",
    y = "Adjusted log10(area_px)",
    color = "Lineage",
    fill = "Lineage",
    size = "Images"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

```{r adjusted-density-heatmaps}
ggplot(adjusted_density_grid_tbl, aes(x = passage_position, y = log10_area, fill = density)) +
  geom_raster(interpolate = TRUE) +
  facet_wrap(~ lineage_id, scales = "free_x", ncol = 1) +
  scale_fill_viridis_c(option = "C") +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Confluence-adjusted distribution shape along each lineage",
    subtitle = paste("Subset:", subset_label, "| density is computed after within-lineage confluence adjustment"),
    x = "Passage position",
    y = "Adjusted log10(area_px)",
    fill = "Density"
  ) +
  theme_minimal(base_size = 12)
```

## Render

```{r render-help, eval=FALSE}
rmarkdown::render(
  "workflows/trajectory_alignment_exemplar_lineages_streamlined.Rmd",
  params = list(
    filtered_export_csv = "data/lineage_area/filtered_trajectory_images.csv"
  )
)

rmarkdown::render(
  "workflows/lineage_area_review.Rmd",
  params = list(
    all_results_dir = "data/lineage_area",
    filtered_image_csv = "data/lineage_area/filtered_trajectory_images.csv",
    image_subset_mode = "filtered_progression"
  )
)

rmarkdown::render(
  "workflows/lineage_area_review.Rmd",
  params = list(
    all_results_dir = "data/lineage_area",
    filtered_image_csv = "data/lineage_area/filtered_trajectory_images.csv",
    image_subset_mode = "day0_only"
  )
)
```
