---
title: "Lineage Embedding Review Dev"
output:
  html_document:
    toc: true
    toc_float: true
params:
  lineage_ids_csv: "manual_1,manual_2,manual_3"
  control_lineage_id: "manual_3"
  lineage_rds: "core_data/lineages.Rds"
  segmentation_manifest: "data/all_images/manifests/segmentation_images.csv"
  embedding_dir: "data/all_images/embeddings"
  cell_count_min: 500
  cell_count_max: 1200
  n_pcs: 50
  max_umap_cells: 10000
  umap_neighbors: 20
  umap_min_dist: 0.15
  seed: 1
  cache_dir: "/tmp/lineage_embedding_review_dev_cache"
  refresh_cache: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6
)
```

## Overview

This notebook builds a pooled embedding view across selected lineages directly from the raw per-object embedding CSVs.

Workflow:

1. Resolve all lineage images for the requested lineages.
2. Load per-object embedding tables for those images.
3. Keep only images with between `r params$cell_count_min` and `r params$cell_count_max` cells.
4. Run one pooled PCA across all retained cells using `embedding_*` features.
5. Downsample to at most `r format(params$max_umap_cells, big.mark = ",")` cells for UMAP.
6. Build balanced-subsample UMAP summaries, differential density views, and targeted PC comparisons.

This document can cache PCA and UMAP intermediates under `r params$cache_dir` to speed up repeated renders.

## Packages

```{r packages, include=FALSE}
required_packages <- c(
  "data.table",
  "dplyr",
  "ggplot2",
  "purrr",
  "readr",
  "stringr",
  "tibble",
  "tidyr",
  "scales",
  "uwot"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    sprintf(
      "Install required packages before rendering: %s",
      paste(missing_packages, collapse = ", ")
    ),
    call. = FALSE
  )
}

library(data.table)
library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(tidyr)
library(scales)
```

## Helpers

```{r helpers, include=FALSE}
resolve_project_root <- function() {
  candidates <- c(
    normalizePath(".", winslash = "/", mustWork = FALSE),
    normalizePath("..", winslash = "/", mustWork = FALSE),
    normalizePath("../..", winslash = "/", mustWork = FALSE)
  )

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "README.md")) &&
        dir.exists(file.path(candidate, "workflows")) &&
        dir.exists(file.path(candidate, "core_data"))) {
      return(candidate)
    }
  }

  stop("Could not locate project root from current working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

project_path <- function(path) {
  is_absolute <- grepl("^[A-Za-z]:[/\\\\]|^/", path)
  out <- path
  out[!is_absolute] <- file.path(project_root, path[!is_absolute])
  out
}

normalize_relpath <- function(path) {
  gsub("\\\\", "/", path)
}

cache_dir <- project_path(params$cache_dir)
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

cache_stub <- function(...) {
  raw <- paste(..., sep = "__")
  gsub("[^A-Za-z0-9._-]+", "_", raw)
}

cache_path <- function(name) {
  file.path(
    cache_dir,
    paste0(
      cache_stub(
        params$lineage_ids_csv,
        params$control_lineage_id,
        params$cell_count_min,
        params$cell_count_max,
        params$n_pcs,
        params$max_umap_cells,
        params$umap_neighbors,
        params$umap_min_dist
      ),
      "__",
      name
    )
  )
}

safe_readRDS <- function(path) {
  if (!file.exists(path)) {
    return(NULL)
  }

  tryCatch(
    readRDS(path),
    error = function(e) {
      message(sprintf("Ignoring unreadable cache file: %s", path))
      unlink(path)
      NULL
    }
  )
}

safe_saveRDS <- function(object, path) {
  tmp_path <- sprintf("%s.tmp.%s", path, Sys.getpid())
  on.exit(unlink(tmp_path), add = TRUE)
  saveRDS(object, tmp_path)
  ok <- file.rename(tmp_path, path)
  if (!ok) {
    file.copy(tmp_path, path, overwrite = TRUE)
    unlink(tmp_path)
  }
  invisible(path)
}

parse_lineage_ids <- function(x) {
  str_split(x, ",", simplify = TRUE) %>%
    as.vector() %>%
    str_trim() %>%
    .[nzchar(.)]
}

extract_lineage <- function(lineages, lineage_id) {
  if (lineage_id %in% names(lineages)) {
    return(lineages[[lineage_id]])
  }

  hit <- vapply(
    lineages,
    function(x) identical(as.character(x$lineage_id), as.character(lineage_id)),
    logical(1)
  )
  if (!any(hit)) {
    stop(sprintf("Lineage id not found: %s", lineage_id), call. = FALSE)
  }

  lineages[[which(hit)[[1]]]]
}

build_lineage_image_meta <- function(lineage_obj) {
  passage_positions <- stats::setNames(seq_along(lineage_obj$passage_ids), lineage_obj$passage_ids)

  meta_list <- lapply(names(lineage_obj$passages), function(passage_id) {
    current_passage_id <- as.character(passage_id)
    current_passage_number <- as.integer(passage_positions[[current_passage_id]])
    passage_dt <- data.table::as.data.table(lineage_obj$passages[[current_passage_id]])
    if (!nrow(passage_dt)) {
      return(NULL)
    }

    out <- data.table::copy(passage_dt)
    keep_cols <- intersect(c("id", "filename", "filepath", "field", "date"), names(out))
    keep_dt <- out[, ..keep_cols]
    keep_dt[, image_id := if ("filename" %in% names(keep_dt)) {
      ifelse(!is.na(filename) & nzchar(filename), as.character(filename), as.character(id))
    } else {
      as.character(id)
    }]
    keep_dt[, passage_id := current_passage_id]
    keep_dt[, passage_number := current_passage_number]
    keep_dt
  })

  meta_dt <- data.table::rbindlist(meta_list, use.names = TRUE, fill = TRUE)
  meta_dt[, id := as.character(id)]
  if ("filename" %in% names(meta_dt)) {
    meta_dt[, filename := as.character(filename)]
  }
  unique(meta_dt, by = intersect(c("image_id", "filename", "passage_id"), names(meta_dt)))
}

candidate_embedding_paths <- function(embedding_dir, image_id, filename = NA_character_) {
  stems <- unique(stats::na.omit(c(
    as.character(image_id),
    tools::file_path_sans_ext(basename(as.character(filename)))
  )))
  file.path(embedding_dir, paste0(stems, "_embeddings.csv"))
}

match_segmentation_row <- function(seg_dt, row) {
  seg_row <- NULL

  if ("filename" %in% names(seg_dt) && "filename" %in% names(row)) {
    seg_row <- seg_dt[filename == row$filename[[1]]]
    if (nrow(seg_row)) return(seg_row[1])
  }

  if ("source_filepath" %in% names(seg_dt) && "filepath" %in% names(row)) {
    seg_row <- seg_dt[source_filepath == row$filepath[[1]]]
    if (nrow(seg_row)) return(seg_row[1])
  }

  if ("source_basename" %in% names(seg_dt) && "filepath" %in% names(row)) {
    seg_row <- seg_dt[source_basename == basename(row$filepath[[1]])]
    if (nrow(seg_row)) return(seg_row[1])
  }

  if ("id" %in% names(seg_dt)) {
    seg_row <- seg_dt[id == row$id[[1]]]
    if (nrow(seg_row)) return(seg_row[1])
  }

  NULL
}

resolve_embedding_meta <- function(image_meta, seg_dt, embedding_dir, lineage_id) {
  resolved_list <- lapply(seq_len(nrow(image_meta)), function(i) {
    row <- image_meta[i]
    seg_row <- match_segmentation_row(seg_dt, row)

    embedding_path <- NA_character_

    if (!is.null(seg_row) && nrow(seg_row)) {
      if ("embedding_relpath" %in% names(seg_row) && file.exists(seg_row$embedding_relpath[[1]])) {
        embedding_path <- seg_row$embedding_relpath[[1]]
      }
    }

    if (!isTRUE(nzchar(embedding_path)) || is.na(embedding_path)) {
      candidates <- candidate_embedding_paths(
        embedding_dir = embedding_dir,
        image_id = row$image_id[[1]],
        filename = if ("filename" %in% names(row)) row$filename[[1]] else NA_character_
      )
      existing <- candidates[file.exists(candidates)]
      if (length(existing)) {
        embedding_path <- existing[[1]]
      }
    }

    data.table(
      lineage_id = lineage_id,
      image_id = as.character(row$image_id[[1]]),
      id = as.character(row$id[[1]]),
      filename = if ("filename" %in% names(row)) as.character(row$filename[[1]]) else NA_character_,
      field = if ("field" %in% names(row)) as.character(row$field[[1]]) else NA_character_,
      filepath = if ("filepath" %in% names(row)) as.character(row$filepath[[1]]) else NA_character_,
      date = if ("date" %in% names(row)) as.character(row$date[[1]]) else NA_character_,
      passage_id = as.character(row$passage_id[[1]]),
      passage_number = as.integer(row$passage_number[[1]]),
      embedding_path = embedding_path
    )
  })

  resolved_dt <- rbindlist(resolved_list, use.names = TRUE, fill = TRUE)
  resolved_dt[order(lineage_id, image_id, passage_number)][, .SD[1], by = .(lineage_id, image_id)]
}

load_embedding_table <- function(meta_row) {
  path <- meta_row$embedding_path[[1]]
  if (is.na(path) || !nzchar(path) || !file.exists(path)) {
    return(NULL)
  }

  dt <- data.table::fread(path)
  emb_cols <- grep("^embedding_[0-9]{3}$", names(dt), value = TRUE)
  if (!length(emb_cols)) {
    stop(sprintf("No embedding columns found in %s", path), call. = FALSE)
  }

  keep_cols <- unique(c("label", "log1p_area", emb_cols))
  keep_cols <- intersect(keep_cols, names(dt))
  dt <- dt[, ..keep_cols]
  dt[, lineage_id := meta_row$lineage_id[[1]]]
  dt[, image_id := meta_row$image_id[[1]]]
  dt[, id := meta_row$id[[1]]]
  dt[, filename := meta_row$filename[[1]]]
  dt[, field := meta_row$field[[1]]]
  dt[, date := meta_row$date[[1]]]
  dt[, passage_id := meta_row$passage_id[[1]]]
  dt[, passage_number := meta_row$passage_number[[1]]]
  dt
}

plot_umap <- function(df, color_col, title, discrete = FALSE, point_alpha = 0.6) {
  p <- ggplot(df, aes(x = UMAP1, y = UMAP2, color = .data[[color_col]])) +
    geom_point(size = 0.35, alpha = point_alpha) +
    labs(
      title = title,
      x = "UMAP1",
      y = "UMAP2",
      color = color_col
    ) +
    theme_bw(base_size = 11)

  if (discrete) {
    p + guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
  } else {
    p + scale_color_viridis_c(option = "C")
  }
}

plot_binned_embedding <- function(df, x_col, y_col, title, facet_col = NULL, bins = 55) {
  p <- ggplot(df, aes(x = .data[[x_col]], y = .data[[y_col]])) +
    stat_bin_2d(bins = bins) +
    scale_fill_viridis_c(option = "C", trans = "sqrt") +
    labs(
      title = title,
      x = x_col,
      y = y_col,
      fill = "Cells/bin"
    ) +
    coord_equal() +
    theme_bw(base_size = 11)

  if (!is.null(facet_col)) {
    p <- p + facet_wrap(stats::as.formula(paste("~", facet_col)))
  }

  p
}

plot_density_contours <- function(df, x_col, y_col, title, facet_col = NULL, bins = 10) {
  p <- ggplot(df, aes(x = .data[[x_col]], y = .data[[y_col]])) +
    stat_density_2d_filled(contour_var = "ndensity", bins = bins, alpha = 0.9) +
    labs(
      title = title,
      x = x_col,
      y = y_col,
      fill = "Norm. density"
    ) +
    coord_equal() +
    theme_bw(base_size = 11)

  if (!is.null(facet_col)) {
    p <- p + facet_wrap(stats::as.formula(paste("~", facet_col)))
  }

  p
}

compute_binned_density_difference <- function(df, x_col, y_col, group_col, group_a, group_b, bins = 70) {
  stopifnot(group_a != group_b)

  x_breaks <- seq(min(df[[x_col]], na.rm = TRUE), max(df[[x_col]], na.rm = TRUE), length.out = bins + 1)
  y_breaks <- seq(min(df[[y_col]], na.rm = TRUE), max(df[[y_col]], na.rm = TRUE), length.out = bins + 1)

  counts <- df %>%
    filter(.data[[group_col]] %in% c(group_a, group_b)) %>%
    mutate(
      x_bin = cut(.data[[x_col]], breaks = x_breaks, include.lowest = TRUE, labels = FALSE),
      y_bin = cut(.data[[y_col]], breaks = y_breaks, include.lowest = TRUE, labels = FALSE)
    ) %>%
    filter(!is.na(x_bin), !is.na(y_bin)) %>%
    count(.data[[group_col]], x_bin, y_bin, name = "n") %>%
    group_by(.data[[group_col]]) %>%
    mutate(density = n / sum(n)) %>%
    ungroup() %>%
    select(all_of(group_col), x_bin, y_bin, density)

  density_wide <- counts %>%
    pivot_wider(names_from = all_of(group_col), values_from = density, values_fill = 0)

  x_mid <- head(x_breaks, -1) + diff(x_breaks) / 2
  y_mid <- head(y_breaks, -1) + diff(y_breaks) / 2

  density_wide %>%
    mutate(
      x = x_mid[x_bin],
      y = y_mid[y_bin],
      density_diff = .data[[group_a]] - .data[[group_b]]
    )
}

plot_density_difference <- function(df, x_col, y_col, group_col, group_a, group_b, title, bins = 70) {
  diff_df <- compute_binned_density_difference(df, x_col, y_col, group_col, group_a, group_b, bins = bins)

  ggplot(diff_df, aes(x = x, y = y, fill = density_diff)) +
    geom_raster() +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B") +
    coord_equal() +
    labs(
      title = title,
      x = x_col,
      y = y_col,
      fill = sprintf("%s - %s", group_a, group_b)
    ) +
    theme_bw(base_size = 11)
}

balanced_sample_by_lineage <- function(df, lineage_col = "lineage_id", max_n = 10000, seed = 1) {
  if (nrow(df) <= max_n) {
    return(df)
  }

  set.seed(seed)
  df <- df %>%
    mutate(.sample_row_id = row_number())

  lineage_counts <- df %>%
    count(.data[[lineage_col]], name = "n_cells") %>%
    arrange(.data[[lineage_col]])

  n_groups <- nrow(lineage_counts)
  base_quota <- max(1L, floor(max_n / n_groups))

  sampled <- df %>%
    group_by(.data[[lineage_col]]) %>%
    group_modify(~ {
      keep_n <- min(nrow(.x), base_quota)
      slice_sample(.x, n = keep_n)
    }) %>%
    ungroup()

  n_remaining <- max_n - nrow(sampled)
  if (n_remaining <= 0) {
    return(sampled %>% select(-.sample_row_id))
  }

  leftovers <- anti_join(
    df,
    sampled %>% select(.sample_row_id),
    by = ".sample_row_id"
  )

  if (!nrow(leftovers)) {
    return(sampled %>% select(-.sample_row_id))
  }

  sampled_extra <- leftovers %>%
    group_by(.data[[lineage_col]]) %>%
    group_modify(~ {
      if (!nrow(.x)) return(.x[0, , drop = FALSE])
      keep_n <- min(nrow(.x), ceiling(n_remaining / n_groups))
      slice_sample(.x, n = keep_n)
    }) %>%
    ungroup() %>%
    slice_head(n = n_remaining)

  bind_rows(sampled, sampled_extra) %>%
    select(-.sample_row_id)
}

compute_lineage_pc_separation <- function(df, pc_cols, control_lineage_id) {
  df <- df %>%
    mutate(lineage_group = if_else(lineage_id == control_lineage_id, "control", "treated"))

  bind_rows(lapply(pc_cols, function(pc_name) {
    pc_values <- df[[pc_name]]
    control_vals <- pc_values[df$lineage_group == "control" & is.finite(pc_values)]
    treated_vals <- pc_values[df$lineage_group == "treated" & is.finite(pc_values)]

    pooled_sd <- sqrt(mean(c(stats::var(control_vals, na.rm = TRUE), stats::var(treated_vals, na.rm = TRUE)), na.rm = TRUE))
    pooled_sd <- ifelse(is.finite(pooled_sd) && pooled_sd > 0, pooled_sd, NA_real_)

    tibble(
      pc = pc_name,
      control_mean = mean(control_vals, na.rm = TRUE),
      treated_mean = mean(treated_vals, na.rm = TRUE),
      mean_diff = mean(treated_vals, na.rm = TRUE) - mean(control_vals, na.rm = TRUE),
      abs_mean_diff = abs(mean(treated_vals, na.rm = TRUE) - mean(control_vals, na.rm = TRUE)),
      cohens_d = abs((mean(treated_vals, na.rm = TRUE) - mean(control_vals, na.rm = TRUE)) / pooled_sd)
    )
  })) %>%
    arrange(desc(cohens_d), desc(abs_mean_diff), pc)
}
```

## Resolve Selected Images

```{r resolve-images}
set.seed(params$seed)

lineage_ids <- parse_lineage_ids(params$lineage_ids_csv)
if (!length(lineage_ids)) {
  stop("No lineage ids were provided.", call. = FALSE)
}

lineages <- readRDS(project_path(params$lineage_rds))

seg_dt <- fread(project_path(params$segmentation_manifest))
if ("id" %in% names(seg_dt)) {
  seg_dt[, id := as.character(id)]
}
if ("filename" %in% names(seg_dt)) {
  seg_dt[, filename := as.character(filename)]
}
if ("source_filepath" %in% names(seg_dt)) {
  seg_dt[, source_filepath := as.character(source_filepath)]
  seg_dt[, source_basename := basename(source_filepath)]
  seg_dt[, source_filepath := project_path(normalize_relpath(source_filepath))]
}
for (path_col in intersect(c("raw_relpath", "mask_relpath", "embedding_relpath"), names(seg_dt))) {
  seg_dt[, (path_col) := project_path(normalize_relpath(get(path_col)))]
}

embedding_dir <- project_path(params$embedding_dir)

resolved_image_tbl <- map_dfr(lineage_ids, function(lineage_id) {
  lineage_obj <- extract_lineage(lineages, lineage_id)
  image_meta <- build_lineage_image_meta(lineage_obj)
  resolve_embedding_meta(image_meta, seg_dt, embedding_dir, lineage_id) %>%
    as_tibble()
})

resolved_image_tbl %>%
  group_by(lineage_id) %>%
  summarise(
    n_images = n(),
    n_embedding_paths = sum(!is.na(embedding_path) & nzchar(embedding_path) & file.exists(embedding_path)),
    .groups = "drop"
  ) %>%
  knitr::kable()
```

## Load Embeddings

```{r load-embeddings}
embedding_load_cache <- cache_path("embedding_load_cache.rds")
embedding_cache_obj <- NULL

if (file.exists(embedding_load_cache) && !isTRUE(params$refresh_cache)) {
  embedding_cache_obj <- safe_readRDS(embedding_load_cache)
}

if (!is.null(embedding_cache_obj) && !isTRUE(params$refresh_cache)) {
  object_tbl <- embedding_cache_obj$object_tbl
  embedding_cols <- embedding_cache_obj$embedding_cols
  image_cell_summary <- embedding_cache_obj$image_cell_summary
  lineage_load_summary <- embedding_cache_obj$lineage_load_summary
} else {
  object_tbl <- map_dfr(seq_len(nrow(resolved_image_tbl)), function(i) {
    load_embedding_table(as.data.table(resolved_image_tbl[i, , drop = FALSE])) %>%
      as_tibble()
  })

  if (!nrow(object_tbl)) {
    stop("No object embeddings could be loaded for the requested lineages.", call. = FALSE)
  }

  embedding_cols <- grep("^embedding_[0-9]{3}$", names(object_tbl), value = TRUE)
  if (!length(embedding_cols)) {
    stop("No embedding columns were found after loading the object tables.", call. = FALSE)
  }

  image_cell_summary <- object_tbl %>%
    count(lineage_id, image_id, filename, passage_id, passage_number, name = "cell_count") %>%
    arrange(lineage_id, passage_number, filename)

  lineage_load_summary <- image_cell_summary %>%
    group_by(lineage_id) %>%
    summarise(
      n_images = n(),
      median_cells_per_image = median(cell_count),
      min_cells_per_image = min(cell_count),
      max_cells_per_image = max(cell_count),
      n_cells = sum(cell_count),
      .groups = "drop"
    )

  safe_saveRDS(
    list(
      object_tbl = object_tbl,
      embedding_cols = embedding_cols,
      image_cell_summary = image_cell_summary,
      lineage_load_summary = lineage_load_summary
    ),
    embedding_load_cache
  )
}

knitr::kable(lineage_load_summary, digits = 1)
```

## Filter Images By Cell Count

```{r filter-images}
filter_cache <- cache_path("filter_cache.rds")
filter_cache_obj <- NULL

if (file.exists(filter_cache) && !isTRUE(params$refresh_cache)) {
  filter_cache_obj <- safe_readRDS(filter_cache)
}

if (!is.null(filter_cache_obj) && !isTRUE(params$refresh_cache)) {
  eligible_images <- filter_cache_obj$eligible_images
  filtered_object_tbl <- filter_cache_obj$filtered_object_tbl
  filter_summary_tbl <- filter_cache_obj$filter_summary_tbl
} else {
  eligible_images <- image_cell_summary %>%
    filter(
      cell_count >= params$cell_count_min,
      cell_count <= params$cell_count_max
    )

  filtered_object_tbl <- object_tbl %>%
    inner_join(
      eligible_images %>%
        select(lineage_id, image_id),
      by = c("lineage_id", "image_id")
    )

  filter_summary_tbl <- eligible_images %>%
    group_by(lineage_id) %>%
    summarise(
      n_images_retained = n(),
      n_cells_retained = sum(cell_count),
      median_cells_per_image = median(cell_count),
      .groups = "drop"
    ) %>%
    full_join(
      image_cell_summary %>%
        group_by(lineage_id) %>%
        summarise(
          n_images_total = n(),
          n_cells_total = sum(cell_count),
          .groups = "drop"
        ),
      by = "lineage_id"
    ) %>%
    mutate(
      image_retention = percent(n_images_retained / n_images_total, accuracy = 0.1),
      cell_retention = percent(n_cells_retained / n_cells_total, accuracy = 0.1)
    ) %>%
    arrange(lineage_id)

  safe_saveRDS(
    list(
      eligible_images = eligible_images,
      filtered_object_tbl = filtered_object_tbl,
      filter_summary_tbl = filter_summary_tbl
    ),
    filter_cache
  )
}

knitr::kable(filter_summary_tbl, digits = 1)
```

```{r filter-cell-count-plot}
ggplot(image_cell_summary, aes(x = cell_count, fill = lineage_id)) +
  geom_histogram(bins = 40, alpha = 0.65, position = "identity") +
  geom_vline(
    xintercept = c(params$cell_count_min, params$cell_count_max),
    linetype = "dashed",
    linewidth = 0.7
  ) +
  facet_wrap(~ lineage_id, scales = "free_y") +
  labs(
    title = "Per-image cell counts with retained range",
    x = "Cells per image",
    y = "Images",
    fill = "Lineage"
  ) +
  theme_bw(base_size = 11)
```

## Run Pooled PCA

```{r pooled-pca}
pca_cache <- cache_path("pca_cache.rds")
pca_cache_obj <- NULL

if (file.exists(pca_cache) && !isTRUE(params$refresh_cache)) {
  pca_cache_obj <- safe_readRDS(pca_cache)
}

if (!is.null(pca_cache_obj) && !isTRUE(params$refresh_cache)) {
  pca_model <- pca_cache_obj$pca_model
  pca_tbl <- pca_cache_obj$pca_tbl
  pca_variance_tbl <- pca_cache_obj$pca_variance_tbl
  pc_cols <- pca_cache_obj$pc_cols
} else {
  feature_mat <- filtered_object_tbl %>%
    select(all_of(embedding_cols)) %>%
    as.matrix()

  keep_rows <- stats::complete.cases(feature_mat)
  feature_mat <- feature_mat[keep_rows, , drop = FALSE]
  pca_input_tbl <- filtered_object_tbl[keep_rows, , drop = FALSE]

  if (!nrow(feature_mat)) {
    stop("No complete embedding rows remained after filtering.", call. = FALSE)
  }

  scaled_mat <- scale(feature_mat)
  pca_model <- prcomp(scaled_mat, center = FALSE, scale. = FALSE)
  n_pcs <- min(params$n_pcs, ncol(pca_model$x))
  pc_cols <- sprintf("PC%02d", seq_len(n_pcs))

  pca_tbl <- pca_input_tbl %>%
    bind_cols(
      as_tibble(pca_model$x[, seq_len(n_pcs), drop = FALSE], .name_repair = "minimal") %>%
        stats::setNames(pc_cols)
    )

  pca_variance_tbl <- tibble(
    pc = factor(pc_cols, levels = pc_cols),
    variance_explained = (pca_model$sdev[seq_len(n_pcs)] ^ 2) / sum(pca_model$sdev ^ 2),
    cumulative_variance = cumsum((pca_model$sdev[seq_len(n_pcs)] ^ 2) / sum(pca_model$sdev ^ 2))
  )

  safe_saveRDS(
    list(
      pca_model = pca_model,
      pca_tbl = pca_tbl,
      pca_variance_tbl = pca_variance_tbl,
      pc_cols = pc_cols
    ),
    pca_cache
  )
}

tibble(
  retained_images = nrow(eligible_images),
  retained_cells = nrow(filtered_object_tbl),
  complete_cells_for_pca = nrow(pca_tbl),
  embedding_dims = length(embedding_cols),
  n_pcs = length(pc_cols),
  pca_cache = pca_cache
) %>%
  knitr::kable()
```

```{r pca-variance-plot}
ggplot(pca_variance_tbl, aes(x = pc, y = variance_explained)) +
  geom_col(fill = "#2C7FB8") +
  geom_line(aes(y = cumulative_variance, group = 1), color = "#D95F0E", linewidth = 0.8) +
  geom_point(aes(y = cumulative_variance), color = "#D95F0E", size = 1.2) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Pooled PCA variance explained",
    x = NULL,
    y = "Variance explained"
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Downsample For UMAP

```{r umap-input}
umap_cache <- cache_path("umap_cache.rds")
umap_cache_obj <- NULL

if (file.exists(umap_cache) && !isTRUE(params$refresh_cache)) {
  umap_cache_obj <- safe_readRDS(umap_cache)
}

if (!is.null(umap_cache_obj) && !isTRUE(params$refresh_cache)) {
  umap_tbl <- umap_cache_obj$umap_tbl
  n_neighbors <- umap_cache_obj$n_neighbors
  umap_source_tbl <- umap_cache_obj$umap_source_tbl
} else {
  umap_source_tbl <- balanced_sample_by_lineage(
    pca_tbl,
    lineage_col = "lineage_id",
    max_n = params$max_umap_cells,
    seed = params$seed
  )

  n_neighbors <- min(params$umap_neighbors, max(2, nrow(umap_source_tbl) - 1))
  if (nrow(umap_source_tbl) < 3) {
    stop("UMAP requires at least 3 cells after filtering/downsampling.", call. = FALSE)
  }

  set.seed(params$seed)
  umap_coords <- uwot::umap(
    umap_source_tbl %>% select(all_of(pc_cols)) %>% as.matrix(),
    n_neighbors = n_neighbors,
    min_dist = params$umap_min_dist,
    metric = "euclidean",
    verbose = FALSE,
    ret_model = FALSE
  )

  umap_tbl <- umap_source_tbl %>%
    mutate(
      UMAP1 = umap_coords[, 1],
      UMAP2 = umap_coords[, 2]
    )

  safe_saveRDS(
    list(
      umap_tbl = umap_tbl,
      umap_source_tbl = umap_source_tbl,
      n_neighbors = n_neighbors
    ),
    umap_cache
  )
}

tibble(
  cells_entering_umap = nrow(umap_tbl),
  downsampled = nrow(pca_tbl) > nrow(umap_tbl),
  umap_neighbors = n_neighbors,
  umap_min_dist = params$umap_min_dist,
  umap_cache = umap_cache
) %>%
  knitr::kable()
```

```{r umap-balance-summary}
umap_tbl %>%
  count(lineage_id, name = "n_umap_cells") %>%
  left_join(pca_tbl %>% count(lineage_id, name = "n_pca_cells"), by = "lineage_id") %>%
  mutate(sample_fraction = n_umap_cells / n_pca_cells) %>%
  arrange(lineage_id) %>%
  knitr::kable(digits = 3)
```

## UMAP Summary

```{r umap-summary-plot}
umap_summary_tbl <- umap_tbl %>%
  group_by(lineage_id, passage_number) %>%
  summarise(
    n_cells = n(),
    mean_log1p_area = mean(log1p_area, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(umap_summary_tbl, aes(x = passage_number, y = lineage_id)) +
  geom_point(aes(size = n_cells, color = mean_log1p_area), alpha = 0.9) +
  scale_size_area(max_size = 10) +
  scale_color_viridis_c(option = "C") +
  labs(
    title = "UMAP sample composition by lineage and passage",
    x = "Passage number",
    y = "Lineage",
    size = "UMAP cells",
    color = "Mean log1p area"
  ) +
  theme_bw(base_size = 11)
```

## Lineage-Separating PCs

```{r separating-pcs}
pc_separation_tbl <- compute_lineage_pc_separation(
  pca_tbl,
  pc_cols = pc_cols,
  control_lineage_id = params$control_lineage_id
)

top_sep_pcs <- head(pc_separation_tbl$pc, min(4, nrow(pc_separation_tbl)))
top_sep_pc_pair <- head(pc_separation_tbl$pc, min(2, nrow(pc_separation_tbl)))

pc_separation_tbl %>%
  slice_head(n = 10) %>%
  knitr::kable(digits = 3)
```

```{r separating-pcs-bar}
ggplot(pc_separation_tbl %>% slice_head(n = 10), aes(x = reorder(pc, cohens_d), y = cohens_d)) +
  geom_col(fill = "#C44E52") +
  coord_flip() +
  labs(
    title = sprintf("Top PCs separating treated lineages from control (%s)", params$control_lineage_id),
    x = NULL,
    y = "Absolute Cohen's d"
  ) +
  theme_bw(base_size = 11)
```

```{r separating-pcs-density}
pca_tbl %>%
  select(lineage_id, all_of(top_sep_pcs)) %>%
  pivot_longer(cols = all_of(top_sep_pcs), names_to = "pc", values_to = "score") %>%
  mutate(
    lineage_group = if_else(lineage_id == params$control_lineage_id, "control", "treated")
  ) %>%
  ggplot(aes(x = score, color = lineage_group, fill = lineage_group)) +
  geom_density(alpha = 0.18, linewidth = 0.8) +
  facet_wrap(~ pc, scales = "free", ncol = 2) +
  labs(
    title = "Top separating PCs: treated vs control density",
    x = "PC score",
    y = "Density",
    color = NULL,
    fill = NULL
  ) +
  theme_bw(base_size = 11)
```

```{r separating-pcs-by-lineage}
pca_tbl %>%
  select(lineage_id, all_of(top_sep_pcs)) %>%
  pivot_longer(cols = all_of(top_sep_pcs), names_to = "pc", values_to = "score") %>%
  mutate(lineage_id = factor(lineage_id, levels = lineage_ids)) %>%
  ggplot(aes(x = score, color = lineage_id, fill = lineage_id)) +
  geom_density(alpha = 0.14, linewidth = 0.8) +
  facet_wrap(~ pc, scales = "free", ncol = 2) +
  labs(
    title = "Top separating PCs by lineage",
    x = "PC score",
    y = "Density",
    color = "Lineage",
    fill = "Lineage"
  ) +
  theme_bw(base_size = 11)
```

```{r separating-pcs-2d-density}
if (length(top_sep_pc_pair) >= 2) {
  plot_density_contours(
    pca_tbl %>% mutate(lineage_id = factor(lineage_id, levels = lineage_ids)),
    x_col = top_sep_pc_pair[[1]],
    y_col = top_sep_pc_pair[[2]],
    title = sprintf("2D density for top discriminative PCs: %s vs %s", top_sep_pc_pair[[1]], top_sep_pc_pair[[2]]),
    facet_col = "lineage_id",
    bins = 9
  )
}
```

```{r separating-pcs-2d-diff}
if (length(top_sep_pc_pair) >= 2) {
  plot_density_difference(
    pca_tbl %>% mutate(lineage_group = if_else(lineage_id == params$control_lineage_id, "control", "treated")),
    x_col = top_sep_pc_pair[[1]],
    y_col = top_sep_pc_pair[[2]],
    group_col = "lineage_group",
    group_a = "treated",
    group_b = "control",
    title = sprintf("Differential density for top discriminative PCs: %s vs %s", top_sep_pc_pair[[1]], top_sep_pc_pair[[2]]),
    bins = 70
  )
}
```

```{r separating-pcs-passage-quantiles}
pca_tbl %>%
  select(lineage_id, passage_number, all_of(top_sep_pcs)) %>%
  pivot_longer(cols = all_of(top_sep_pcs), names_to = "pc", values_to = "score") %>%
  group_by(lineage_id, passage_number, pc) %>%
  summarise(
    q10 = quantile(score, probs = 0.10, na.rm = TRUE),
    q50 = quantile(score, probs = 0.50, na.rm = TRUE),
    q90 = quantile(score, probs = 0.90, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(lineage_id = factor(lineage_id, levels = lineage_ids)) %>%
  ggplot(aes(x = passage_number, group = lineage_id)) +
  geom_ribbon(aes(ymin = q10, ymax = q90, fill = lineage_id), alpha = 0.18, color = NA) +
  geom_line(aes(y = q50, color = lineage_id), linewidth = 0.9) +
  geom_point(aes(y = q50, color = lineage_id), size = 1.3) +
  facet_wrap(~ pc, scales = "free_y", ncol = 2) +
  labs(
    title = "Top separating PCs across passage",
    x = "Passage number",
    y = "PC score quantiles",
    color = "Lineage",
    fill = "Lineage"
  ) +
  theme_bw(base_size = 11)
```

```{r separating-pcs-passage-treated-vs-control}
pca_tbl %>%
  mutate(lineage_group = if_else(lineage_id == params$control_lineage_id, "control", "treated")) %>%
  select(lineage_group, passage_number, all_of(top_sep_pcs)) %>%
  pivot_longer(cols = all_of(top_sep_pcs), names_to = "pc", values_to = "score") %>%
  group_by(lineage_group, passage_number, pc) %>%
  summarise(
    q10 = quantile(score, probs = 0.10, na.rm = TRUE),
    q50 = quantile(score, probs = 0.50, na.rm = TRUE),
    q90 = quantile(score, probs = 0.90, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = passage_number, group = lineage_group)) +
  geom_ribbon(aes(ymin = q10, ymax = q90, fill = lineage_group), alpha = 0.18, color = NA) +
  geom_line(aes(y = q50, color = lineage_group), linewidth = 0.9) +
  geom_point(aes(y = q50, color = lineage_group), size = 1.3) +
  facet_wrap(~ pc, scales = "free_y", ncol = 2) +
  labs(
    title = "Top separating PCs across passage: treated vs control",
    x = "Passage number",
    y = "PC score quantiles",
    color = NULL,
    fill = NULL
  ) +
  theme_bw(base_size = 11)
```

## Alternative Views

For these dense pooled embeddings, occupancy and density views are easier to read than raw points. The plots below summarize local cell concentration and explicitly compare treated versus control density.

```{r umap-lineage-heatmap}
plot_binned_embedding(
  umap_tbl %>% mutate(lineage_id = factor(lineage_id, levels = lineage_ids)),
  x_col = "UMAP1",
  y_col = "UMAP2",
  title = "UMAP occupancy heatmap by lineage",
  facet_col = "lineage_id",
  bins = 50
)
```

```{r umap-lineage-contours}
plot_density_contours(
  umap_tbl %>% mutate(lineage_id = factor(lineage_id, levels = lineage_ids)),
  x_col = "UMAP1",
  y_col = "UMAP2",
  title = "UMAP density contours by lineage",
  facet_col = "lineage_id",
  bins = 9
)
```

```{r umap-treated-vs-control-diff}
plot_density_difference(
  umap_tbl %>% mutate(lineage_group = if_else(lineage_id == params$control_lineage_id, "control", "treated")),
  x_col = "UMAP1",
  y_col = "UMAP2",
  group_col = "lineage_group",
  group_a = "treated",
  group_b = "control",
  title = "UMAP differential density: treated minus control",
  bins = 70
)
```

```{r discriminative-pc-heatmap}
if (length(top_sep_pc_pair) >= 2) {
  plot_binned_embedding(
    pca_tbl %>% mutate(lineage_id = factor(lineage_id, levels = lineage_ids)),
    x_col = top_sep_pc_pair[[1]],
    y_col = top_sep_pc_pair[[2]],
    title = sprintf("Occupancy heatmap for top discriminative PCs: %s vs %s", top_sep_pc_pair[[1]], top_sep_pc_pair[[2]]),
    facet_col = "lineage_id",
    bins = 50
  )
}
```
