---
title: "Lineage Embedding Review"
output:
  html_document:
    toc: true
    toc_float: true
params:
  results_dir: "data/lineage_embeddings"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5
)
```

## Overview

This notebook reviews the per-lineage embedding analysis outputs saved under `data/lineage_embeddings`.

For each lineage, the workflow:

1. Loads all segmentation embedding tables for lineage images.
2. Computes per-image cell counts and divides images into count tertiles.
3. Runs a separate PCA within each tertile using `log1p_area` and `embedding_*`.
4. Correlates each PC score with passage number.
5. Exports tiled TIFF panels showing representative low-score and high-score cells for the strongest PCs.

## Packages

```{r packages}
required_packages <- c(
  "data.table",
  "dplyr",
  "ggplot2",
  "purrr",
  "readr",
  "stringr",
  "tibble",
  "tidyr",
  "tiff"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    sprintf("Install required packages before rendering: %s", paste(missing_packages, collapse = ", ")),
    call. = FALSE
  )
}

library(data.table)
library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(tidyr)
library(tiff)
```

## Helpers

```{r helpers}
resolve_project_root <- function() {
  candidates <- c(
    normalizePath(".", winslash = "/", mustWork = FALSE),
    normalizePath("..", winslash = "/", mustWork = FALSE),
    normalizePath("../..", winslash = "/", mustWork = FALSE)
  )

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "README.md")) &&
        dir.exists(file.path(candidate, "workflows"))) {
      return(candidate)
    }
  }

  stop("Could not locate project root from current working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

project_path <- function(path) {
  is_absolute <- grepl("^[A-Za-z]:[/\\\\]|^/", path)
  out <- path
  out[!is_absolute] <- file.path(project_root, path[!is_absolute])
  out
}

read_if_exists <- function(path, reader) {
  if (!file.exists(path)) {
    return(NULL)
  }
  reader(path)
}

render_tiff_panel <- function(path, title = basename(path)) {
  img <- tiff::readTIFF(path, native = FALSE, all = FALSE)
  if (length(dim(img)) == 3L) {
    img <- apply(img[, , seq_len(min(3, dim(img)[3])), drop = FALSE], c(1, 2), mean)
  }
  img <- as.matrix(img)
  plot.new()
  usr <- par("usr")
  raster_img <- as.raster(img[nrow(img):1, , drop = FALSE], max = 1)
  rasterImage(raster_img, usr[1], usr[3], usr[2], usr[4], interpolate = FALSE)
  title(main = title, cex.main = 0.9)
  invisible(NULL)
}

discover_lineage_dirs <- function(results_dir) {
  root <- project_path(results_dir)
  if (!dir.exists(root)) {
    stop(sprintf("Results dir not found: %s", root), call. = FALSE)
  }
  dirs <- list.dirs(root, recursive = FALSE, full.names = TRUE)
  dirs[grepl("^lineage_", basename(dirs))]
}

discover_quantile_dirs <- function(lineage_dir) {
  dirs <- list.dirs(lineage_dir, recursive = FALSE, full.names = TRUE)
  dirs[basename(dirs) %in% c("Q1_low", "Q2_mid", "Q3_high")]
}

read_lineage_bundle <- function(lineage_dir) {
  lineage_id <- sub("^lineage_", "", basename(lineage_dir))
  list(
    lineage_id = lineage_id,
    lineage_dir = lineage_dir,
    image_counts = read_if_exists(file.path(lineage_dir, "image_cell_counts.csv"), fread),
    image_plot = file.path(lineage_dir, "image_cell_count_distribution.png"),
    warning_log = read_if_exists(file.path(lineage_dir, "warning_log.txt"), readLines),
    lineage_scores = read_if_exists(file.path(lineage_dir, "lineage_pc_scores.csv"), fread),
    quantiles = setNames(lapply(discover_quantile_dirs(lineage_dir), function(q_dir) {
      list(
        dir = q_dir,
        name = basename(q_dir),
        correlations = read_if_exists(file.path(q_dir, "pc_passage_correlations.csv"), fread),
        scores = read_if_exists(file.path(q_dir, "pc_scores.csv"), fread),
        top3_panels = list.dirs(file.path(q_dir, "top3_abs_correlation"), recursive = FALSE, full.names = TRUE),
        top5_panels = list.dirs(file.path(q_dir, "top5_abs_correlation"), recursive = FALSE, full.names = TRUE)
      )
    }), basename(discover_quantile_dirs(lineage_dir)))
  )
}

build_overview_tbl <- function(lineage_bundles) {
  bind_rows(lapply(lineage_bundles, function(bundle) {
    image_counts <- bundle$image_counts
    tibble(
      lineage_id = bundle$lineage_id,
      n_images = if (is.null(image_counts)) NA_integer_ else nrow(image_counts),
      n_images_with_cells = if (is.null(image_counts)) NA_integer_ else sum(is.finite(image_counts$cell_count) & image_counts$cell_count > 0),
      median_cell_count = if (is.null(image_counts)) NA_real_ else median(image_counts$cell_count, na.rm = TRUE),
      max_cell_count = if (is.null(image_counts)) NA_real_ else max(image_counts$cell_count, na.rm = TRUE),
      n_warnings = length(bundle$warning_log %||% character())
    )
  }))
}

`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

plot_top_pc_correlations <- function(cor_dt, lineage_id, quantile_name, top_n = 10L) {
  if (is.null(cor_dt) || !nrow(cor_dt)) {
    return(NULL)
  }
  plot_dt <- cor_dt %>%
    as_tibble() %>%
    slice_head(n = min(top_n, n())) %>%
    mutate(pc = factor(pc, levels = rev(pc)))

  ggplot(plot_dt, aes(x = pc, y = correlation, fill = correlation > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c(`TRUE` = "#2C7BB6", `FALSE` = "#D7191C"), guide = "none") +
    labs(
      title = sprintf("%s: top PC correlations with passage in %s", lineage_id, quantile_name),
      x = NULL,
      y = "Pearson correlation with passage number"
    ) +
    theme_bw(base_size = 11)
}

plot_top_pc_trajectories <- function(score_dt, cor_dt, lineage_id, quantile_name, top_n = 3L) {
  if (is.null(score_dt) || !nrow(score_dt) || is.null(cor_dt) || !nrow(cor_dt)) {
    return(NULL)
  }
  top_pcs <- head(cor_dt$pc, min(top_n, nrow(cor_dt)))
  if (!length(top_pcs)) {
    return(NULL)
  }
  plot_dt <- score_dt %>%
    as_tibble() %>%
    select(any_of(c("passage_number", top_pcs))) %>%
    pivot_longer(cols = all_of(top_pcs), names_to = "pc", values_to = "score") %>%
    group_by(passage_number, pc) %>%
    summarise(
      median_score = median(score, na.rm = TRUE),
      q25 = quantile(score, probs = 0.25, na.rm = TRUE),
      q75 = quantile(score, probs = 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  ggplot(plot_dt, aes(x = passage_number, y = median_score, color = pc, fill = pc)) +
    geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.15, linewidth = 0) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.2) +
    labs(
      title = sprintf("%s: top PC trajectories in %s", lineage_id, quantile_name),
      x = "Passage number",
      y = "Median PC score"
    ) +
    theme_bw(base_size = 11)
}

collect_panel_files <- function(panel_dirs) {
  if (is.null(panel_dirs) || !length(panel_dirs)) {
    return(character())
  }
  unlist(lapply(panel_dirs, function(dir) {
    files <- list.files(dir, pattern = "_panel\\.tiff$", full.names = TRUE)
    files[order(files)]
  }))
}
```

## Load Results

```{r load-results}
lineage_dirs <- discover_lineage_dirs(params$results_dir)
lineage_bundles <- lapply(lineage_dirs, read_lineage_bundle)

if (!length(lineage_bundles)) {
  stop("No lineage embedding outputs were found.", call. = FALSE)
}
```

## Output Summary

```{r overview-table}
overview_tbl <- build_overview_tbl(lineage_bundles)
knitr::kable(overview_tbl, digits = 2)
```

## Cell Count Overview

```{r cross-lineage-cell-counts}
count_plot_dt <- bind_rows(lapply(lineage_bundles, function(bundle) {
  if (is.null(bundle$image_counts)) return(NULL)
  as_tibble(bundle$image_counts) %>% mutate(lineage_id = bundle$lineage_id)
}))

ggplot(count_plot_dt, aes(x = cell_count, fill = lineage_id)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  facet_wrap(~ lineage_id, scales = "free_y") +
  labs(
    title = "Per-image cell count distributions by lineage",
    x = "Cells per image",
    y = "Images"
  ) +
  theme_bw(base_size = 11)
```

## Detailed Review

```{r detailed-review, results='asis'}
for (bundle in lineage_bundles) {
  cat("\n\n## ", bundle$lineage_id, "\n\n", sep = "")
  cat("Output directory: `", normalizePath(bundle$lineage_dir, winslash = "/", mustWork = FALSE), "`\n\n", sep = "")

  if (file.exists(bundle$image_plot)) {
    cat("### Image Count Tertiles\n\n")
    print(knitr::include_graphics(bundle$image_plot))
    cat("\n\n")
  }

  if (length(bundle$warning_log %||% character())) {
    cat("### Warnings\n\n")
    warning_preview <- head(bundle$warning_log, 20)
    cat("```text\n", paste(warning_preview, collapse = "\n"), "\n```\n\n", sep = "")
  }

  for (quantile_name in names(bundle$quantiles)) {
    q_bundle <- bundle$quantiles[[quantile_name]]
    cat("### ", quantile_name, "\n\n", sep = "")

    cor_plot <- plot_top_pc_correlations(q_bundle$correlations, bundle$lineage_id, quantile_name)
    if (!is.null(cor_plot)) {
      print(cor_plot)
    }

    traj_plot <- plot_top_pc_trajectories(q_bundle$scores, q_bundle$correlations, bundle$lineage_id, quantile_name)
    if (!is.null(traj_plot)) {
      print(traj_plot)
    }

    if (!is.null(q_bundle$correlations) && nrow(q_bundle$correlations)) {
      cat("\nTop PCs by absolute passage correlation:\n\n")
      print(knitr::kable(as_tibble(q_bundle$correlations) %>% slice_head(n = min(10, n())), digits = 3))
      cat("\n\n")
    }

    top3_panels <- collect_panel_files(q_bundle$top3_panels)
    if (length(top3_panels)) {
      cat("#### Representative Cell Panels: Top 3 Correlated PCs\n\n")
      for (panel_path in top3_panels) {
        panel_title <- paste(basename(dirname(panel_path)), basename(panel_path))
        render_tiff_panel(panel_path, panel_title)
      }
      cat("\n")
    }

    top5_panels <- collect_panel_files(q_bundle$top5_panels)
    if (length(top5_panels)) {
      cat("#### Representative Cell Panels: Top 5 Correlated PCs\n\n")
      for (panel_path in top5_panels) {
        panel_title <- paste(basename(dirname(panel_path)), basename(panel_path))
        render_tiff_panel(panel_path, panel_title)
      }
      cat("\n")
    }
  }
}
```
