---
title: "Exemplar Lineage Trajectory Alignment"
output:
  html_document:
    toc: true
    toc_float: true
params:
  lineage_rds: "core_data/lineages.Rds"
  embedding_csv: "data/all_images/manifests/raw_image_embeddings.csv"
  manifest_csv: "manifests/image_manifest.csv"
  filtered_export_csv: "data/lineage_area/filtered_trajectory_images.csv"
  exemplar_lineages: null
  control_lineage: null
  n_top_pcs: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5
)
```

## Overview

This streamlined notebook keeps the core diagnostics for exemplar-lineage trajectory alignment and removes the exploratory side paths.

```{r packages, include=FALSE}
required_packages <- c(
  "data.table",
  "dplyr",
  "ggplot2",
  "mgcv",
  "patchwork",
  "purrr",
  "readr",
  "scales",
  "stringr",
  "tibble",
  "tidyr"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    sprintf("Install required packages before rendering: %s", paste(missing_packages, collapse = ", ")),
    call. = FALSE
  )
}

library(data.table)
library(dplyr)
library(ggplot2)
library(mgcv)
library(patchwork)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)
```

```{r plot-style, include=FALSE}
theme_set(
  theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "right"
    )
)

lineage_palette <- c(
  "#1B4F72",
  "#B03A2E",
  "#117864",
  "#7D6608",
  "#6C3483"
)
```

```{r helpers, include=FALSE}
`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

resolve_project_root <- function() {
  candidates <- c(
    normalizePath(".", winslash = "/", mustWork = FALSE),
    normalizePath("..", winslash = "/", mustWork = FALSE),
    normalizePath("../..", winslash = "/", mustWork = FALSE)
  )

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "README.md")) &&
        dir.exists(file.path(candidate, "workflows"))) {
      return(candidate)
    }
  }

  stop("Could not locate project root from current working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

project_path <- function(path) {
  is_absolute <- grepl("^[A-Za-z]:[/\\\\]|^/", path)
  out <- path
  out[!is_absolute] <- file.path(project_root, path[!is_absolute])
  out
}

ensure_parent_dir <- function(path) {
  dir.create(dirname(project_path(path)), recursive = TRUE, showWarnings = FALSE)
}

parse_mixed_datetime <- function(x) {
  x <- as.character(x)
  out <- as.POSIXct(rep(NA_character_, length(x)), tz = "UTC")
  formats <- c(
    "%Y-%m-%d %H:%M:%S",
    "%Y-%m-%dT%H:%M:%SZ",
    "%m/%d/%Y %H:%M",
    "%m/%d/%Y %H:%M:%S"
  )

  for (fmt in formats) {
    missing <- is.na(out) & !is.na(x) & nzchar(x)
    if (!any(missing)) next
    out[missing] <- as.POSIXct(x[missing], format = fmt, tz = "UTC")
  }

  out
}

embedding_column_names <- function(x) {
  grep("^embedding_[0-9]{3}$", names(x), value = TRUE)
}

extract_lineage <- function(lineages, lineage_id) {
  if (lineage_id %in% names(lineages)) {
    return(lineages[[lineage_id]])
  }

  hit <- vapply(lineages, function(x) identical(as.character(x$lineage_id), as.character(lineage_id)), logical(1))
  if (!any(hit)) {
    stop(sprintf("Lineage id not found: %s", lineage_id), call. = FALSE)
  }

  lineages[[which(hit)[[1]]]]
}

build_lineage_image_meta <- function(lineage_obj) {
  passage_positions <- stats::setNames(seq_along(lineage_obj$passage_ids), lineage_obj$passage_ids)

  image_meta_list <- lapply(names(lineage_obj$passages), function(passage_id) {
    passage_dt <- data.table::as.data.table(lineage_obj$passages[[passage_id]])
    if (!nrow(passage_dt)) return(NULL)

    keep_cols <- intersect(
      c("id", "filename", "field", "date", "areaOccupied_um2", "cellCount", "condition_label", "event", "passage"),
      names(passage_dt)
    )
    image_meta <- unique(passage_dt[, ..keep_cols], by = "filename")
    image_meta[, passage_id := as.character(passage_id)]
    image_meta[, lineage_id := as.character(lineage_obj$lineage_id)]
    image_meta[, passage_position := unname(passage_positions[passage_id])]
    image_meta
  })

  image_meta <- data.table::rbindlist(image_meta_list, fill = TRUE)
  if (!nrow(image_meta)) {
    stop(sprintf("No manifest rows found in lineage %s.", lineage_obj$lineage_id), call. = FALSE)
  }

  image_meta[, image_time := parse_mixed_datetime(date)]
  image_meta[, passage_start_time := suppressWarnings(min(image_time, na.rm = TRUE)), by = passage_id]
  image_meta[!is.finite(as.numeric(passage_start_time)), passage_start_time := as.POSIXct(NA, tz = "UTC")]
  image_meta[, time_since_seeding_hours := as.numeric(difftime(image_time, passage_start_time, units = "hours"))]
  image_meta[, time_since_seeding_days := time_since_seeding_hours / 24]
  image_meta[, confluence := suppressWarnings(as.numeric(areaOccupied_um2))]
  image_meta[, cell_count := suppressWarnings(as.numeric(cellCount))]
  image_meta[, log_confluence := log1p(pmax(confluence, 0))]
  image_meta[, log_cell_count := log1p(pmax(cell_count, 0))]

  image_meta
}

choose_exemplar_lineages <- function(lineages, embedding_tbl, n = 3L) {
  coverage_tbl <- purrr::map_dfr(names(lineages), function(lineage_name) {
    lineage_obj <- lineages[[lineage_name]]
    meta <- build_lineage_image_meta(lineage_obj)
    tibble(
      lineage_id = as.character(lineage_obj$lineage_id),
      n_manifest_images = nrow(meta),
      n_matched_images = sum(meta$filename %in% embedding_tbl$filename)
    )
  })

  coverage_tbl %>%
    arrange(desc(n_matched_images), desc(n_manifest_images), lineage_id) %>%
    slice_head(n = n) %>%
    pull(lineage_id)
}

load_exemplar_panel <- function(lineage_rds, embedding_csv, exemplar_lineages = NULL) {
  lineages <- readRDS(project_path(lineage_rds))
  embedding_tbl <- data.table::fread(project_path(embedding_csv)) %>%
    as_tibble()

  embedding_cols <- embedding_column_names(embedding_tbl)
  if (!length(embedding_cols)) {
    stop("No embedding columns were found in the image-level embedding CSV.", call. = FALSE)
  }

  if (is.null(exemplar_lineages) || !length(exemplar_lineages)) {
    exemplar_lineages <- choose_exemplar_lineages(lineages, embedding_tbl, n = 3L)
  }

  lineage_meta <- purrr::map_dfr(exemplar_lineages, function(lineage_id) {
    build_lineage_image_meta(extract_lineage(lineages, lineage_id)) %>% as_tibble()
  })

  if (!nrow(lineage_meta)) {
    stop("No lineage metadata rows were recovered for the exemplar lineages.", call. = FALSE)
  }

  joined <- lineage_meta %>%
    inner_join(embedding_tbl, by = "filename")

  if (!nrow(joined)) {
    stop("No exemplar-lineage images matched the embedding CSV.", call. = FALSE)
  }

  joined %>%
    mutate(
      lineage_id = factor(lineage_id, levels = exemplar_lineages),
      passage_position = as.integer(passage_position)
    ) %>%
    distinct(lineage_id, passage_id, filename, .keep_all = TRUE)
}

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || diff(rng) <= 0) {
    return(rep(NA_real_, length(x)))
  }
  (x - rng[1]) / diff(rng)
}

load_png_cache_index <- function(manifest_csv = "manifests/image_manifest.csv", area_dir = "data/lineage_area", out_root = "data/lineage_png") {
  manifest_tbl <- readr::read_csv(project_path(manifest_csv), show_col_types = FALSE) %>%
    filter(artifact_type == "raw_image") %>%
    distinct(filename, filepath, .keep_all = TRUE)

  image_summary_files <- list.files(project_path(area_dir), pattern = "image_summary\\.csv$", recursive = TRUE, full.names = TRUE)

  if (!length(image_summary_files)) {
    return(tibble(
      filename = character(),
      raw_filepath = character(),
      png_relpath = character(),
      png_abspath = character(),
      png_exists = logical(),
      cache_lineage_id = character()
    ))
  }

  purrr::map_dfr(image_summary_files, function(path) {
    lineage_dir <- basename(dirname(path))

    readr::read_csv(path, show_col_types = FALSE) %>%
      distinct(filename, .keep_all = TRUE) %>%
      transmute(
        filename,
        cache_lineage_id = sub("^lineage_", "", lineage_dir)
      ) %>%
      left_join(manifest_tbl %>% transmute(filename, raw_filepath = filepath), by = "filename") %>%
      mutate(
        png_relpath = file.path(out_root, lineage_dir, sub("\\.[^.]+$", ".png", basename(filename))),
        png_abspath = project_path(png_relpath),
        png_exists = file.exists(png_abspath)
      )
  }) %>%
    distinct(filename, .keep_all = TRUE)
}

pick_evenly_spaced_images <- function(df, n_keep = 12L) {
  dat <- df %>%
    filter(is.finite(latent_score), !is.na(png_exists), png_exists) %>%
    arrange(latent_score)

  if (!nrow(dat)) {
    return(dat)
  }

  if (nrow(dat) <= n_keep) {
    return(
      dat %>%
        mutate(selection_rank = row_number())
    )
  }

  targets <- seq(0.025, 0.975, length.out = n_keep)
  picked_rows <- integer(0)

  for (target in targets) {
    remaining_idx <- setdiff(seq_len(nrow(dat)), picked_rows)
    remaining <- dat[remaining_idx, , drop = FALSE]
    if (!nrow(remaining)) break

    chosen <- remaining %>%
      mutate(distance_to_target = abs(latent_score - target)) %>%
      slice_min(distance_to_target, n = 1, with_ties = FALSE)

    picked_rows <- c(picked_rows, match(chosen$filename[[1]], dat$filename))
  }

  dat[picked_rows, , drop = FALSE] %>%
    distinct(filename, .keep_all = TRUE) %>%
    arrange(latent_score) %>%
    mutate(selection_rank = row_number())
}

robust_high_confluence <- function(x, prob = 0.99) {
  x <- x[is.finite(x)]
  if (!length(x)) {
    return(NA_real_)
  }
  as.numeric(stats::quantile(x, probs = prob, na.rm = TRUE, names = FALSE))
}

trajectory_crop_card <- function(img_src, title_text, subtitle_text, crop_scale = sqrt(10)) {
  sprintf(
    paste0(
      "<div style=\"width:150px;\">",
      "<div style=\"width:150px;height:112px;overflow:hidden;border:1px solid #d0d0d0;background:#fafafa;\">",
      "<img src=\"%s\" style=\"width:100%%;height:100%%;object-fit:cover;transform:scale(%.4f);transform-origin:center center;display:block;\" />",
      "</div>",
      "<div style=\"font-size:11px;line-height:1.25;margin-top:4px;\">",
      "<div><strong>%s</strong></div>",
      "<div>%s</div>",
      "</div></div>"
    ),
    img_src,
    crop_scale,
    title_text,
    subtitle_text
  )
}

fit_shared_pc12_curve <- function(df, grid_n = 500L) {
  dat <- df %>%
    filter(is.finite(PC1), is.finite(PC2), is.finite(time_since_seeding_days))

  if (nrow(dat) < 10 || dplyr::n_distinct(round(dat$PC1, 4)) < 6) {
    stop("Too few distinct PC1/PC2 points to fit a shared spline trajectory.", call. = FALSE)
  }

  k_basis <- min(7, max(4, floor(dplyr::n_distinct(round(dat$PC1, 3)) / 6)))
  spline_fit <- mgcv::gam(PC2 ~ s(PC1, k = k_basis, bs = "cr"), data = dat, method = "REML")

  x_grid <- seq(min(dat$PC1, na.rm = TRUE), max(dat$PC1, na.rm = TRUE), length.out = grid_n)
  y_grid <- as.numeric(predict(spline_fit, newdata = tibble(PC1 = x_grid)))

  step_len <- c(0, sqrt(diff(x_grid)^2 + diff(y_grid)^2))
  lambda <- cumsum(step_len)

  grid_tbl <- tibble(
    PC1 = x_grid,
    PC2 = y_grid,
    lambda = lambda
  )

  nearest_idx <- vapply(seq_len(nrow(dat)), function(i) {
    which.min((grid_tbl$PC1 - dat$PC1[[i]])^2 + (grid_tbl$PC2 - dat$PC2[[i]])^2)
  }, integer(1))

  projection_tbl <- dat %>%
    mutate(
      curve_PC1 = grid_tbl$PC1[nearest_idx],
      curve_PC2 = grid_tbl$PC2[nearest_idx],
      lambda = grid_tbl$lambda[nearest_idx],
      distance_to_curve = sqrt((PC1 - curve_PC1)^2 + (PC2 - curve_PC2)^2)
    )

  lambda_cor <- suppressWarnings(cor(projection_tbl$lambda, projection_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman"))
  if (is.finite(lambda_cor) && lambda_cor < 0) {
    lambda_max <- max(grid_tbl$lambda, na.rm = TRUE)
    grid_tbl <- grid_tbl %>%
      mutate(lambda = lambda_max - lambda) %>%
      arrange(lambda)
    projection_tbl <- projection_tbl %>%
      mutate(lambda = lambda_max - lambda)
  }

  list(
    curve_path_tbl = grid_tbl,
    curve_projection_tbl = projection_tbl
  )
}
```

```{r load-data, include=FALSE}
panel_tbl <- load_exemplar_panel(
  lineage_rds = params$lineage_rds,
  embedding_csv = params$embedding_csv,
  exemplar_lineages = params$exemplar_lineages
)

png_cache_index <- load_png_cache_index(manifest_csv = params$manifest_csv)

embedding_cols <- embedding_column_names(panel_tbl)

analysis_tbl <- panel_tbl %>%
  mutate(
    lineage_id = factor(lineage_id, levels = unique(lineage_id)),
    passage_label = paste0("P", passage_position)
  )

lineage_levels <- levels(analysis_tbl$lineage_id)
```

## Sample Coverage by Lineage and Passage

```{r coverage-plots}
coverage_tbl <- analysis_tbl %>%
  dplyr::count(lineage_id, passage_position, time_since_seeding_days, name = "n_images")

ggplot(coverage_tbl, aes(x = time_since_seeding_days, y = factor(passage_position), size = n_images, color = lineage_id)) +
  geom_point(alpha = 0.8) +
  scale_size_continuous(range = c(2.5, 7)) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Sampling coverage by lineage, passage, and time since seeding",
    x = "Time since seeding (days)",
    y = "Passage position",
    color = "Lineage",
    size = "Images"
  )
```

## Confluence and Cell Count Versus Time Since Seeding

```{r covariate-pairs}
covariate_long <- analysis_tbl %>%
  select(lineage_id, passage_position, time_since_seeding_days, confluence, cell_count) %>%
  pivot_longer(
    cols = c(confluence, cell_count),
    names_to = "metric",
    values_to = "value"
  )

ggplot(covariate_long, aes(x = time_since_seeding_days, y = value, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Confluence and cell count versus time since seeding",
    x = "Time since seeding (days)",
    y = NULL,
    color = "Lineage"
  )
```

## PCA Variance Explained

```{r pca, include=FALSE}
complete_tbl <- analysis_tbl %>%
  filter(
    if_all(all_of(embedding_cols), is.finite),
    is.finite(time_since_seeding_days)
  )

embedding_mat <- as.matrix(complete_tbl[, embedding_cols])
storage.mode(embedding_mat) <- "double"

pca_fit <- prcomp(embedding_mat, center = TRUE, scale. = TRUE)

n_top_pcs <- min(params$n_top_pcs, ncol(pca_fit$x))
score_cols <- paste0("PC", seq_len(n_top_pcs))
score_tbl <- bind_cols(
  complete_tbl %>% select(-all_of(embedding_cols)),
  as_tibble(pca_fit$x[, seq_len(n_top_pcs), drop = FALSE], .name_repair = "minimal")
)
names(score_tbl)[(ncol(score_tbl) - n_top_pcs + 1):ncol(score_tbl)] <- score_cols

var_explained <- (pca_fit$sdev^2) / sum(pca_fit$sdev^2)
var_tbl <- tibble(
  pc = paste0("PC", seq_along(var_explained)),
  variance_explained = var_explained,
  cumulative_variance = cumsum(var_explained)
)
```

```{r variance-explained}
ggplot(var_tbl %>% slice_head(n = 15), aes(x = reorder(pc, variance_explained), y = variance_explained)) +
  geom_col(fill = "#2C7FB8") +
  geom_text(aes(label = percent(variance_explained, accuracy = 0.1)), hjust = -0.05, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Variance explained by the leading principal components",
    x = NULL,
    y = "Fraction of total embedding variance"
  )
```

## PCA State Space Views

```{r pca-color-views}
pca_by_lineage_plot <- ggplot(score_tbl, aes(x = PC1, y = PC2, color = lineage_id)) +
  geom_point(alpha = 0.8, size = 2.2) +
  stat_ellipse(linewidth = 0.7, alpha = 0.4) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Colored by lineage",
    x = "PC1",
    y = "PC2",
    color = "Lineage"
  )

plot_by_numeric <- function(var_name, title_text) {
  ggplot(score_tbl, aes(x = PC1, y = PC2, color = .data[[var_name]])) +
    geom_point(size = 2.1, alpha = 0.85) +
    scale_color_viridis_c(option = "C", na.value = "grey80") +
    labs(
      title = title_text,
      x = "PC1",
      y = "PC2",
      color = var_name
    )
}

pca_by_confluence_plot <- plot_by_numeric("confluence", "Colored by confluence")
pca_by_cell_count_plot <- plot_by_numeric("cell_count", "Colored by cell count")
pca_by_time_plot <- plot_by_numeric("time_since_seeding_days", "Colored by time since seeding")

(pca_by_lineage_plot + pca_by_confluence_plot) /
  (pca_by_cell_count_plot + pca_by_time_plot) +
  plot_annotation(title = "PCA state space views")
```

```{r latent-prep, include=FALSE}
pc1_oriented <- score_tbl$PC1
if (stats::cor(pc1_oriented, score_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman") < 0) {
  pc1_oriented <- -pc1_oriented
}

latent_tbl <- score_tbl %>%
  mutate(
    latent_score = rescale01(pc1_oriented),
    distance_to_curve = NA_real_
  )

curve_path_tbl <- tibble(
  PC1 = numeric(),
  PC2 = numeric(),
  lambda = numeric()
)

shared_curve_fit <- tryCatch(
  fit_shared_pc12_curve(score_tbl),
  error = function(e) NULL
)

if (!is.null(shared_curve_fit)) {
  curve_path_tbl <- shared_curve_fit$curve_path_tbl

  latent_tbl <- latent_tbl %>%
    select(-distance_to_curve) %>%
    left_join(
      shared_curve_fit$curve_projection_tbl %>%
        transmute(
          filename,
          latent_score = rescale01(lambda),
          distance_to_curve = distance_to_curve
        ),
      by = "filename",
      suffix = c("_pc1", "")
    ) %>%
    mutate(
      latent_score = coalesce(latent_score, latent_score_pc1)
    ) %>%
    select(-latent_score_pc1)
}
```

```{r filtered-trajectory-prep, include=FALSE}
latent_tbl <- latent_tbl %>%
  left_join(
    png_cache_index %>%
      select(filename, raw_filepath, png_relpath, png_abspath, png_exists, cache_lineage_id),
    by = "filename"
  ) %>%
  mutate(
    png_exists = dplyr::coalesce(png_exists, FALSE)
  )

confluence_reference_max <- robust_high_confluence(
  latent_tbl$confluence[is.finite(latent_tbl$time_since_seeding_days) & latent_tbl$time_since_seeding_days > 0],
  prob = 0.99
)
confluence_keep_threshold <- 0.70 * confluence_reference_max

filtered_tbl <- latent_tbl %>%
  filter(
    is.finite(time_since_seeding_days),
    time_since_seeding_days > 0,
    is.finite(confluence),
    confluence <= confluence_keep_threshold
  )

filtered_export_tbl <- filtered_tbl %>%
  transmute(
    lineage_id,
    passage_id,
    passage_position,
    filename,
    time_since_seeding_days,
    confluence,
    cell_count,
    latent_score,
    confluence_reference_max,
    confluence_keep_threshold
  ) %>%
  distinct(lineage_id, passage_id, passage_position, filename, .keep_all = TRUE) %>%
  arrange(lineage_id, passage_position, time_since_seeding_days, filename)

ensure_parent_dir(params$filtered_export_csv)
readr::write_csv(filtered_export_tbl, project_path(params$filtered_export_csv))

filtered_crop_tbl <- filtered_tbl %>%
  group_by(lineage_id) %>%
  group_modify(~ pick_evenly_spaced_images(.x, n_keep = 12L)) %>%
  ungroup()

passage_dist_plots <- purrr::map(lineage_levels, function(lineage_id) {
  plot_tbl <- filtered_tbl %>%
    filter(lineage_id == .env$lineage_id) %>%
    count(passage_position, name = "n_images")

  ggplot(plot_tbl, aes(x = factor(passage_position), y = n_images)) +
    geom_col(fill = lineage_palette[match(lineage_id, lineage_levels)]) +
    labs(
      title = paste("Passage distribution:", lineage_id),
      x = "Passage position",
      y = "Surviving images"
    )
})
```

## Filtered Trajectory in PCA Space

Images are retained only if `time_since_seeding_days > 0` and `confluence <= 70%` of a robust high-confluence reference, where the reference is the 99th percentile of nonzero-time confluence.

```{r filtered-trajectory-pca}
filtered_pca_plot <- ggplot(latent_tbl, aes(x = PC1, y = PC2)) +
  geom_point(color = "grey82", alpha = 0.45, size = 1.5) +
  geom_point(
    data = filtered_tbl,
    aes(color = lineage_id),
    alpha = 0.85,
    size = 2.1
  ) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Filtered trajectory view in PCA space",
    subtitle = paste0(
      "Reference max confluence = ", scales::label_number(accuracy = 0.1, scale_cut = scales::cut_short_scale())(confluence_reference_max),
      "; keep threshold = ", scales::label_number(accuracy = 0.1, scale_cut = scales::cut_short_scale())(confluence_keep_threshold),
      "; surviving images = ", nrow(filtered_tbl)
    ),
    x = "PC1",
    y = "PC2",
    color = "Lineage"
  )

if (nrow(curve_path_tbl) >= 2) {
  filtered_pca_plot <- filtered_pca_plot +
    geom_path(
      data = curve_path_tbl,
      mapping = aes(x = PC1, y = PC2),
      inherit.aes = FALSE,
      linewidth = 1.1,
      color = "black"
    )
}

filtered_pca_plot
```

## Passage Distributions of Surviving Images

```{r filtered-passage-distributions}
wrap_plots(passage_dist_plots, ncol = 1)
```

## Representative Filtered Crops

These crops use the center 10% of each cached lineage PNG. Within each lineage, up to 12 images are sampled to span the filtered latent trajectory; if fewer than 12 images survive the filter, all survivors are shown.

```{r filtered-crop-gallery, results='asis'}
if (!nrow(filtered_crop_tbl)) {
  cat("No filter-passing cached PNGs were available.\n")
} else {
  for (lineage_id in lineage_levels) {
    lineage_tbl <- filtered_crop_tbl %>%
      filter(lineage_id == .env$lineage_id)

    if (!nrow(lineage_tbl)) next

    cards <- purrr::pmap_chr(
      list(
        img_src = lineage_tbl$png_relpath,
        rank = lineage_tbl$selection_rank,
        passage = lineage_tbl$passage_position,
        time_days = lineage_tbl$time_since_seeding_days
      ),
      function(img_src, rank, passage, time_days) {
        trajectory_crop_card(
          img_src = file.path("..", img_src),
          title_text = sprintf("%s #%02d", lineage_id, rank),
          subtitle_text = sprintf("P%s | day %.2f", passage, time_days)
        )
      }
    )

    cat(sprintf("<h3>%s (%d shown)</h3>", lineage_id, nrow(lineage_tbl)))
    cat(
      "<div style=\"display:flex;flex-wrap:wrap;gap:10px;align-items:flex-start;\">",
      paste(cards, collapse = ""),
      "</div>"
    )
  }
}
```
