---
title: "Exemplar Lineage Trajectory Alignment"
output:
  html_document:
    toc: true
    toc_float: true
params:
  lineage_rds: "core_data/lineages.Rds"
  embedding_csv: "data/all_images/manifests/raw_image_embeddings.csv"
  exemplar_lineages: null
  control_lineage: null
  n_top_pcs: 8
  n_stage_bins: 5
  candidate_top_n: 20
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5
)
```

## Overview

This notebook is a first-pass exploratory workflow for asking whether three exemplar lineages appear to move through a shared morphological progression in image-embedding space.

The emphasis is on interpretable diagnostics rather than a black-box trajectory method. The workflow starts with direct checks of coverage and covariates, uses PCA as the main low-dimensional representation, and only then adds a simple latent progression score for provisional cross-lineage alignment.

Evidence for a shared trajectory would look like this:

- the lineages overlap in the same low-dimensional regions rather than occupying disjoint clouds
- the major embedding axes vary smoothly with `time_since_seeding`
- images with similar latent progression scores can come from different lineages at different absolute times
- the latent score still carries temporal structure after checking `confluence` and `cell_count`

Evidence against a shared trajectory would look like this:

- lineage-specific clusters with little overlap in PCA space
- different lineages following visibly different directions in low-dimensional space
- latent progression being almost entirely reducible to trivial density proxies such as `cell_count` or `confluence`
- stage bins that separate lineages rather than mixing them

## Packages

The core workflow uses standard packages. `uwot` and `princurve` are optional extras for a secondary UMAP view and an optional principal-curve score.

```{r packages}
required_packages <- c(
  "data.table",
  "dplyr",
  "ggplot2",
  "mgcv",
  "purrr",
  "readr",
  "scales",
  "stringr",
  "tibble",
  "tidyr"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    sprintf("Install required packages before rendering: %s", paste(missing_packages, collapse = ", ")),
    call. = FALSE
  )
}

library(data.table)
library(dplyr)
library(ggplot2)
library(mgcv)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

has_uwot <- requireNamespace("uwot", quietly = TRUE)
has_princurve <- requireNamespace("princurve", quietly = TRUE)
```

## Plot Style

The notebook uses a single theme and a stable lineage palette so plots are easy to compare.

```{r plot-style}
theme_set(
  theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "right"
    )
)

lineage_palette <- c(
  "#1B4F72",
  "#B03A2E",
  "#117864",
  "#7D6608",
  "#6C3483"
)
```

Short plot-reading convention used throughout:

- points are individual images
- smooth curves summarize lineage-level trends without assuming equal sampling
- `time_since_seeding` is treated as a within-passage clock, not as a synonym for morphological state

## Helpers

The notebook reconstructs image metadata from `lineages.Rds`, then joins it to the one-row-per-image embedding table.

```{r helpers}
`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

resolve_project_root <- function() {
  candidates <- c(
    normalizePath(".", winslash = "/", mustWork = FALSE),
    normalizePath("..", winslash = "/", mustWork = FALSE),
    normalizePath("../..", winslash = "/", mustWork = FALSE)
  )

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "README.md")) &&
        dir.exists(file.path(candidate, "workflows"))) {
      return(candidate)
    }
  }

  stop("Could not locate project root from current working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

project_path <- function(path) {
  is_absolute <- grepl("^[A-Za-z]:[/\\\\]|^/", path)
  out <- path
  out[!is_absolute] <- file.path(project_root, path[!is_absolute])
  out
}

normalize_relpath <- function(path) {
  gsub("\\\\", "/", path)
}

parse_mixed_datetime <- function(x) {
  x <- as.character(x)
  out <- as.POSIXct(rep(NA_character_, length(x)), tz = "UTC")
  formats <- c(
    "%Y-%m-%d %H:%M:%S",
    "%Y-%m-%dT%H:%M:%SZ",
    "%m/%d/%Y %H:%M",
    "%m/%d/%Y %H:%M:%S"
  )

  for (fmt in formats) {
    missing <- is.na(out) & !is.na(x) & nzchar(x)
    if (!any(missing)) next
    out[missing] <- as.POSIXct(x[missing], format = fmt, tz = "UTC")
  }

  out
}

embedding_column_names <- function(x) {
  grep("^embedding_[0-9]{3}$", names(x), value = TRUE)
}

extract_lineage <- function(lineages, lineage_id) {
  if (lineage_id %in% names(lineages)) {
    return(lineages[[lineage_id]])
  }

  hit <- vapply(lineages, function(x) identical(as.character(x$lineage_id), as.character(lineage_id)), logical(1))
  if (!any(hit)) {
    stop(sprintf("Lineage id not found: %s", lineage_id), call. = FALSE)
  }

  lineages[[which(hit)[[1]]]]
}

build_lineage_image_meta <- function(lineage_obj) {
  passage_positions <- stats::setNames(seq_along(lineage_obj$passage_ids), lineage_obj$passage_ids)

  image_meta_list <- lapply(names(lineage_obj$passages), function(passage_id) {
    passage_dt <- data.table::as.data.table(lineage_obj$passages[[passage_id]])
    if (!nrow(passage_dt)) return(NULL)

    keep_cols <- intersect(
      c("id", "filename", "field", "date", "areaOccupied_um2", "cellCount", "condition_label", "event", "passage"),
      names(passage_dt)
    )
    image_meta <- unique(passage_dt[, ..keep_cols], by = "filename")
    image_meta[, passage_id := as.character(passage_id)]
    image_meta[, lineage_id := as.character(lineage_obj$lineage_id)]
    image_meta[, passage_position := unname(passage_positions[passage_id])]
    image_meta[, image_key := sub("\\.[^.]+$", "", basename(filename))]
    image_meta
  })

  image_meta <- data.table::rbindlist(image_meta_list, fill = TRUE)
  if (!nrow(image_meta)) {
    stop(sprintf("No manifest rows found in lineage %s.", lineage_obj$lineage_id), call. = FALSE)
  }

  image_meta[, image_time := parse_mixed_datetime(date)]
  image_meta[, passage_start_time := suppressWarnings(min(image_time, na.rm = TRUE)), by = passage_id]
  image_meta[!is.finite(as.numeric(passage_start_time)), passage_start_time := as.POSIXct(NA, tz = "UTC")]

  # Treat the first observed image within a passage as the local seeding-time anchor.
  image_meta[, time_since_seeding_hours := as.numeric(difftime(image_time, passage_start_time, units = "hours"))]
  image_meta[, time_since_seeding_days := time_since_seeding_hours / 24]

  image_meta[, confluence := suppressWarnings(as.numeric(areaOccupied_um2))]
  image_meta[, cell_count := suppressWarnings(as.numeric(cellCount))]
  image_meta[, log_confluence := log1p(pmax(confluence, 0))]
  image_meta[, log_cell_count := log1p(pmax(cell_count, 0))]

  image_meta
}

choose_exemplar_lineages <- function(lineages, embedding_tbl, n = 3L) {
  coverage_tbl <- purrr::map_dfr(names(lineages), function(lineage_name) {
    lineage_obj <- lineages[[lineage_name]]
    meta <- build_lineage_image_meta(lineage_obj)
    tibble(
      lineage_id = as.character(lineage_obj$lineage_id),
      n_manifest_images = nrow(meta),
      n_matched_images = sum(meta$filename %in% embedding_tbl$filename)
    )
  })

  coverage_tbl %>%
    arrange(desc(n_matched_images), desc(n_manifest_images), lineage_id) %>%
    slice_head(n = n) %>%
    pull(lineage_id)
}

load_exemplar_panel <- function(lineage_rds, embedding_csv, exemplar_lineages = NULL) {
  lineages <- readRDS(project_path(lineage_rds))
  embedding_tbl <- data.table::fread(project_path(embedding_csv)) %>%
    as_tibble()

  embedding_cols <- embedding_column_names(embedding_tbl)
  if (!length(embedding_cols)) {
    stop("No embedding columns were found in the image-level embedding CSV.", call. = FALSE)
  }

  if (is.null(exemplar_lineages) || !length(exemplar_lineages)) {
    exemplar_lineages <- choose_exemplar_lineages(lineages, embedding_tbl, n = 3L)
  }

  lineage_meta <- purrr::map_dfr(exemplar_lineages, function(lineage_id) {
    build_lineage_image_meta(extract_lineage(lineages, lineage_id)) %>% as_tibble()
  })

  if (!nrow(lineage_meta)) {
    stop("No lineage metadata rows were recovered for the exemplar lineages.", call. = FALSE)
  }

  # Join on filename because the image-level embedding table is one row per image file.
  joined <- lineage_meta %>%
    inner_join(embedding_tbl, by = "filename")

  if (!nrow(joined)) {
    stop("No exemplar-lineage images matched the embedding CSV.", call. = FALSE)
  }

  joined %>%
    mutate(
      lineage_id = factor(lineage_id, levels = exemplar_lineages),
      passage_position = as.integer(passage_position)
    ) %>%
    distinct(lineage_id, passage_id, filename, .keep_all = TRUE)
}

fit_pc_association <- function(score, predictor, predictor_type = c("numeric", "factor")) {
  predictor_type <- match.arg(predictor_type)

  dat <- tibble(score = score, predictor = predictor) %>%
    filter(is.finite(score), !is.na(predictor))

  if (nrow(dat) < 8) {
    return(tibble(metric = NA_real_, n = nrow(dat)))
  }

  if (predictor_type == "numeric") {
    unique_n <- dplyr::n_distinct(dat$predictor)
    if (unique_n < 4) {
      fit <- lm(score ~ predictor, data = dat)
      out <- summary(fit)$adj.r.squared
    } else {
      k_basis <- max(4, min(8, unique_n - 1))
      fit <- mgcv::gam(score ~ s(predictor, k = k_basis), data = dat, method = "REML")
      out <- summary(fit)$r.sq
    }
  } else {
    if (dplyr::n_distinct(dat$predictor) < 2) {
      out <- NA_real_
    } else {
      fit <- lm(score ~ predictor, data = dat)
      out <- summary(fit)$adj.r.squared
    }
  }

  tibble(metric = out, n = nrow(dat))
}

weighted_pc_association <- function(score_tbl, score_cols, var_explained, predictor, predictor_type) {
  stats_tbl <- purrr::map_dfr(seq_along(score_cols), function(i) {
    fit_pc_association(score_tbl[[score_cols[[i]]]], predictor = predictor, predictor_type = predictor_type) %>%
      mutate(
        pc = score_cols[[i]],
        variance_weight = var_explained[[i]]
      )
  })

  tibble(
    weighted_r2 = weighted.mean(stats_tbl$metric, w = stats_tbl$variance_weight, na.rm = TRUE),
    mean_pc_r2 = mean(stats_tbl$metric, na.rm = TRUE),
    n_pcs = length(score_cols)
  )
}

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || diff(rng) <= 0) {
    return(rep(NA_real_, length(x)))
  }
  (x - rng[1]) / diff(rng)
}

nearest_other_lineage_distance <- function(mat, lineage) {
  out <- rep(NA_real_, nrow(mat))

  for (i in seq_len(nrow(mat))) {
    keep <- lineage != lineage[[i]]
    if (!any(keep)) next
    d <- sqrt(rowSums((t(t(mat[keep, , drop = FALSE]) - mat[i, ]))^2))
    out[[i]] <- min(d, na.rm = TRUE)
  }

  out
}

predict_time_to_stage <- function(df, lineage_levels, stage_grid) {
  purrr::map_dfr(lineage_levels, function(lineage_id) {
    dat <- df %>%
      filter(lineage_id == .env$lineage_id) %>%
      filter(is.finite(time_since_seeding_days), is.finite(latent_score))

    if (nrow(dat) < 8 || dplyr::n_distinct(dat$time_since_seeding_days) < 4) {
      return(tibble())
    }

    fit <- mgcv::gam(
      latent_score ~ s(time_since_seeding_days, k = min(8, dplyr::n_distinct(dat$time_since_seeding_days) - 1)),
      data = dat,
      method = "REML"
    )

    time_grid <- seq(
      min(dat$time_since_seeding_days, na.rm = TRUE),
      max(dat$time_since_seeding_days, na.rm = TRUE),
      length.out = 200
    )

    pred_tbl <- tibble(
      time_since_seeding_days = time_grid,
      latent_score = as.numeric(predict(fit, newdata = tibble(time_since_seeding_days = time_grid)))
    )

    purrr::map_dfr(stage_grid, function(target_score) {
      hit <- pred_tbl %>%
        mutate(distance = abs(latent_score - target_score)) %>%
        slice_min(order_by = distance, n = 1, with_ties = FALSE)

      tibble(
        lineage_id = lineage_id,
        stage_score = target_score,
        estimated_time_days = hit$time_since_seeding_days
      )
    })
  })
}

fit_lineage_trajectory_grid <- function(df, lineage_levels, pc_x = "PC1", pc_y = "PC2", n_points = 100) {
  purrr::map_dfr(lineage_levels, function(lineage_id) {
    dat <- df %>%
      filter(lineage_id == .env$lineage_id) %>%
      filter(is.finite(time_since_seeding_days))

    if (nrow(dat) < 8 || dplyr::n_distinct(dat$time_since_seeding_days) < 4) {
      return(tibble())
    }

    time_grid <- seq(
      min(dat$time_since_seeding_days, na.rm = TRUE),
      max(dat$time_since_seeding_days, na.rm = TRUE),
      length.out = n_points
    )

    fit_x <- mgcv::gam(
      reformulate(sprintf("s(time_since_seeding_days, k = %d)", min(8, dplyr::n_distinct(dat$time_since_seeding_days) - 1)), response = pc_x),
      data = dat,
      method = "REML"
    )
    fit_y <- mgcv::gam(
      reformulate(sprintf("s(time_since_seeding_days, k = %d)", min(8, dplyr::n_distinct(dat$time_since_seeding_days) - 1)), response = pc_y),
      data = dat,
      method = "REML"
    )

    tibble(
      lineage_id = lineage_id,
      time_since_seeding_days = time_grid,
      x = as.numeric(predict(fit_x, newdata = tibble(time_since_seeding_days = time_grid))),
      y = as.numeric(predict(fit_y, newdata = tibble(time_since_seeding_days = time_grid))),
      view = paste(pc_x, pc_y, sep = "_")
    )
  })
}

infer_control_lineage <- function(df, requested_control = NULL) {
  lineage_levels <- levels(df$lineage_id) %||% unique(as.character(df$lineage_id))
  if (!is.null(requested_control) && nzchar(requested_control)) {
    return(as.character(requested_control))
  }

  if ("condition_label" %in% names(df)) {
    cond_tbl <- df %>%
      mutate(condition_label = ifelse(is.na(condition_label), "", as.character(condition_label))) %>%
      group_by(lineage_id) %>%
      summarise(
        control_label_score = mean(grepl("untreated|control|normox|normoxia", tolower(condition_label)), na.rm = TRUE),
        .groups = "drop"
      ) %>%
      arrange(desc(control_label_score), lineage_id)
    if (nrow(cond_tbl) && is.finite(cond_tbl$control_label_score[[1]]) && cond_tbl$control_label_score[[1]] > 0) {
      return(as.character(cond_tbl$lineage_id[[1]]))
    }
  }

  sampling_tbl <- df %>%
    filter(is.finite(time_since_seeding_days)) %>%
    mutate(time_bin = round(time_since_seeding_days, 2)) %>%
    dplyr::count(lineage_id, passage_id, time_bin, name = "n_images") %>%
    dplyr::count(lineage_id, passage_id, name = "n_timepoints") %>%
    group_by(lineage_id) %>%
    summarise(
      median_timepoints_per_passage = median(n_timepoints, na.rm = TRUE),
      n_images = n(),
      .groups = "drop"
    ) %>%
    arrange(median_timepoints_per_passage, n_images, lineage_id)

  as.character(sampling_tbl$lineage_id[[1]])
}

nn_match_between_groups <- function(query_tbl, target_tbl, feature_cols) {
  if (!nrow(query_tbl) || !nrow(target_tbl)) {
    return(tibble(
      query_row = integer(),
      nn_index = integer(),
      nn_distance = double()
    ))
  }

  query_mat <- as.matrix(query_tbl[, feature_cols])
  target_mat <- as.matrix(target_tbl[, feature_cols])

  out <- vector("list", nrow(query_tbl))
  for (i in seq_len(nrow(query_tbl))) {
    d <- sqrt(rowSums((t(t(target_mat) - query_mat[i, ]))^2))
    j <- which.min(d)
    out[[i]] <- tibble(
      query_row = i,
      nn_index = j,
      nn_distance = d[[j]]
    )
  }

  bind_rows(out)
}

curvature_diagnostic_tbl <- function(df) {
  linear_pc2 <- lm(PC2 ~ PC1, data = df)
  smooth_pc2 <- mgcv::gam(PC2 ~ s(PC1, k = min(8, max(4, dplyr::n_distinct(round(df$PC1, 3)) - 1))), data = df, method = "REML")

  out <- tibble(
    linear_r2_pc2_on_pc1 = summary(linear_pc2)$adj.r.squared,
    smooth_r2_pc2_on_pc1 = summary(smooth_pc2)$r.sq,
    nonlinear_gain_pc2_on_pc1 = summary(smooth_pc2)$r.sq - summary(linear_pc2)$adj.r.squared
  )

  if ("PC3" %in% names(df)) {
    linear_pc3 <- lm(PC3 ~ PC1, data = df)
    smooth_pc3 <- mgcv::gam(PC3 ~ s(PC1, k = min(8, max(4, dplyr::n_distinct(round(df$PC1, 3)) - 1))), data = df, method = "REML")
    out <- out %>%
      mutate(
        linear_r2_pc3_on_pc1 = summary(linear_pc3)$adj.r.squared,
        smooth_r2_pc3_on_pc1 = summary(smooth_pc3)$r.sq,
        nonlinear_gain_pc3_on_pc1 = summary(smooth_pc3)$r.sq - summary(linear_pc3)$adj.r.squared
      )
  }

  out
}
```

## Load the Exemplar Panel

Start by loading the image-level embedding table and reconstructing metadata for three exemplar lineages from `lineages.Rds`.

If `params$exemplar_lineages` is left empty, the notebook will pick the three lineages with the most matched embedded images. That is only a fallback for convenience; for serious review, it is better to set the intended lineages explicitly.

```{r load-data}
panel_tbl <- load_exemplar_panel(
  lineage_rds = params$lineage_rds,
  embedding_csv = params$embedding_csv,
  exemplar_lineages = params$exemplar_lineages
)

embedding_cols <- embedding_column_names(panel_tbl)

analysis_tbl <- panel_tbl %>%
  mutate(
    lineage_id = factor(lineage_id, levels = unique(lineage_id)),
    passage_label = paste0("P", passage_position)
  )

lineage_levels <- levels(analysis_tbl$lineage_id)

analysis_overview <- tibble(
  n_images = nrow(analysis_tbl),
  n_lineages = dplyr::n_distinct(analysis_tbl$lineage_id),
  n_passages = dplyr::n_distinct(interaction(analysis_tbl$lineage_id, analysis_tbl$passage_id, drop = TRUE)),
  n_embedding_dims = length(embedding_cols),
  exemplar_lineages = paste(lineage_levels, collapse = ", ")
)

knitr::kable(analysis_overview)
```

Interpretation:

- this table is the quick sanity check that the notebook is operating on image-level rows
- if the selected lineages are not the intended exemplars, set `params$exemplar_lineages` explicitly before interpreting any downstream plot

## Basic QC and Coverage

Before doing any dimensionality reduction, check missingness, numeric ranges, and whether coverage is balanced enough to support a trajectory comparison.

```{r qc-summary-table}
qc_tbl <- analysis_tbl %>%
  summarise(
    across(
      c(time_since_seeding_days, confluence, cell_count, all_of(embedding_cols)),
      ~ mean(is.na(.x))
    )
  ) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_fraction") %>%
  arrange(desc(missing_fraction))

knitr::kable(head(qc_tbl, 12), digits = 3)
```

The next table summarizes each lineage separately.

```{r lineage-qc}
lineage_qc_tbl <- analysis_tbl %>%
  group_by(lineage_id) %>%
  summarise(
    n_images = n(),
    n_passages = n_distinct(passage_id),
    time_min_days = min(time_since_seeding_days, na.rm = TRUE),
    time_max_days = max(time_since_seeding_days, na.rm = TRUE),
    median_confluence = median(confluence, na.rm = TRUE),
    median_cell_count = median(cell_count, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(lineage_qc_tbl, digits = 2)
```

Interpretation:

- large differences in `n_images` or `time_max_days` are acceptable, but they matter when comparing smooths
- if one lineage is concentrated in only a narrow time window, that lineage can still inform overlap in state space but not broad alignment

Coverage plots are the first place to see whether irregular sampling or passage imbalance could create spurious apparent trajectories.

```{r coverage-plots}
coverage_tbl <- analysis_tbl %>%
  dplyr::count(lineage_id, passage_position, time_since_seeding_days, name = "n_images")

ggplot(coverage_tbl, aes(x = time_since_seeding_days, y = factor(passage_position), size = n_images, color = lineage_id)) +
  geom_point(alpha = 0.8) +
  scale_size_continuous(range = c(2.5, 7)) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Sampling coverage by lineage, passage, and time since seeding",
    x = "Time since seeding (days)",
    y = "Passage position",
    color = "Lineage",
    size = "Images"
  )
```

```{r time-density}
ggplot(analysis_tbl, aes(x = time_since_seeding_days, fill = lineage_id)) +
  geom_histogram(binwidth = 0.5, position = "identity", alpha = 0.45, color = "white") +
  facet_wrap(~ lineage_id, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)], guide = "none") +
  labs(
    title = "Time-since-seeding coverage within each lineage",
    x = "Time since seeding (days)",
    y = "Image count"
  )
```

Interpretation:

- these plots should be read before trusting any smooth trend
- if one lineage has strong gaps in time or is represented mainly by one passage, apparent rate differences may just reflect missing intermediate states

## Covariate Checks

The embeddings may capture density-related biology, but they may also be dominated by simple occupancy effects. These direct covariate plots are therefore both biological summaries and sanity checks.

```{r covariate-pairs}
covariate_long <- analysis_tbl %>%
  select(lineage_id, passage_position, time_since_seeding_days, confluence, cell_count) %>%
  pivot_longer(
    cols = c(confluence, cell_count),
    names_to = "metric",
    values_to = "value"
  )

ggplot(covariate_long, aes(x = time_since_seeding_days, y = value, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Confluence and cell count versus time since seeding",
    x = "Time since seeding (days)",
    y = NULL,
    color = "Lineage"
  )
```

Interpretation:

- if `confluence` and `cell_count` are nearly deterministic functions of time in every lineage, then a strong temporal signal in the embeddings could still be largely a density signal
- if lineages differ in density at the same time, that creates a useful stress test for whether embeddings are capturing something beyond raw time

## Preprocess the Embeddings

The embedding dimensions are centered and scaled before PCA. This prevents large-variance coordinates from dominating the first PCs for purely numeric reasons.

```{r pca}
complete_tbl <- analysis_tbl %>%
  filter(
    if_all(all_of(embedding_cols), is.finite),
    is.finite(time_since_seeding_days)
  )

embedding_mat <- as.matrix(complete_tbl[, embedding_cols])
storage.mode(embedding_mat) <- "double"

# Scale each embedding dimension so PCA reflects structure rather than raw scale.
pca_fit <- prcomp(embedding_mat, center = TRUE, scale. = TRUE)

n_top_pcs <- min(params$n_top_pcs, ncol(pca_fit$x))
score_cols <- paste0("PC", seq_len(n_top_pcs))
score_tbl <- bind_cols(
  complete_tbl %>% select(-all_of(embedding_cols)),
  as_tibble(pca_fit$x[, seq_len(n_top_pcs), drop = FALSE], .name_repair = "minimal")
)
names(score_tbl)[(ncol(score_tbl) - n_top_pcs + 1):ncol(score_tbl)] <- score_cols

var_explained <- (pca_fit$sdev^2) / sum(pca_fit$sdev^2)
var_tbl <- tibble(
  pc = paste0("PC", seq_along(var_explained)),
  variance_explained = var_explained,
  cumulative_variance = cumsum(var_explained)
)
```

The variance-explained profile tells us how concentrated or diffuse image-state variation is.

```{r variance-explained}
ggplot(var_tbl %>% slice_head(n = 15), aes(x = reorder(pc, variance_explained), y = variance_explained)) +
  geom_col(fill = "#2C7FB8") +
  geom_text(aes(label = percent(variance_explained, accuracy = 0.1)), hjust = -0.05, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Variance explained by the leading principal components",
    x = NULL,
    y = "Fraction of total embedding variance"
  )
```

Interpretation:

- if PC1 is clearly dominant and its biology is sensible, it becomes a strong candidate latent progression axis
- if variance is spread across many PCs, shared progression may still exist, but it is less likely to be captured by a single simple coordinate

Loading summaries are kept compact here: the goal is to identify whether a few embedding dimensions dominate each PC.

```{r loading-summaries}
loading_tbl <- as_tibble(pca_fit$rotation[, seq_len(min(4, ncol(pca_fit$rotation))), drop = FALSE], rownames = "embedding_dim") %>%
  pivot_longer(-embedding_dim, names_to = "pc", values_to = "loading") %>%
  group_by(pc) %>%
  slice_max(order_by = abs(loading), n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(pc, desc(abs(loading)))

knitr::kable(loading_tbl, digits = 3)
```

Interpretation:

- these are not biologically interpretable features by themselves, but they help detect whether a PC is broad-based or dominated by a small corner of the embedding vector

## PCA Views of Image State

PCA is the main basis of inference in this notebook. The first view asks whether the three lineages occupy overlapping or distinct regions of low-dimensional embedding space.

```{r pca-by-lineage}
ggplot(score_tbl, aes(x = PC1, y = PC2, color = lineage_id)) +
  geom_point(alpha = 0.8, size = 2.2) +
  stat_ellipse(linewidth = 0.7, alpha = 0.4) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "PCA state space colored by lineage",
    x = "PC1",
    y = "PC2",
    color = "Lineage"
  )
```

Interpretation:

- strong overlap is the first qualitative sign that a shared trajectory may exist
- cleanly separated lineage clouds would instead suggest lineage-specific morphology or strong uncorrected batch structure

The next views ask whether the same PCA space orders images by time and by density-related covariates.

```{r pca-color-views}
plot_by_numeric <- function(var_name, title_text) {
  ggplot(score_tbl, aes(x = PC1, y = PC2, color = .data[[var_name]])) +
    geom_point(size = 2.1, alpha = 0.85) +
    scale_color_viridis_c(option = "C", na.value = "grey80") +
    labs(
      title = title_text,
      x = "PC1",
      y = "PC2",
      color = var_name
    )
}

plot_by_numeric("time_since_seeding_days", "PCA state space colored by time since seeding")
plot_by_numeric("confluence", "PCA state space colored by confluence")
plot_by_numeric("cell_count", "PCA state space colored by cell count")
```

Interpretation:

- a smooth color gradient along the same geometric axis in all three plots suggests a shared dominant progression
- if the time gradient and density gradients point in different directions, that is useful evidence that morphological state is not reducible to a single occupancy proxy

## Quantify Association Between Embeddings and Key Covariates

To summarize which variables explain the major axes of embedding variation, the notebook computes variance-weighted per-PC association scores across the leading PCs.

This is intentionally simple: it is meant as an interpretable first-pass summary, not a formal variance-partitioning model.

```{r pc-association-summary}
association_tbl <- bind_rows(
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$time_since_seeding_days, "numeric") %>%
    mutate(variable = "time_since_seeding"),
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$confluence, "numeric") %>%
    mutate(variable = "confluence"),
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$cell_count, "numeric") %>%
    mutate(variable = "cell_count"),
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$lineage_id, "factor") %>%
    mutate(variable = "lineage"),
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$passage_position, "numeric") %>%
    mutate(variable = "passage_position")
) %>%
  select(variable, weighted_r2, mean_pc_r2, n_pcs)

knitr::kable(association_tbl, digits = 3)
```

```{r pc-association-plot}
ggplot(association_tbl, aes(x = reorder(variable, weighted_r2), y = weighted_r2)) +
  geom_col(fill = "#21618C") +
  geom_text(aes(label = round(weighted_r2, 2)), hjust = -0.05, size = 3.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Variance-weighted association of leading PCs with key variables",
    x = NULL,
    y = "Weighted R-squared across leading PCs"
  )
```

Interpretation:

- if `time_since_seeding` is strongest but `confluence` and `cell_count` are also strong, the embedding is likely tracking both temporal progression and density-related biology
- if `lineage` dominates, shared alignment becomes less plausible
- if `passage_position` remains strong after time, that suggests slow lineage history effects beyond within-passage seeding time

## PC Scores Versus Time and Density Covariates

The next plots examine the leading PCs directly. These are now used as explicit diagnostics for whether a straight-axis progression is adequate or whether the trajectory is genuinely curved in low-dimensional space.

```{r pc-trends}
pc_long <- score_tbl %>%
  select(lineage_id, passage_position, time_since_seeding_days, confluence, cell_count, all_of(score_cols[seq_len(min(4, length(score_cols)))])) %>%
  pivot_longer(cols = starts_with("PC"), names_to = "pc", values_to = "score")

ggplot(pc_long, aes(x = time_since_seeding_days, y = score, color = lineage_id)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_wrap(~ pc, scales = "free_y", ncol = 2) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Leading PC scores versus time since seeding",
    x = "Time since seeding (days)",
    y = "PC score",
    color = "Lineage"
  )
```

```{r pc-vs-covariates}
pc_cov_long <- score_tbl %>%
  select(lineage_id, any_of(c("PC1", "PC2", "PC3")), confluence, cell_count) %>%
  pivot_longer(cols = starts_with("PC"), names_to = "pc", values_to = "score") %>%
  pivot_longer(cols = c(confluence, cell_count), names_to = "covariate", values_to = "value")

ggplot(pc_cov_long, aes(x = value, y = score, color = lineage_id)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_grid(pc ~ covariate, scales = "free_x") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Leading PCs versus confluence and cell count",
    x = NULL,
    y = "PC score",
    color = "Lineage"
  )
```

Interpretation:

- if time is expressed mainly along one straight PC axis, then `PC1` can remain a reasonable baseline summary
- if multiple PCs change smoothly and jointly, especially with turning structure in `PC1-PC2` or `PC1-PC2-PC3`, then a curved trajectory is more appropriate than `PC1` alone

## Curvature Diagnostics in PCA Space

These views are the main visual test of whether a curved latent trajectory is warranted.

```{r pca-curvature-views}
pc_view_tbl <- bind_rows(
  score_tbl %>%
    transmute(lineage_id, passage_position, time_since_seeding_days, confluence, cell_count, view = "PC1 vs PC2", x = PC1, y = PC2),
  score_tbl %>%
    mutate(PC3 = if ("PC3" %in% names(score_tbl)) PC3 else NA_real_) %>%
    transmute(lineage_id, passage_position, time_since_seeding_days, confluence, cell_count, view = "PC1 vs PC3", x = PC1, y = PC3)
)

plot_pc_view <- function(color_var, title_text, scale_fn = scale_color_viridis_c(option = "C", na.value = "grey80")) {
  ggplot(pc_view_tbl, aes(x = x, y = y, color = .data[[color_var]])) +
    geom_point(size = 2, alpha = 0.85) +
    facet_wrap(~ view, scales = "free") +
    scale_fn +
    labs(
      title = title_text,
      x = NULL,
      y = NULL,
      color = color_var
    )
}

plot_pc_view("time_since_seeding_days", "PC-space views colored by time since seeding")
plot_pc_view("confluence", "PC-space views colored by confluence")
plot_pc_view("cell_count", "PC-space views colored by cell count")

ggplot(pc_view_tbl, aes(x = x, y = y, color = lineage_id)) +
  geom_point(size = 2, alpha = 0.85) +
  facet_wrap(~ view, scales = "free") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "PC-space views colored by lineage",
    x = NULL,
    y = NULL,
    color = "Lineage"
  )

ggplot(pc_view_tbl, aes(x = x, y = y, color = passage_position)) +
  geom_point(size = 2, alpha = 0.85) +
  facet_wrap(~ view, scales = "free") +
  scale_color_viridis_c(option = "D", na.value = "grey80") +
  labs(
    title = "PC-space views colored by passage position",
    x = NULL,
    y = NULL,
    color = "Passage"
  )
```

```{r smooth-trajectories}
trajectory_grid_12 <- fit_lineage_trajectory_grid(score_tbl, lineage_levels, pc_x = "PC1", pc_y = "PC2")
trajectory_grid_13 <- if ("PC3" %in% names(score_tbl)) {
  fit_lineage_trajectory_grid(score_tbl, lineage_levels, pc_x = "PC1", pc_y = "PC3")
} else {
  tibble()
}

trajectory_grid <- bind_rows(trajectory_grid_12, trajectory_grid_13)

ggplot(pc_view_tbl, aes(x = x, y = y, color = lineage_id)) +
  geom_point(alpha = 0.3, size = 1.8) +
  geom_path(data = trajectory_grid, linewidth = 1.15, lineend = "round") +
  facet_wrap(~ view, scales = "free") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Lineage-specific smooth trajectories in PCA space",
    subtitle = "These curves summarize how each lineage moves through the leading PCs over time",
    x = NULL,
    y = NULL,
    color = "Lineage"
  )
```

```{r curvature-summary}
curve_diag_tbl <- curvature_diagnostic_tbl(score_tbl)
knitr::kable(curve_diag_tbl, digits = 3)
```

Interpretation:

- a curved trajectory is justified when the time gradient bends visibly through `PC1-PC2` or `PC1-PC2-PC3`, and when smooth fits of `PC2` or `PC3` as functions of `PC1` outperform a straight-line description
- if those gains are trivial and the point cloud is effectively collinear, `PC1` remains a fair summary baseline

## Secondary Nonlinear View

UMAP is included only as a secondary visualization. It can help reveal local neighborhoods, but it is not used as the main basis of inference.

```{r umap, eval=has_uwot}
set.seed(1)

umap_fit <- uwot::umap(
  embedding_mat,
  n_neighbors = min(20, nrow(embedding_mat) - 1),
  min_dist = 0.25,
  metric = "euclidean",
  scale = FALSE,
  verbose = FALSE
)

umap_tbl <- score_tbl %>%
  mutate(
    UMAP1 = umap_fit[, 1],
    UMAP2 = umap_fit[, 2]
  )

ggplot(umap_tbl, aes(x = UMAP1, y = UMAP2, color = lineage_id)) +
  geom_point(alpha = 0.8, size = 2.1) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Secondary UMAP view of image state",
    subtitle = "Use as a neighborhood view, not as the primary trajectory basis",
    x = "UMAP1",
    y = "UMAP2",
    color = "Lineage"
  )
```

```{r umap-note, eval=!has_uwot, results='asis'}
cat("`uwot` is not installed, so the secondary UMAP view is skipped.\n")
```

Interpretation:

- if UMAP preserves the same broad lineage overlap already seen in PCA, that increases confidence that the overlap is not an artifact of one linear projection
- if UMAP produces lineage islands that are absent in PCA, inspect carefully rather than over-interpreting either view

## Do the Lineages Occupy Similar Regions of State Space?

A shared trajectory requires at least partial overlap in low-dimensional state space. As a simple summary, the next plot and table look at cross-lineage proximity in the first three PCs.

```{r overlap-diagnostics}
pc_overlap_cols <- c("PC1", "PC2", score_cols[min(3, length(score_cols))])
pc_overlap_cols <- unique(pc_overlap_cols)
pc_overlap_mat <- as.matrix(score_tbl[, pc_overlap_cols])
score_tbl$nearest_other_lineage_dist <- nearest_other_lineage_distance(pc_overlap_mat, score_tbl$lineage_id)

ggplot(score_tbl, aes(x = lineage_id, y = nearest_other_lineage_dist, fill = lineage_id)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.35) +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)], guide = "none") +
  labs(
    title = "Distance to the nearest image from a different lineage",
    subtitle = "Smaller values imply stronger cross-lineage overlap in low-dimensional state space",
    x = "Lineage",
    y = "Nearest other-lineage distance in PC space"
  )
```

```{r overlap-summary}
overlap_summary_tbl <- score_tbl %>%
  group_by(lineage_id) %>%
  summarise(
    median_nearest_other_lineage_dist = median(nearest_other_lineage_dist, na.rm = TRUE),
    q75_nearest_other_lineage_dist = quantile(nearest_other_lineage_dist, probs = 0.75, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(overlap_summary_tbl, digits = 3)
```

Interpretation:

- low cross-lineage nearest-neighbor distances support the idea that different lineages revisit similar image states
- very large distances would argue for lineage-specific trajectories even if each lineage changes smoothly on its own

## Derive a Refined Latent Progression Score

The straight-axis baseline is still useful, but the notebook now defines one explicit main latent variable:

- baseline comparator: oriented `PC1`
- main latent trajectory: arc length along one shared principal curve fit in `PC1-PC2` across all lineages

This keeps the trajectory interpretable and visible in the same 2D PCA space used throughout the notebook. `PC3` remains a curvature diagnostic, but the default staging variable is the shared `PC1-PC2` curve score rather than `PC1` alone.

```{r latent-score}
pc1_oriented <- score_tbl$PC1
if (stats::cor(pc1_oriented, score_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman") < 0) {
  pc1_oriented <- -pc1_oriented
}

curve_feature_cols <- c("PC1", "PC2")

latent_tbl <- score_tbl %>%
  mutate(
    latent_score_pc1 = rescale01(pc1_oriented),
    latent_score = latent_score_pc1,
    latent_arc_length = latent_score_pc1,
    distance_to_curve = NA_real_,
    latent_score_method = "PC1_baseline",
    curve_basis = "PC1"
  )

curve_diag_tbl <- curve_diag_tbl %>%
  mutate(curved_trajectory_warranted = FALSE)

curve_path_tbl <- tibble(
  PC1 = numeric(),
  PC2 = numeric(),
  lambda = numeric()
)
curve_projection_tbl <- tibble(
  filename = character(),
  lineage_id = factor(),
  PC1 = numeric(),
  PC2 = numeric(),
  curve_PC1 = numeric(),
  curve_PC2 = numeric(),
  lambda = numeric(),
  distance_to_curve = numeric()
)

if (has_princurve) {
  curve_input <- as.matrix(score_tbl[, curve_feature_cols])
  principal_curve_fit <- princurve::principal_curve(curve_input)

  curve_lambda <- principal_curve_fit$lambda
  if (stats::cor(curve_lambda, score_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman") < 0) {
    curve_lambda <- -curve_lambda
  }

  curve_lambda <- curve_lambda - min(curve_lambda, na.rm = TRUE)
  curve_lambda01 <- rescale01(curve_lambda)
  curve_distance <- principal_curve_fit$dist

  path_length_ratio <- sum(sqrt(rowSums(diff(principal_curve_fit$s[order(curve_lambda), , drop = FALSE])^2)), na.rm = TRUE) /
    sqrt(sum((principal_curve_fit$s[which.max(curve_lambda), ] - principal_curve_fit$s[which.min(curve_lambda), ])^2))

  curve_vs_pc1 <- stats::cor(curve_lambda01, latent_tbl$latent_score_pc1, use = "pairwise.complete.obs", method = "spearman")
  curve_vs_time <- stats::cor(curve_lambda01, score_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman")

  curved_trajectory_warranted <- is.finite(path_length_ratio) &&
    (
      path_length_ratio > 1.08 ||
      (("nonlinear_gain_pc2_on_pc1" %in% names(curve_diag_tbl)) && curve_diag_tbl$nonlinear_gain_pc2_on_pc1[[1]] > 0.05)
    )

  latent_tbl <- latent_tbl %>%
    mutate(
      latent_score_curve = curve_lambda01,
      latent_arc_length_curve = curve_lambda,
      distance_to_curve = curve_distance
    )

  if (curved_trajectory_warranted) {
    latent_tbl <- latent_tbl %>%
      mutate(
        latent_score = latent_score_curve,
        latent_arc_length = latent_arc_length_curve,
        latent_score_method = "shared_principal_curve_arc_length",
        curve_basis = "PC1 + PC2"
      )
  }

  curve_path_tbl <- tibble(
    PC1 = principal_curve_fit$s[, 1],
    PC2 = principal_curve_fit$s[, 2],
    lambda = curve_lambda
  ) %>%
    arrange(lambda)

  curve_projection_tbl <- latent_tbl %>%
    transmute(
      filename,
      lineage_id,
      PC1,
      PC2,
      curve_PC1 = principal_curve_fit$s[, 1],
      curve_PC2 = principal_curve_fit$s[, 2],
      lambda = curve_lambda,
      distance_to_curve = curve_distance
    )

  curve_diag_tbl <- curve_diag_tbl %>%
    mutate(
      path_length_ratio = path_length_ratio,
      spearman_curve_vs_pc1 = curve_vs_pc1,
      spearman_curve_vs_time = curve_vs_time,
      curved_trajectory_warranted = curved_trajectory_warranted
    )
} else {
  latent_tbl <- latent_tbl %>%
    mutate(
      latent_score_curve = NA_real_,
      latent_arc_length_curve = NA_real_
    )
}

# Normalize the curve table schema so downstream plots never see a malformed object.
if (!exists("curve_path_tbl") ||
    !is.data.frame(curve_path_tbl) ||
    !all(c("PC1", "PC2", "lambda") %in% names(curve_path_tbl))) {
  curve_path_tbl <- tibble(
    PC1 = numeric(),
    PC2 = numeric(),
    lambda = numeric()
  )
}

if (!exists("curve_projection_tbl") ||
    !is.data.frame(curve_projection_tbl) ||
    !all(c("PC1", "PC2", "curve_PC1", "curve_PC2", "lambda", "distance_to_curve") %in% names(curve_projection_tbl))) {
  curve_projection_tbl <- tibble(
    filename = character(),
    lineage_id = factor(),
    PC1 = numeric(),
    PC2 = numeric(),
    curve_PC1 = numeric(),
    curve_PC2 = numeric(),
    lambda = numeric(),
    distance_to_curve = numeric()
  )
}

latent_method_tbl <- tibble(
  chosen_method = unique(latent_tbl$latent_score_method),
  curve_basis = unique(latent_tbl$curve_basis),
  spearman_pc1_vs_time = cor(latent_tbl$latent_score_pc1, latent_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman"),
  spearman_curve_vs_time = if ("latent_score_curve" %in% names(latent_tbl)) cor(latent_tbl$latent_score_curve, latent_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman") else NA_real_,
  spearman_main_latent_vs_time = cor(latent_tbl$latent_score, latent_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman"),
  median_distance_to_curve = median(latent_tbl$distance_to_curve, na.rm = TRUE)
)

knitr::kable(latent_method_tbl, digits = 3)
knitr::kable(curve_diag_tbl, digits = 3)
```

Interpretation:

- the main staging variable used below is `latent_tbl$latent_score`
- if the chosen method is `shared_principal_curve_arc_length`, then every downstream alignment and stage plot is based on distance along one shared curved path in `PC1-PC2`
- the main diagnostics to watch are the path-length ratio, the nonlinear gain in `PC2`, and the distance-to-curve distribution

Mapping rule used here:

- each image is projected to the nearest point on the fitted shared curve in `PC1/PC2`
- `distance_to_curve` is the Euclidean distance from the observed point to that projection
- `latent_score` is rescaled arc length (`lambda`) of that projected point along the curve

```{r shared-curve-in-pca}
shared_curve_plot <- ggplot(latent_tbl, aes(x = PC1, y = PC2, color = latent_score)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "C", na.value = "grey80") +
  labs(
    title = "Shared curved trajectory in PC1-PC2 space",
    subtitle = paste("Main latent score:", unique(latent_tbl$latent_score_method)),
    x = "PC1",
    y = "PC2",
    color = "Latent score"
  )

if (exists("curve_path_tbl") &&
    is.data.frame(curve_path_tbl) &&
    all(c("PC1", "PC2") %in% names(curve_path_tbl)) &&
    nrow(curve_path_tbl) >= 2) {
  shared_curve_plot <- shared_curve_plot +
    geom_path(
      data = curve_path_tbl,
      mapping = aes(x = PC1, y = PC2),
      inherit.aes = FALSE,
      linewidth = 1.3,
      color = "black"
    )
}

shared_curve_plot
```

```{r curve-projection-diagnostic}
projection_plot_tbl <- curve_projection_tbl %>%
  filter(is.finite(distance_to_curve))

if (nrow(projection_plot_tbl) > 80) {
  projection_plot_tbl <- bind_rows(
    projection_plot_tbl %>% slice_min(order_by = distance_to_curve, n = 40, with_ties = FALSE),
    projection_plot_tbl %>% slice_max(order_by = distance_to_curve, n = 40, with_ties = FALSE)
  ) %>%
    distinct(filename, .keep_all = TRUE)
}

if (nrow(curve_path_tbl) >= 2 && nrow(projection_plot_tbl) >= 1) {
  ggplot() +
    geom_path(data = curve_path_tbl, aes(x = PC1, y = PC2), linewidth = 1.2, color = "black") +
    geom_segment(
      data = projection_plot_tbl,
      aes(x = PC1, y = PC2, xend = curve_PC1, yend = curve_PC2, color = lineage_id),
      alpha = 0.25,
      linewidth = 0.4
    ) +
    geom_point(data = projection_plot_tbl, aes(x = PC1, y = PC2, color = lineage_id), alpha = 0.75, size = 1.8) +
    geom_point(data = projection_plot_tbl, aes(x = curve_PC1, y = curve_PC2), color = "black", alpha = 0.5, size = 1) +
    scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
    labs(
      title = "Projection of images onto the shared curve",
      subtitle = "Segments show Euclidean projection in PC1/PC2; black points are the assigned closest points on the curve",
      x = "PC1",
      y = "PC2",
      color = "Lineage"
    )
} else {
  cat("Projection diagnostic skipped because the shared curve was not fit.\n")
}
```

## Straight-Axis Versus Curved Progression

This section makes the comparison explicit rather than assuming the curve is automatically better.

```{r straight-vs-curved}
latent_compare_long <- latent_tbl %>%
  select(lineage_id, time_since_seeding_days, latent_score_pc1, latent_score_curve, distance_to_curve) %>%
  pivot_longer(cols = c(latent_score_pc1, latent_score_curve), names_to = "score_type", values_to = "score") %>%
  filter(is.finite(score))

ggplot(latent_compare_long, aes(x = time_since_seeding_days, y = score, color = lineage_id)) +
  geom_point(alpha = 0.65, size = 1.8) +
  geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6), linewidth = 0.85) +
  facet_wrap(~ score_type, scales = "free_y") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Straight-axis versus curved latent progression",
    subtitle = "Use this to judge whether the curved score resolves structure that PC1 compresses",
    x = "Time since seeding (days)",
    y = "Latent score",
    color = "Lineage"
  )
```

```{r distance-to-curve}
distance_plot_tbl <- latent_tbl %>%
  filter(is.finite(distance_to_curve))

if (nrow(distance_plot_tbl) >= 2 && dplyr::n_distinct(distance_plot_tbl$distance_to_curve) >= 2) {
  ggplot(distance_plot_tbl, aes(x = distance_to_curve, fill = lineage_id)) +
    geom_histogram(alpha = 0.45, position = "identity", bins = 30) +
    facet_wrap(~ lineage_id, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)], guide = "none") +
    labs(
      title = "Distance to the fitted curved trajectory",
      subtitle = "Images far from the curve may represent noise, off-trajectory states, or geometry not captured by one path",
      x = "Distance to principal curve",
      y = "Image count"
    )
} else {
  cat("Distance-to-curve histogram skipped because there are too few finite distinct values to bin safely.\n")
}
```

Interpretation:

- the curved score is worthwhile when it separates images that `PC1` compresses, especially in the central part of the trajectory
- a useful principal curve should produce moderate distances for most images rather than fitting only a thin subset of the cloud

## Latent Progression Score Versus Time Since Seeding

This is the key first-pass alignment plot: it asks whether the same latent state can appear at different absolute times in different lineages.

```{r latent-vs-time}
ggplot(latent_tbl, aes(x = time_since_seeding_days, y = latent_score, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6), linewidth = 0.9) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Latent progression score versus time since seeding",
    subtitle = paste("Chosen score:", unique(latent_tbl$latent_score_method)),
    x = "Time since seeding (days)",
    y = "Latent progression score",
    color = "Lineage"
  )
```

Interpretation:

- parallel but shifted curves suggest similar morphological states reached at different times
- sharply different curve shapes suggest lineage-specific progressions rather than simple rate differences

## Latent Progression Score Versus Confluence and Cell Count

These plots show how much the inferred progression score tracks density-related covariates.

```{r latent-vs-density}
latent_cov_long <- latent_tbl %>%
  select(lineage_id, latent_score, confluence, cell_count) %>%
  pivot_longer(cols = c(confluence, cell_count), names_to = "covariate", values_to = "value")

ggplot(latent_cov_long, aes(x = value, y = latent_score, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 1.9) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_wrap(~ covariate, scales = "free_x") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Latent progression score versus confluence and cell count",
    x = NULL,
    y = "Latent progression score",
    color = "Lineage"
  )
```

Interpretation:

- a strong relationship here is not a problem by itself; it becomes a problem only if latent progression is nothing more than a density proxy
- if the same latent score occurs across a range of confluence or cell-count values, that supports a richer notion of morphological state

## Lineage Occupancy Along the Latent Trajectory

This view makes the sparse-versus-dense sampling problem explicit. It is especially useful for asking whether the control lineage contributes images to the middle of the shared trajectory rather than only the endpoints.

```{r latent-density-by-lineage}
latent_density_tbl <- latent_tbl %>%
  filter(is.finite(latent_score))

if (nrow(latent_density_tbl) >= 2 && dplyr::n_distinct(latent_density_tbl$latent_score) >= 2) {
  ggplot(latent_density_tbl, aes(x = latent_score, fill = lineage_id)) +
    geom_density(alpha = 0.4) +
    facet_wrap(~ lineage_id, ncol = 1) +
    scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)], guide = "none") +
    labs(
      title = "Lineage occupancy along the latent trajectory",
      subtitle = "A useful curved trajectory should make the intermediate region visible rather than collapsing it into endpoint bins",
      x = "Latent progression score",
      y = "Density"
    )
} else {
  cat("Latent-score density plot skipped because there are too few finite distinct values for a stable density estimate.\n")
}
```

## Does Temporal Structure Remain After Density Adjustment?

To check whether the latent score reflects more than trivial density effects, fit a simple additive model and inspect the residual temporal structure.

```{r residualized-latent}
density_adjust_fit <- mgcv::gam(
  latent_score ~ s(log_confluence, k = 6) + s(log_cell_count, k = 6) + lineage_id,
  data = latent_tbl,
  method = "REML"
)

latent_tbl <- latent_tbl %>%
  mutate(latent_resid_density = resid(density_adjust_fit, type = "response"))

ggplot(latent_tbl, aes(x = time_since_seeding_days, y = latent_resid_density, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 1.9) +
  geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6), linewidth = 0.9) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Latent progression residual after adjusting for confluence and cell count",
    subtitle = "Persistent temporal structure here argues against a purely density-driven embedding",
    x = "Time since seeding (days)",
    y = "Residual latent score",
    color = "Lineage"
  )
```

Interpretation:

- if temporal structure largely disappears here, the progression score is mostly a density axis
- if clear lineage-specific time structure remains, the embeddings are likely capturing morphology beyond simple occupancy proxies

```{r latent-partial-density}
partial_density_fit <- mgcv::gam(
  latent_score ~ s(time_since_seeding_days, k = 6) + s(log_confluence, k = 6) + s(log_cell_count, k = 6) + lineage_id,
  data = latent_tbl,
  method = "REML"
)

partial_terms <- predict(partial_density_fit, type = "terms")
confluence_term_idx <- grep("log_confluence", colnames(partial_terms))
cell_count_term_idx <- grep("log_cell_count", colnames(partial_terms))

confluence_term <- if (length(confluence_term_idx)) partial_terms[, confluence_term_idx[[1]]] else 0
cell_count_term <- if (length(cell_count_term_idx)) partial_terms[, cell_count_term_idx[[1]]] else 0

latent_tbl <- latent_tbl %>%
  mutate(
    latent_without_confluence = latent_score - confluence_term,
    latent_without_cell_count = latent_score - cell_count_term
  )

partial_long <- latent_tbl %>%
  select(lineage_id, confluence, cell_count, latent_without_confluence, latent_without_cell_count) %>%
  pivot_longer(cols = c(latent_without_confluence, latent_without_cell_count), names_to = "partial_type", values_to = "partial_score") %>%
  mutate(x_value = ifelse(partial_type == "latent_without_confluence", confluence, cell_count))

ggplot(partial_long, aes(x = x_value, y = partial_score, color = lineage_id)) +
  geom_point(alpha = 0.6, size = 1.7) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_wrap(~ partial_type, scales = "free_x") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Partial checks against simple density re-encoding",
    subtitle = "These plots help assess whether the curved latent score carries structure beyond confluence and cell count",
    x = NULL,
    y = "Partial latent score",
    color = "Lineage"
  )
```

Interpretation:

- if the latent score still varies in a structured way after subtracting a confluence or cell-count contribution, it is doing more than merely re-encoding density
- if these partial views are nearly flat, then the curved trajectory may still mostly represent density-linked growth state

## A Simple Alignment View Based on the Latent Score

Instead of forcing a formal time-warping model, use the latent score itself as the shared progression axis and estimate when each lineage reaches the same score levels.

```{r alignment-grid}
stage_score_grid <- seq(0.1, 0.9, by = 0.1)
alignment_tbl <- predict_time_to_stage(latent_tbl, lineage_levels, stage_score_grid)

ggplot(alignment_tbl, aes(x = stage_score, y = estimated_time_days, color = lineage_id)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Estimated time required to reach the same latent progression levels",
    subtitle = "Differences between curves suggest different traversal rates along a common axis",
    x = "Latent progression score",
    y = "Estimated time since seeding (days)",
    color = "Lineage"
  )
```

Interpretation:

- this is the simplest alignment diagnostic in the notebook
- if lineages have similar stage ordering but systematically different times-to-stage, that supports the idea of shared morphology traversed at different rates
- if the curves cross erratically or cannot be estimated for large parts of the score range, the notion of one shared progression is weaker

## Trajectory Zones for Interpretability

Instead of creating many narrow bins, the notebook now defines a small number of trajectory zones anchored on the shared curve:

- `Early / Seeded`
- `Intermediate`
- `Late / Near-Confluent`

The middle zone is intentionally broad enough to isolate biologically interesting intermediate states without collapsing them into endpoint structure.

```{r stage-bins}
early_cut <- quantile(latent_tbl$latent_score, probs = 0.20, na.rm = TRUE)
late_cut <- quantile(latent_tbl$latent_score, probs = 0.80, na.rm = TRUE)
core_lo <- quantile(latent_tbl$latent_score, probs = 0.35, na.rm = TRUE)
core_hi <- quantile(latent_tbl$latent_score, probs = 0.65, na.rm = TRUE)

latent_tbl <- latent_tbl %>%
  mutate(
    trajectory_zone = case_when(
      latent_score <= early_cut ~ "Early / Seeded",
      latent_score >= late_cut ~ "Late / Near-Confluent",
      TRUE ~ "Intermediate"
    ),
    intermediate_region = latent_score >= core_lo & latent_score <= core_hi
  )
```

```{r stage-in-pca}
stage_pca_plot <- ggplot(latent_tbl, aes(x = PC1, y = PC2, color = trajectory_zone)) +
  geom_point(alpha = 0.85, size = 2.1) +
  scale_color_manual(
    values = c(
      "Early / Seeded" = "#2C7BB6",
      "Intermediate" = "#D95F02",
      "Late / Near-Confluent" = "#1A9850"
    )
  ) +
  labs(
    title = "Trajectory zones mapped onto the shared curve in PC1-PC2 space",
    subtitle = "These zones are derived from arc length along the shared curved path, not from PC1 thresholds alone",
    x = "PC1",
    y = "PC2",
    color = "Trajectory zone"
  )

if (exists("curve_path_tbl") &&
    is.data.frame(curve_path_tbl) &&
    all(c("PC1", "PC2") %in% names(curve_path_tbl)) &&
    nrow(curve_path_tbl) >= 2) {
  stage_pca_plot <- stage_pca_plot +
    geom_path(
      data = curve_path_tbl,
      mapping = aes(x = PC1, y = PC2),
      inherit.aes = FALSE,
      linewidth = 1.2,
      color = "black"
    )
}

stage_pca_plot
```

```{r stage-composition}
stage_comp_tbl <- latent_tbl %>%
  dplyr::count(trajectory_zone, lineage_id, name = "n_images") %>%
  group_by(trajectory_zone) %>%
  mutate(stage_fraction = n_images / sum(n_images)) %>%
  ungroup()

ggplot(stage_comp_tbl, aes(x = trajectory_zone, y = stage_fraction, fill = lineage_id)) +
  geom_col(position = "stack", color = "white") +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Lineage composition within shared trajectory zones",
    x = "Trajectory zone",
    y = "Fraction of images in zone",
    fill = "Lineage"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r stage-time-boxplots}
ggplot(latent_tbl, aes(x = trajectory_zone, y = time_since_seeding_days, fill = lineage_id)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.35, position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Time since seeding within shared trajectory zones",
    subtitle = "Comparable intermediate stages should be populated at different times across lineages",
    x = "Trajectory zone",
    y = "Time since seeding (days)",
    fill = "Lineage"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r stage-density-boxplots}
stage_density_long <- latent_tbl %>%
  select(trajectory_zone, lineage_id, confluence, cell_count) %>%
  pivot_longer(cols = c(confluence, cell_count), names_to = "covariate", values_to = "value")

ggplot(stage_density_long, aes(x = trajectory_zone, y = value, fill = lineage_id)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.3, position = position_dodge(width = 0.8)) +
  facet_wrap(~ covariate, scales = "free_y") +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Confluence and cell count within shared trajectory zones",
    x = "Trajectory zone",
    y = NULL,
    fill = "Lineage"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

Interpretation:

- the `Intermediate` zone is the main target here: it should recover intermediate-growth images rather than splitting the data only into seeded and confluent endpoints
- if control images now appear in the intermediate bins, the curved latent score is doing something biologically useful that `PC1` was missing

## Candidate Control Images in Intermediate States

This section is the main deliverable of the notebook: it explicitly searches for control-lineage images that plausibly occupy intermediate shared states rather than early or late endpoints.

```{r candidate-control-images}
control_lineage_id <- infer_control_lineage(latent_tbl, params$control_lineage)
comparison_lineages <- setdiff(lineage_levels, control_lineage_id)

control_tbl <- latent_tbl %>% filter(as.character(lineage_id) == control_lineage_id)
comparison_tbl <- latent_tbl %>% filter(as.character(lineage_id) %in% comparison_lineages)

control_time_floor <- quantile(control_tbl$time_since_seeding_days, probs = 0.15, na.rm = TRUE)
control_confluence_ceiling <- quantile(control_tbl$confluence, probs = 0.90, na.rm = TRUE)
curve_distance_ceiling <- quantile(latent_tbl$distance_to_curve, probs = 0.75, na.rm = TRUE)

candidate_control_tbl <- control_tbl %>%
  mutate(
    centrality = abs(latent_score - 0.5),
    endpoint_distance = pmin(latent_score, 1 - latent_score)
  ) %>%
  filter(
    intermediate_region,
    is.finite(time_since_seeding_days),
    time_since_seeding_days > control_time_floor,
    is.finite(confluence),
    confluence < control_confluence_ceiling,
    is.finite(distance_to_curve),
    distance_to_curve <= curve_distance_ceiling
  )

comparison_intermediate_tbl <- comparison_tbl %>%
  filter(
    intermediate_region,
    is.finite(distance_to_curve),
    distance_to_curve <= curve_distance_ceiling
  )

nn_feature_cols <- c("latent_score", "PC1", "PC2", "PC3")
nn_feature_cols <- nn_feature_cols[nn_feature_cols %in% names(latent_tbl)]

latent_scaled_tbl <- latent_tbl
for (col in nn_feature_cols) {
  latent_scaled_tbl[[col]] <- as.numeric(scale(latent_scaled_tbl[[col]]))
}

scaled_query_tbl <- latent_scaled_tbl %>%
  filter(filename %in% candidate_control_tbl$filename) %>%
  mutate(.query_order = match(filename, candidate_control_tbl$filename)) %>%
  arrange(.query_order)
scaled_target_tbl <- latent_scaled_tbl %>%
  filter(filename %in% comparison_intermediate_tbl$filename) %>%
  mutate(.target_order = match(filename, comparison_intermediate_tbl$filename)) %>%
  arrange(.target_order)

nn_tbl <- nn_match_between_groups(scaled_query_tbl, scaled_target_tbl, nn_feature_cols)

if (!nrow(candidate_control_tbl)) {
  candidate_rank_tbl <- tibble(
    image_identifier = character(),
    lineage = character(),
    passage = integer(),
    time_since_seeding_days = double(),
    confluence = double(),
    cell_count = double(),
    latent_score = double(),
    distance_to_curve = double(),
    nearest_oxygen_starved_image = character(),
    nearest_oxygen_starved_lineage = character(),
    nearest_neighbor_distance = double(),
    nearest_neighbor_time_days = double(),
    nearest_neighbor_confluence = double(),
    nearest_neighbor_cell_count = double()
  )
} else {
  candidate_rank_tbl <- candidate_control_tbl %>%
    mutate(query_row = row_number()) %>%
    left_join(nn_tbl, by = "query_row") %>%
    mutate(
      matched_filename = dplyr::if_else(is.finite(nn_index), comparison_intermediate_tbl$filename[nn_index], NA_character_),
      matched_lineage = dplyr::if_else(is.finite(nn_index), as.character(comparison_intermediate_tbl$lineage_id[nn_index]), NA_character_),
      matched_time_since_seeding_days = dplyr::if_else(is.finite(nn_index), comparison_intermediate_tbl$time_since_seeding_days[nn_index], NA_real_),
      matched_confluence = dplyr::if_else(is.finite(nn_index), comparison_intermediate_tbl$confluence[nn_index], NA_real_),
      matched_cell_count = dplyr::if_else(is.finite(nn_index), comparison_intermediate_tbl$cell_count[nn_index], NA_real_),
      candidate_priority = rank(centrality, ties.method = "average") + rank(dplyr::coalesce(nn_distance, Inf), ties.method = "average") + rank(distance_to_curve, ties.method = "average")
    ) %>%
    arrange(candidate_priority, nn_distance, centrality) %>%
    transmute(
      image_identifier = filename,
      lineage = as.character(lineage_id),
      passage = passage_position,
      time_since_seeding_days,
      confluence,
      cell_count,
      latent_score,
      distance_to_curve,
      nearest_oxygen_starved_image = matched_filename,
      nearest_oxygen_starved_lineage = matched_lineage,
      nearest_neighbor_distance = nn_distance,
      nearest_neighbor_time_days = matched_time_since_seeding_days,
      nearest_neighbor_confluence = matched_confluence,
      nearest_neighbor_cell_count = matched_cell_count
    ) %>%
    slice_head(n = params$candidate_top_n)
}

knitr::kable(candidate_rank_tbl, digits = 3)
```

Interpretation:

- these are the control images most plausibly aligned with intermediate states from the oxygen-starved lineages
- the highest-priority candidates are central along the trajectory, reasonably close to the fitted curve, not immediate post-seeding, and close to intermediate non-control neighbors

## Alignment Views Focused on the Control Lineage

The next plots focus on whether the sparse control images align with intermediate states represented more densely in the oxygen-starved lineages.

```{r control-alignment-plots}
ggplot(latent_tbl, aes(x = latent_score, y = confluence, color = lineage_id)) +
  geom_point(alpha = 0.75, size = 2) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Confluence versus latent trajectory position",
    x = "Latent progression score",
    y = "Confluence",
    color = "Lineage"
  )

ggplot(latent_tbl, aes(x = latent_score, y = cell_count, color = lineage_id)) +
  geom_point(alpha = 0.75, size = 2) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Cell count versus latent trajectory position",
    x = "Latent progression score",
    y = "Cell count",
    color = "Lineage"
  )
```

```{r control-highlighted-in-pca}
highlight_tbl <- latent_tbl %>%
  mutate(
    highlight_group = case_when(
      filename %in% candidate_rank_tbl$image_identifier ~ "Top control candidates",
      as.character(lineage_id) == control_lineage_id ~ "Other control images",
      intermediate_region ~ "Intermediate non-control images",
      TRUE ~ "Other images"
    )
  )

ggplot(highlight_tbl, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = highlight_group), alpha = 0.75, size = 2.1) +
  scale_color_manual(
    values = c(
      "Top control candidates" = "#C0392B",
      "Other control images" = "#7F8C8D",
      "Intermediate non-control images" = "#117864",
      "Other images" = "#D5D8DC"
    )
  ) +
  labs(
    title = "Candidate control images highlighted in PCA space",
    subtitle = "This plot asks whether the top control candidates sit among intermediate images from the other lineages",
    x = "PC1",
    y = "PC2",
    color = NULL
  )
```

## Representative Images for Manual Review

If raw image paths are available, it is useful to review examples ordered by latent score. This chunk at minimum exports representative image identifiers for later inspection, with extra emphasis on the middle of the trajectory.

```{r representative-images}
representative_image_tbl <- latent_tbl %>%
  mutate(
    representative_group = case_when(
      latent_score <= quantile(latent_score, probs = 0.10, na.rm = TRUE) ~ "Early endpoint",
      latent_score >= quantile(latent_score, probs = 0.90, na.rm = TRUE) ~ "Late endpoint",
      intermediate_region ~ "Intermediate region",
      TRUE ~ "Transition"
    ),
    target_score = case_when(
      representative_group == "Early endpoint" ~ 0.08,
      representative_group == "Intermediate region" ~ 0.50,
      representative_group == "Late endpoint" ~ 0.92,
      TRUE ~ 0.30
    ),
    score_distance = abs(latent_score - target_score)
  ) %>%
  group_by(representative_group, lineage_id) %>%
  slice_min(order_by = score_distance, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  select(
    representative_group,
    lineage_id,
    filename,
    passage_position,
    time_since_seeding_days,
    confluence,
    cell_count,
    latent_score,
    distance_to_curve,
    any_of(c("filepath", "source_filepath"))
  ) %>%
  arrange(representative_group, lineage_id, latent_score)

knitr::kable(representative_image_tbl, digits = 3)
```

Interpretation:

- this table is meant for direct manual image review around the endpoints and especially around the intermediate region
- if the candidate control images also look morphologically intermediate by eye, that is strong practical support for the refined trajectory

## Provisional Conclusions

This notebook intentionally stops short of a production trajectory model. At this stage, the main outputs to trust are:

1. the curvature diagnostics in `PC1-PC2` and `PC1-PC3`
2. the principal-curve arc-length score and distance-to-curve diagnostic
3. the latent-score versus time and density checks
4. the refined middle-resolving stage bins
5. the ranked control-image candidate table and highlighted PCA views

Reasonable next steps depend on what these diagnostics show:

- If the curved trajectory resolves intermediate control images that align closely with intermediate oxygen-starved images, then the shared-trajectory interpretation is biologically useful.
- If control images still collapse almost entirely into endpoint regions even after curvature is modeled, then the issue is likely sparse sampling rather than trajectory parameterization alone.
- If most of the refined trajectory disappears after accounting for `confluence` and `cell_count`, then the embedding is still behaving mainly as a density summary.

The most important practical output is `candidate_rank_tbl`: it is the notebook's current best answer to which control-lineage images most plausibly correspond to intermediate morphological states along the shared trajectory.
