---
title: "Exemplar Lineage Trajectory Alignment"
output:
  html_document:
    toc: true
    toc_float: true
params:
  lineage_rds: "core_data/lineages.Rds"
  embedding_csv: "data/all_images/manifests/raw_image_embeddings.csv"
  exemplar_lineages: null
  n_top_pcs: 8
  n_stage_bins: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5
)
```

## Overview

This notebook is a first-pass exploratory workflow for asking whether three exemplar lineages appear to move through a shared morphological progression in image-embedding space.

The emphasis is on interpretable diagnostics rather than a black-box trajectory method. The workflow starts with direct checks of coverage and covariates, uses PCA as the main low-dimensional representation, and only then adds a simple latent progression score for provisional cross-lineage alignment.

Evidence for a shared trajectory would look like this:

- the lineages overlap in the same low-dimensional regions rather than occupying disjoint clouds
- the major embedding axes vary smoothly with `time_since_seeding`
- images with similar latent progression scores can come from different lineages at different absolute times
- the latent score still carries temporal structure after checking `confluence` and `cell_count`

Evidence against a shared trajectory would look like this:

- lineage-specific clusters with little overlap in PCA space
- different lineages following visibly different directions in low-dimensional space
- latent progression being almost entirely reducible to trivial density proxies such as `cell_count` or `confluence`
- stage bins that separate lineages rather than mixing them

## Packages

The core workflow uses standard packages. `uwot` and `princurve` are optional extras for a secondary UMAP view and an optional principal-curve score.

```{r packages}
required_packages <- c(
  "data.table",
  "dplyr",
  "ggplot2",
  "mgcv",
  "purrr",
  "readr",
  "scales",
  "stringr",
  "tibble",
  "tidyr"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    sprintf("Install required packages before rendering: %s", paste(missing_packages, collapse = ", ")),
    call. = FALSE
  )
}

library(data.table)
library(dplyr)
library(ggplot2)
library(mgcv)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

has_uwot <- requireNamespace("uwot", quietly = TRUE)
has_princurve <- requireNamespace("princurve", quietly = TRUE)
```

## Plot Style

The notebook uses a single theme and a stable lineage palette so plots are easy to compare.

```{r plot-style}
theme_set(
  theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "right"
    )
)

lineage_palette <- c(
  "#1B4F72",
  "#B03A2E",
  "#117864",
  "#7D6608",
  "#6C3483"
)
```

Short plot-reading convention used throughout:

- points are individual images
- smooth curves summarize lineage-level trends without assuming equal sampling
- `time_since_seeding` is treated as a within-passage clock, not as a synonym for morphological state

## Helpers

The notebook reconstructs image metadata from `lineages.Rds`, then joins it to the one-row-per-image embedding table.

```{r helpers}
`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

resolve_project_root <- function() {
  candidates <- c(
    normalizePath(".", winslash = "/", mustWork = FALSE),
    normalizePath("..", winslash = "/", mustWork = FALSE),
    normalizePath("../..", winslash = "/", mustWork = FALSE)
  )

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "README.md")) &&
        dir.exists(file.path(candidate, "workflows"))) {
      return(candidate)
    }
  }

  stop("Could not locate project root from current working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

project_path <- function(path) {
  is_absolute <- grepl("^[A-Za-z]:[/\\\\]|^/", path)
  out <- path
  out[!is_absolute] <- file.path(project_root, path[!is_absolute])
  out
}

normalize_relpath <- function(path) {
  gsub("\\\\", "/", path)
}

parse_mixed_datetime <- function(x) {
  x <- as.character(x)
  out <- as.POSIXct(rep(NA_character_, length(x)), tz = "UTC")
  formats <- c(
    "%Y-%m-%d %H:%M:%S",
    "%Y-%m-%dT%H:%M:%SZ",
    "%m/%d/%Y %H:%M",
    "%m/%d/%Y %H:%M:%S"
  )

  for (fmt in formats) {
    missing <- is.na(out) & !is.na(x) & nzchar(x)
    if (!any(missing)) next
    out[missing] <- as.POSIXct(x[missing], format = fmt, tz = "UTC")
  }

  out
}

embedding_column_names <- function(x) {
  grep("^embedding_[0-9]{3}$", names(x), value = TRUE)
}

extract_lineage <- function(lineages, lineage_id) {
  if (lineage_id %in% names(lineages)) {
    return(lineages[[lineage_id]])
  }

  hit <- vapply(lineages, function(x) identical(as.character(x$lineage_id), as.character(lineage_id)), logical(1))
  if (!any(hit)) {
    stop(sprintf("Lineage id not found: %s", lineage_id), call. = FALSE)
  }

  lineages[[which(hit)[[1]]]]
}

build_lineage_image_meta <- function(lineage_obj) {
  passage_positions <- stats::setNames(seq_along(lineage_obj$passage_ids), lineage_obj$passage_ids)

  image_meta_list <- lapply(names(lineage_obj$passages), function(passage_id) {
    passage_dt <- data.table::as.data.table(lineage_obj$passages[[passage_id]])
    if (!nrow(passage_dt)) return(NULL)

    keep_cols <- intersect(
      c("id", "filename", "field", "date", "areaOccupied_um2", "cellCount", "condition_label", "event", "passage"),
      names(passage_dt)
    )
    image_meta <- unique(passage_dt[, ..keep_cols], by = "filename")
    image_meta[, passage_id := as.character(passage_id)]
    image_meta[, lineage_id := as.character(lineage_obj$lineage_id)]
    image_meta[, passage_position := unname(passage_positions[passage_id])]
    image_meta[, image_key := sub("\\.[^.]+$", "", basename(filename))]
    image_meta
  })

  image_meta <- data.table::rbindlist(image_meta_list, fill = TRUE)
  if (!nrow(image_meta)) {
    stop(sprintf("No manifest rows found in lineage %s.", lineage_obj$lineage_id), call. = FALSE)
  }

  image_meta[, image_time := parse_mixed_datetime(date)]
  image_meta[, passage_start_time := suppressWarnings(min(image_time, na.rm = TRUE)), by = passage_id]
  image_meta[!is.finite(as.numeric(passage_start_time)), passage_start_time := as.POSIXct(NA, tz = "UTC")]

  # Treat the first observed image within a passage as the local seeding-time anchor.
  image_meta[, time_since_seeding_hours := as.numeric(difftime(image_time, passage_start_time, units = "hours"))]
  image_meta[, time_since_seeding_days := time_since_seeding_hours / 24]

  image_meta[, confluence := suppressWarnings(as.numeric(areaOccupied_um2))]
  image_meta[, cell_count := suppressWarnings(as.numeric(cellCount))]
  image_meta[, log_confluence := log1p(pmax(confluence, 0))]
  image_meta[, log_cell_count := log1p(pmax(cell_count, 0))]

  image_meta
}

choose_exemplar_lineages <- function(lineages, embedding_tbl, n = 3L) {
  coverage_tbl <- purrr::map_dfr(names(lineages), function(lineage_name) {
    lineage_obj <- lineages[[lineage_name]]
    meta <- build_lineage_image_meta(lineage_obj)
    tibble(
      lineage_id = as.character(lineage_obj$lineage_id),
      n_manifest_images = nrow(meta),
      n_matched_images = sum(meta$filename %in% embedding_tbl$filename)
    )
  })

  coverage_tbl %>%
    arrange(desc(n_matched_images), desc(n_manifest_images), lineage_id) %>%
    slice_head(n = n) %>%
    pull(lineage_id)
}

load_exemplar_panel <- function(lineage_rds, embedding_csv, exemplar_lineages = NULL) {
  lineages <- readRDS(project_path(lineage_rds))
  embedding_tbl <- data.table::fread(project_path(embedding_csv)) %>%
    as_tibble()

  embedding_cols <- embedding_column_names(embedding_tbl)
  if (!length(embedding_cols)) {
    stop("No embedding columns were found in the image-level embedding CSV.", call. = FALSE)
  }

  if (is.null(exemplar_lineages) || !length(exemplar_lineages)) {
    exemplar_lineages <- choose_exemplar_lineages(lineages, embedding_tbl, n = 3L)
  }

  lineage_meta <- purrr::map_dfr(exemplar_lineages, function(lineage_id) {
    build_lineage_image_meta(extract_lineage(lineages, lineage_id)) %>% as_tibble()
  })

  if (!nrow(lineage_meta)) {
    stop("No lineage metadata rows were recovered for the exemplar lineages.", call. = FALSE)
  }

  # Join on filename because the image-level embedding table is one row per image file.
  joined <- lineage_meta %>%
    inner_join(embedding_tbl, by = "filename")

  if (!nrow(joined)) {
    stop("No exemplar-lineage images matched the embedding CSV.", call. = FALSE)
  }

  joined %>%
    mutate(
      lineage_id = factor(lineage_id, levels = exemplar_lineages),
      passage_position = as.integer(passage_position)
    ) %>%
    distinct(lineage_id, passage_id, filename, .keep_all = TRUE)
}

fit_pc_association <- function(score, predictor, predictor_type = c("numeric", "factor")) {
  predictor_type <- match.arg(predictor_type)

  dat <- tibble(score = score, predictor = predictor) %>%
    filter(is.finite(score), !is.na(predictor))

  if (nrow(dat) < 8) {
    return(tibble(metric = NA_real_, n = nrow(dat)))
  }

  if (predictor_type == "numeric") {
    unique_n <- dplyr::n_distinct(dat$predictor)
    if (unique_n < 4) {
      fit <- lm(score ~ predictor, data = dat)
      out <- summary(fit)$adj.r.squared
    } else {
      k_basis <- max(4, min(8, unique_n - 1))
      fit <- mgcv::gam(score ~ s(predictor, k = k_basis), data = dat, method = "REML")
      out <- summary(fit)$r.sq
    }
  } else {
    if (dplyr::n_distinct(dat$predictor) < 2) {
      out <- NA_real_
    } else {
      fit <- lm(score ~ predictor, data = dat)
      out <- summary(fit)$adj.r.squared
    }
  }

  tibble(metric = out, n = nrow(dat))
}

weighted_pc_association <- function(score_tbl, score_cols, var_explained, predictor, predictor_type) {
  stats_tbl <- purrr::map_dfr(seq_along(score_cols), function(i) {
    fit_pc_association(score_tbl[[score_cols[[i]]]], predictor = predictor, predictor_type = predictor_type) %>%
      mutate(
        pc = score_cols[[i]],
        variance_weight = var_explained[[i]]
      )
  })

  tibble(
    weighted_r2 = weighted.mean(stats_tbl$metric, w = stats_tbl$variance_weight, na.rm = TRUE),
    mean_pc_r2 = mean(stats_tbl$metric, na.rm = TRUE),
    n_pcs = length(score_cols)
  )
}

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || diff(rng) <= 0) {
    return(rep(NA_real_, length(x)))
  }
  (x - rng[1]) / diff(rng)
}

nearest_other_lineage_distance <- function(mat, lineage) {
  out <- rep(NA_real_, nrow(mat))

  for (i in seq_len(nrow(mat))) {
    keep <- lineage != lineage[[i]]
    if (!any(keep)) next
    d <- sqrt(rowSums((t(t(mat[keep, , drop = FALSE]) - mat[i, ]))^2))
    out[[i]] <- min(d, na.rm = TRUE)
  }

  out
}

predict_time_to_stage <- function(df, lineage_levels, stage_grid) {
  purrr::map_dfr(lineage_levels, function(lineage_id) {
    dat <- df %>%
      filter(lineage_id == .env$lineage_id) %>%
      filter(is.finite(time_since_seeding_days), is.finite(latent_score))

    if (nrow(dat) < 8 || dplyr::n_distinct(dat$time_since_seeding_days) < 4) {
      return(tibble())
    }

    fit <- mgcv::gam(
      latent_score ~ s(time_since_seeding_days, k = min(8, dplyr::n_distinct(dat$time_since_seeding_days) - 1)),
      data = dat,
      method = "REML"
    )

    time_grid <- seq(
      min(dat$time_since_seeding_days, na.rm = TRUE),
      max(dat$time_since_seeding_days, na.rm = TRUE),
      length.out = 200
    )

    pred_tbl <- tibble(
      time_since_seeding_days = time_grid,
      latent_score = as.numeric(predict(fit, newdata = tibble(time_since_seeding_days = time_grid)))
    )

    purrr::map_dfr(stage_grid, function(target_score) {
      hit <- pred_tbl %>%
        mutate(distance = abs(latent_score - target_score)) %>%
        slice_min(order_by = distance, n = 1, with_ties = FALSE)

      tibble(
        lineage_id = lineage_id,
        stage_score = target_score,
        estimated_time_days = hit$time_since_seeding_days
      )
    })
  })
}
```

## Load the Exemplar Panel

Start by loading the image-level embedding table and reconstructing metadata for three exemplar lineages from `lineages.Rds`.

If `params$exemplar_lineages` is left empty, the notebook will pick the three lineages with the most matched embedded images. That is only a fallback for convenience; for serious review, it is better to set the intended lineages explicitly.

```{r load-data}
panel_tbl <- load_exemplar_panel(
  lineage_rds = params$lineage_rds,
  embedding_csv = params$embedding_csv,
  exemplar_lineages = params$exemplar_lineages
)

embedding_cols <- embedding_column_names(panel_tbl)

analysis_tbl <- panel_tbl %>%
  mutate(
    lineage_id = factor(lineage_id, levels = unique(lineage_id)),
    passage_label = paste0("P", passage_position)
  )

lineage_levels <- levels(analysis_tbl$lineage_id)

analysis_overview <- tibble(
  n_images = nrow(analysis_tbl),
  n_lineages = dplyr::n_distinct(analysis_tbl$lineage_id),
  n_passages = dplyr::n_distinct(interaction(analysis_tbl$lineage_id, analysis_tbl$passage_id, drop = TRUE)),
  n_embedding_dims = length(embedding_cols),
  exemplar_lineages = paste(lineage_levels, collapse = ", ")
)

knitr::kable(analysis_overview)
```

Interpretation:

- this table is the quick sanity check that the notebook is operating on image-level rows
- if the selected lineages are not the intended exemplars, set `params$exemplar_lineages` explicitly before interpreting any downstream plot

## Basic QC and Coverage

Before doing any dimensionality reduction, check missingness, numeric ranges, and whether coverage is balanced enough to support a trajectory comparison.

```{r qc-summary-table}
qc_tbl <- analysis_tbl %>%
  summarise(
    across(
      c(time_since_seeding_days, confluence, cell_count, all_of(embedding_cols)),
      ~ mean(is.na(.x))
    )
  ) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_fraction") %>%
  arrange(desc(missing_fraction))

knitr::kable(head(qc_tbl, 12), digits = 3)
```

The next table summarizes each lineage separately.

```{r lineage-qc}
lineage_qc_tbl <- analysis_tbl %>%
  group_by(lineage_id) %>%
  summarise(
    n_images = n(),
    n_passages = n_distinct(passage_id),
    time_min_days = min(time_since_seeding_days, na.rm = TRUE),
    time_max_days = max(time_since_seeding_days, na.rm = TRUE),
    median_confluence = median(confluence, na.rm = TRUE),
    median_cell_count = median(cell_count, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(lineage_qc_tbl, digits = 2)
```

Interpretation:

- large differences in `n_images` or `time_max_days` are acceptable, but they matter when comparing smooths
- if one lineage is concentrated in only a narrow time window, that lineage can still inform overlap in state space but not broad alignment

Coverage plots are the first place to see whether irregular sampling or passage imbalance could create spurious apparent trajectories.

```{r coverage-plots}
coverage_tbl <- analysis_tbl %>%
  count(lineage_id, passage_position, time_since_seeding_days, name = "n_images")

ggplot(coverage_tbl, aes(x = time_since_seeding_days, y = factor(passage_position), size = n_images, color = lineage_id)) +
  geom_point(alpha = 0.8) +
  scale_size_continuous(range = c(2.5, 7)) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Sampling coverage by lineage, passage, and time since seeding",
    x = "Time since seeding (days)",
    y = "Passage position",
    color = "Lineage",
    size = "Images"
  )
```

```{r time-density}
ggplot(analysis_tbl, aes(x = time_since_seeding_days, fill = lineage_id)) +
  geom_histogram(binwidth = 0.5, position = "identity", alpha = 0.45, color = "white") +
  facet_wrap(~ lineage_id, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)], guide = "none") +
  labs(
    title = "Time-since-seeding coverage within each lineage",
    x = "Time since seeding (days)",
    y = "Image count"
  )
```

Interpretation:

- these plots should be read before trusting any smooth trend
- if one lineage has strong gaps in time or is represented mainly by one passage, apparent rate differences may just reflect missing intermediate states

## Covariate Checks

The embeddings may capture density-related biology, but they may also be dominated by simple occupancy effects. These direct covariate plots are therefore both biological summaries and sanity checks.

```{r covariate-pairs}
covariate_long <- analysis_tbl %>%
  select(lineage_id, passage_position, time_since_seeding_days, confluence, cell_count) %>%
  pivot_longer(
    cols = c(confluence, cell_count),
    names_to = "metric",
    values_to = "value"
  )

ggplot(covariate_long, aes(x = time_since_seeding_days, y = value, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Confluence and cell count versus time since seeding",
    x = "Time since seeding (days)",
    y = NULL,
    color = "Lineage"
  )
```

Interpretation:

- if `confluence` and `cell_count` are nearly deterministic functions of time in every lineage, then a strong temporal signal in the embeddings could still be largely a density signal
- if lineages differ in density at the same time, that creates a useful stress test for whether embeddings are capturing something beyond raw time

## Preprocess the Embeddings

The embedding dimensions are centered and scaled before PCA. This prevents large-variance coordinates from dominating the first PCs for purely numeric reasons.

```{r pca}
complete_tbl <- analysis_tbl %>%
  filter(
    if_all(all_of(embedding_cols), is.finite),
    is.finite(time_since_seeding_days)
  )

embedding_mat <- as.matrix(complete_tbl[, embedding_cols])
storage.mode(embedding_mat) <- "double"

# Scale each embedding dimension so PCA reflects structure rather than raw scale.
pca_fit <- prcomp(embedding_mat, center = TRUE, scale. = TRUE)

n_top_pcs <- min(params$n_top_pcs, ncol(pca_fit$x))
score_cols <- paste0("PC", seq_len(n_top_pcs))
score_tbl <- bind_cols(
  complete_tbl %>% select(-all_of(embedding_cols)),
  as_tibble(pca_fit$x[, seq_len(n_top_pcs), drop = FALSE], .name_repair = "minimal")
)
names(score_tbl)[(ncol(score_tbl) - n_top_pcs + 1):ncol(score_tbl)] <- score_cols

var_explained <- (pca_fit$sdev^2) / sum(pca_fit$sdev^2)
var_tbl <- tibble(
  pc = paste0("PC", seq_along(var_explained)),
  variance_explained = var_explained,
  cumulative_variance = cumsum(var_explained)
)
```

The variance-explained profile tells us how concentrated or diffuse image-state variation is.

```{r variance-explained}
ggplot(var_tbl %>% slice_head(n = 15), aes(x = reorder(pc, variance_explained), y = variance_explained)) +
  geom_col(fill = "#2C7FB8") +
  geom_text(aes(label = percent(variance_explained, accuracy = 0.1)), hjust = -0.05, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Variance explained by the leading principal components",
    x = NULL,
    y = "Fraction of total embedding variance"
  )
```

Interpretation:

- if PC1 is clearly dominant and its biology is sensible, it becomes a strong candidate latent progression axis
- if variance is spread across many PCs, shared progression may still exist, but it is less likely to be captured by a single simple coordinate

Loading summaries are kept compact here: the goal is to identify whether a few embedding dimensions dominate each PC.

```{r loading-summaries}
loading_tbl <- as_tibble(pca_fit$rotation[, seq_len(min(4, ncol(pca_fit$rotation))), drop = FALSE], rownames = "embedding_dim") %>%
  pivot_longer(-embedding_dim, names_to = "pc", values_to = "loading") %>%
  group_by(pc) %>%
  slice_max(order_by = abs(loading), n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(pc, desc(abs(loading)))

knitr::kable(loading_tbl, digits = 3)
```

Interpretation:

- these are not biologically interpretable features by themselves, but they help detect whether a PC is broad-based or dominated by a small corner of the embedding vector

## PCA Views of Image State

PCA is the main basis of inference in this notebook. The first view asks whether the three lineages occupy overlapping or distinct regions of low-dimensional embedding space.

```{r pca-by-lineage}
ggplot(score_tbl, aes(x = PC1, y = PC2, color = lineage_id)) +
  geom_point(alpha = 0.8, size = 2.2) +
  stat_ellipse(linewidth = 0.7, alpha = 0.4) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "PCA state space colored by lineage",
    x = "PC1",
    y = "PC2",
    color = "Lineage"
  )
```

Interpretation:

- strong overlap is the first qualitative sign that a shared trajectory may exist
- cleanly separated lineage clouds would instead suggest lineage-specific morphology or strong uncorrected batch structure

The next views ask whether the same PCA space orders images by time and by density-related covariates.

```{r pca-color-views}
plot_by_numeric <- function(var_name, title_text) {
  ggplot(score_tbl, aes(x = PC1, y = PC2, color = .data[[var_name]])) +
    geom_point(size = 2.1, alpha = 0.85) +
    scale_color_viridis_c(option = "C", na.value = "grey80") +
    labs(
      title = title_text,
      x = "PC1",
      y = "PC2",
      color = var_name
    )
}

plot_by_numeric("time_since_seeding_days", "PCA state space colored by time since seeding")
plot_by_numeric("confluence", "PCA state space colored by confluence")
plot_by_numeric("cell_count", "PCA state space colored by cell count")
```

Interpretation:

- a smooth color gradient along the same geometric axis in all three plots suggests a shared dominant progression
- if the time gradient and density gradients point in different directions, that is useful evidence that morphological state is not reducible to a single occupancy proxy

## Quantify Association Between Embeddings and Key Covariates

To summarize which variables explain the major axes of embedding variation, the notebook computes variance-weighted per-PC association scores across the leading PCs.

This is intentionally simple: it is meant as an interpretable first-pass summary, not a formal variance-partitioning model.

```{r pc-association-summary}
association_tbl <- bind_rows(
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$time_since_seeding_days, "numeric") %>%
    mutate(variable = "time_since_seeding"),
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$confluence, "numeric") %>%
    mutate(variable = "confluence"),
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$cell_count, "numeric") %>%
    mutate(variable = "cell_count"),
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$lineage_id, "factor") %>%
    mutate(variable = "lineage"),
  weighted_pc_association(score_tbl, score_cols, var_explained[seq_along(score_cols)], score_tbl$passage_position, "numeric") %>%
    mutate(variable = "passage_position")
) %>%
  select(variable, weighted_r2, mean_pc_r2, n_pcs)

knitr::kable(association_tbl, digits = 3)
```

```{r pc-association-plot}
ggplot(association_tbl, aes(x = reorder(variable, weighted_r2), y = weighted_r2)) +
  geom_col(fill = "#21618C") +
  geom_text(aes(label = round(weighted_r2, 2)), hjust = -0.05, size = 3.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Variance-weighted association of leading PCs with key variables",
    x = NULL,
    y = "Weighted R-squared across leading PCs"
  )
```

Interpretation:

- if `time_since_seeding` is strongest but `confluence` and `cell_count` are also strong, the embedding is likely tracking both temporal progression and density-related biology
- if `lineage` dominates, shared alignment becomes less plausible
- if `passage_position` remains strong after time, that suggests slow lineage history effects beyond within-passage seeding time

## PC Scores Versus Time and Density Covariates

The next plots examine the leading PCs directly. They help decide whether a single PC is interpretable enough to serve as a provisional progression axis.

```{r pc-trends}
pc_long <- score_tbl %>%
  select(lineage_id, passage_position, time_since_seeding_days, confluence, cell_count, all_of(score_cols[seq_len(min(4, length(score_cols)))])) %>%
  pivot_longer(cols = starts_with("PC"), names_to = "pc", values_to = "score")

ggplot(pc_long, aes(x = time_since_seeding_days, y = score, color = lineage_id)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_wrap(~ pc, scales = "free_y", ncol = 2) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Leading PC scores versus time since seeding",
    x = "Time since seeding (days)",
    y = "PC score",
    color = "Lineage"
  )
```

```{r pc-vs-covariates}
pc_cov_long <- score_tbl %>%
  select(lineage_id, PC1, PC2, confluence, cell_count) %>%
  pivot_longer(cols = c(PC1, PC2), names_to = "pc", values_to = "score") %>%
  pivot_longer(cols = c(confluence, cell_count), names_to = "covariate", values_to = "value")

ggplot(pc_cov_long, aes(x = value, y = score, color = lineage_id)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_grid(pc ~ covariate, scales = "free_x") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Leading PCs versus confluence and cell count",
    x = NULL,
    y = "PC score",
    color = "Lineage"
  )
```

Interpretation:

- a usable simple progression axis should change smoothly with time in more than one lineage
- if PC1 mostly reproduces confluence or cell count and little else, it is still informative, but it should be interpreted as a density-linked state axis rather than a generic morphology axis

## Secondary Nonlinear View

UMAP is included only as a secondary visualization. It can help reveal local neighborhoods, but it is not used as the main basis of inference.

```{r umap, eval=has_uwot}
set.seed(1)

umap_fit <- uwot::umap(
  embedding_mat,
  n_neighbors = min(20, nrow(embedding_mat) - 1),
  min_dist = 0.25,
  metric = "euclidean",
  scale = FALSE,
  verbose = FALSE
)

umap_tbl <- score_tbl %>%
  mutate(
    UMAP1 = umap_fit[, 1],
    UMAP2 = umap_fit[, 2]
  )

ggplot(umap_tbl, aes(x = UMAP1, y = UMAP2, color = lineage_id)) +
  geom_point(alpha = 0.8, size = 2.1) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Secondary UMAP view of image state",
    subtitle = "Use as a neighborhood view, not as the primary trajectory basis",
    x = "UMAP1",
    y = "UMAP2",
    color = "Lineage"
  )
```

```{r umap-note, eval=!has_uwot, results='asis'}
cat("`uwot` is not installed, so the secondary UMAP view is skipped.\n")
```

Interpretation:

- if UMAP preserves the same broad lineage overlap already seen in PCA, that increases confidence that the overlap is not an artifact of one linear projection
- if UMAP produces lineage islands that are absent in PCA, inspect carefully rather than over-interpreting either view

## Do the Lineages Occupy Similar Regions of State Space?

A shared trajectory requires at least partial overlap in low-dimensional state space. As a simple summary, the next plot and table look at cross-lineage proximity in the first three PCs.

```{r overlap-diagnostics}
pc_overlap_cols <- c("PC1", "PC2", score_cols[min(3, length(score_cols))])
pc_overlap_cols <- unique(pc_overlap_cols)
pc_overlap_mat <- as.matrix(score_tbl[, pc_overlap_cols])
score_tbl$nearest_other_lineage_dist <- nearest_other_lineage_distance(pc_overlap_mat, score_tbl$lineage_id)

ggplot(score_tbl, aes(x = lineage_id, y = nearest_other_lineage_dist, fill = lineage_id)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.35) +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)], guide = "none") +
  labs(
    title = "Distance to the nearest image from a different lineage",
    subtitle = "Smaller values imply stronger cross-lineage overlap in low-dimensional state space",
    x = "Lineage",
    y = "Nearest other-lineage distance in PC space"
  )
```

```{r overlap-summary}
overlap_summary_tbl <- score_tbl %>%
  group_by(lineage_id) %>%
  summarise(
    median_nearest_other_lineage_dist = median(nearest_other_lineage_dist, na.rm = TRUE),
    q75_nearest_other_lineage_dist = quantile(nearest_other_lineage_dist, probs = 0.75, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(overlap_summary_tbl, digits = 3)
```

Interpretation:

- low cross-lineage nearest-neighbor distances support the idea that different lineages revisit similar image states
- very large distances would argue for lineage-specific trajectories even if each lineage changes smoothly on its own

## Smooth Trajectories in Low-Dimensional Space

Now model each lineage as a smooth path through the first two PCs as a function of `time_since_seeding`.

This is still descriptive. The purpose is to see whether the lineages move through similar regions but at different rates, not to force them onto one common curve.

```{r smooth-trajectories}
trajectory_grid <- purrr::map_dfr(lineage_levels, function(lineage_id) {
  dat <- score_tbl %>%
    filter(lineage_id == .env$lineage_id) %>%
    filter(is.finite(time_since_seeding_days))

  if (nrow(dat) < 8 || dplyr::n_distinct(dat$time_since_seeding_days) < 4) {
    return(tibble())
  }

  time_grid <- seq(
    min(dat$time_since_seeding_days, na.rm = TRUE),
    max(dat$time_since_seeding_days, na.rm = TRUE),
    length.out = 100
  )

  fit_pc1 <- mgcv::gam(
    PC1 ~ s(time_since_seeding_days, k = min(8, dplyr::n_distinct(dat$time_since_seeding_days) - 1)),
    data = dat,
    method = "REML"
  )
  fit_pc2 <- mgcv::gam(
    PC2 ~ s(time_since_seeding_days, k = min(8, dplyr::n_distinct(dat$time_since_seeding_days) - 1)),
    data = dat,
    method = "REML"
  )

  tibble(
    lineage_id = lineage_id,
    time_since_seeding_days = time_grid,
    PC1 = as.numeric(predict(fit_pc1, newdata = tibble(time_since_seeding_days = time_grid))),
    PC2 = as.numeric(predict(fit_pc2, newdata = tibble(time_since_seeding_days = time_grid)))
  )
})

ggplot(score_tbl, aes(x = PC1, y = PC2, color = lineage_id)) +
  geom_point(alpha = 0.35, size = 1.8) +
  geom_path(data = trajectory_grid, linewidth = 1.2, lineend = "round") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Lineage-specific smooth trajectories in PCA space",
    subtitle = "Curves are smooths of PC1 and PC2 versus time since seeding",
    x = "PC1",
    y = "PC2",
    color = "Lineage"
  )
```

Interpretation:

- a shared trajectory is most plausible when these smooths trace broadly similar routes through the same region
- if the curves point in different directions or occupy different lobes, that is evidence for genuinely different lineage paths rather than rate differences along a common path

## Derive a Provisional Latent Progression Score

The simplest useful latent score is often just PC1. A principal curve is only worth using if it adds meaningful curvature beyond what PC1 already captures.

The notebook therefore tries both when possible and chooses the simpler score if the principal-curve ordering is nearly the same as PC1.

```{r latent-score}
pc1_oriented <- score_tbl$PC1

# Orient the simple PC1 score so larger values correspond to later average times.
if (stats::cor(pc1_oriented, score_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman") < 0) {
  pc1_oriented <- -pc1_oriented
}

latent_tbl <- score_tbl %>%
  mutate(
    latent_score_pc1 = rescale01(pc1_oriented),
    latent_score = latent_score_pc1,
    latent_score_method = "PC1"
  )

if (has_princurve && length(score_cols) >= 3) {
  curve_input <- as.matrix(score_tbl[, c("PC1", "PC2", "PC3")])
  principal_curve_fit <- princurve::principal_curve(curve_input)

  lambda <- principal_curve_fit$lambda
  if (stats::cor(lambda, score_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman") < 0) {
    lambda <- -lambda
  }

  lambda01 <- rescale01(lambda)
  pc1_vs_curve <- stats::cor(latent_tbl$latent_score_pc1, lambda01, use = "pairwise.complete.obs", method = "spearman")

  if (is.finite(pc1_vs_curve) && abs(pc1_vs_curve) < 0.95) {
    latent_tbl <- latent_tbl %>%
      mutate(
        latent_score_curve = lambda01,
        latent_score = lambda01,
        latent_score_method = "principal_curve"
      )
  } else {
    latent_tbl <- latent_tbl %>%
      mutate(latent_score_curve = lambda01)
  }
} else {
  latent_tbl <- latent_tbl %>%
    mutate(latent_score_curve = NA_real_)
}

latent_method_tbl <- tibble(
  chosen_method = unique(latent_tbl$latent_score_method),
  spearman_pc1_vs_time = cor(latent_tbl$latent_score_pc1, latent_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman"),
  spearman_latent_vs_time = cor(latent_tbl$latent_score, latent_tbl$time_since_seeding_days, use = "pairwise.complete.obs", method = "spearman")
)

knitr::kable(latent_method_tbl, digits = 3)
```

Interpretation:

- if the chosen method is `PC1`, that means the notebook judged the simple linear axis to be adequate
- if the chosen method is `principal_curve`, interpret that as mild evidence that the common progression bends in low-dimensional space

## Latent Progression Score Versus Time Since Seeding

This is the key first-pass alignment plot: it asks whether the same latent state can appear at different absolute times in different lineages.

```{r latent-vs-time}
ggplot(latent_tbl, aes(x = time_since_seeding_days, y = latent_score, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6), linewidth = 0.9) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Latent progression score versus time since seeding",
    subtitle = paste("Chosen score:", unique(latent_tbl$latent_score_method)),
    x = "Time since seeding (days)",
    y = "Latent progression score",
    color = "Lineage"
  )
```

Interpretation:

- parallel but shifted curves suggest similar morphological states reached at different times
- sharply different curve shapes suggest lineage-specific progressions rather than simple rate differences

## Latent Progression Score Versus Confluence and Cell Count

These plots show how much the inferred progression score tracks density-related covariates.

```{r latent-vs-density}
latent_cov_long <- latent_tbl %>%
  select(lineage_id, latent_score, confluence, cell_count) %>%
  pivot_longer(cols = c(confluence, cell_count), names_to = "covariate", values_to = "value")

ggplot(latent_cov_long, aes(x = value, y = latent_score, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 1.9) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 0.8) +
  facet_wrap(~ covariate, scales = "free_x") +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Latent progression score versus confluence and cell count",
    x = NULL,
    y = "Latent progression score",
    color = "Lineage"
  )
```

Interpretation:

- a strong relationship here is not a problem by itself; it becomes a problem only if latent progression is nothing more than a density proxy
- if the same latent score occurs across a range of confluence or cell-count values, that supports a richer notion of morphological state

## Does Temporal Structure Remain After Density Adjustment?

To check whether the latent score reflects more than trivial density effects, fit a simple additive model and inspect the residual temporal structure.

```{r residualized-latent}
density_adjust_fit <- mgcv::gam(
  latent_score ~ s(log_confluence, k = 6) + s(log_cell_count, k = 6) + lineage_id,
  data = latent_tbl,
  method = "REML"
)

latent_tbl <- latent_tbl %>%
  mutate(latent_resid_density = resid(density_adjust_fit, type = "response"))

ggplot(latent_tbl, aes(x = time_since_seeding_days, y = latent_resid_density, color = lineage_id)) +
  geom_point(alpha = 0.7, size = 1.9) +
  geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6), linewidth = 0.9) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Latent progression residual after adjusting for confluence and cell count",
    subtitle = "Persistent temporal structure here argues against a purely density-driven embedding",
    x = "Time since seeding (days)",
    y = "Residual latent score",
    color = "Lineage"
  )
```

Interpretation:

- if temporal structure largely disappears here, the progression score is mostly a density axis
- if clear lineage-specific time structure remains, the embeddings are likely capturing morphology beyond simple occupancy proxies

## A Simple Alignment View Based on the Latent Score

Instead of forcing a formal time-warping model, use the latent score itself as the shared progression axis and estimate when each lineage reaches the same score levels.

```{r alignment-grid}
stage_score_grid <- seq(0.1, 0.9, by = 0.1)
alignment_tbl <- predict_time_to_stage(latent_tbl, lineage_levels, stage_score_grid)

ggplot(alignment_tbl, aes(x = stage_score, y = estimated_time_days, color = lineage_id)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Estimated time required to reach the same latent progression levels",
    subtitle = "Differences between curves suggest different traversal rates along a common axis",
    x = "Latent progression score",
    y = "Estimated time since seeding (days)",
    color = "Lineage"
  )
```

Interpretation:

- this is the simplest alignment diagnostic in the notebook
- if lineages have similar stage ordering but systematically different times-to-stage, that supports the idea of shared morphology traversed at different rates
- if the curves cross erratically or cannot be estimated for large parts of the score range, the notion of one shared progression is weaker

## Provisional Morphological Stage Bins

For a first-pass stratification, divide the latent score into broad bins. Quantile-based bins keep the stage definition robust to unequal lineage sizes and irregular sampling.

```{r stage-bins}
stage_breaks <- unique(quantile(latent_tbl$latent_score, probs = seq(0, 1, length.out = params$n_stage_bins + 1), na.rm = TRUE))
if (length(stage_breaks) < 3) {
  stop("Latent score does not have enough spread to define stage bins.", call. = FALSE)
}

stage_labels <- paste0("Stage ", seq_len(length(stage_breaks) - 1))

latent_tbl <- latent_tbl %>%
  mutate(
    stage_bin = cut(
      latent_score,
      breaks = stage_breaks,
      include.lowest = TRUE,
      labels = stage_labels
    )
  )
```

Plot the stage bins back onto PCA space first, so the discretization can be checked against the continuous geometry.

```{r stage-in-pca}
ggplot(latent_tbl, aes(x = PC1, y = PC2, color = stage_bin)) +
  geom_point(alpha = 0.85, size = 2.1) +
  scale_color_brewer(palette = "Spectral", direction = -1, na.value = "grey80") +
  labs(
    title = "Provisional stage bins mapped onto PCA state space",
    x = "PC1",
    y = "PC2",
    color = "Stage bin"
  )
```

Interpretation:

- good stage bins should tile a coherent progression through PCA space rather than forming noisy checkerboards
- if the bins jump around spatially, the latent score is not stable enough for stage-based comparison

Now check how the lineages populate those stage bins.

```{r stage-composition}
stage_comp_tbl <- latent_tbl %>%
  count(stage_bin, lineage_id, name = "n_images") %>%
  group_by(stage_bin) %>%
  mutate(stage_fraction = n_images / sum(n_images)) %>%
  ungroup()

ggplot(stage_comp_tbl, aes(x = stage_bin, y = stage_fraction, fill = lineage_id)) +
  geom_col(position = "stack", color = "white") +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Lineage composition within provisional morphological stage bins",
    x = "Stage bin",
    y = "Fraction of images in bin",
    fill = "Lineage"
  )
```

```{r stage-time-boxplots}
ggplot(latent_tbl, aes(x = stage_bin, y = time_since_seeding_days, fill = lineage_id)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.35, position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Time since seeding within each provisional stage bin",
    subtitle = "Comparable stages occurring at different times support the alignment idea",
    x = "Stage bin",
    y = "Time since seeding (days)",
    fill = "Lineage"
  )
```

```{r stage-density-boxplots}
stage_density_long <- latent_tbl %>%
  select(stage_bin, lineage_id, confluence, cell_count) %>%
  pivot_longer(cols = c(confluence, cell_count), names_to = "covariate", values_to = "value")

ggplot(stage_density_long, aes(x = stage_bin, y = value, fill = lineage_id)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.3, position = position_dodge(width = 0.8)) +
  facet_wrap(~ covariate, scales = "free_y") +
  scale_fill_manual(values = lineage_palette[seq_along(lineage_levels)]) +
  labs(
    title = "Confluence and cell count within provisional stage bins",
    x = "Stage bin",
    y = NULL,
    fill = "Lineage"
  )
```

Interpretation:

- the strongest support for alignment is when multiple lineages contribute to the same stage bins but reach those bins at different times
- if stage bins are almost lineage-specific, or if each stage is merely a monotone density bucket, the alignment is likely too simplistic

## Provisional Conclusions

This notebook intentionally stops short of a production trajectory model. At this stage, the main outputs to trust are:

1. the coverage plots
2. the PCA geometry and smooth trajectories
3. the latent-score versus time and density diagnostics
4. the stage-bin composition and timing plots

Reasonable next steps depend on what these diagnostics show:

- If the lineages overlap strongly in PCA space and differ mainly by time-to-stage, a more formal alignment model may be worthwhile.
- If progression remains after density adjustment but different lineages follow visibly different paths, the right next step is likely lineage-specific trajectory modeling rather than forced alignment.
- If most structure disappears after accounting for `confluence` and `cell_count`, then the embedding is probably acting more as a density summary than as a richer morphology descriptor.

For now, the stage bins in `latent_tbl$stage_bin` are a sensible provisional stratification of comparable image states across the three exemplar lineages.
