---
title: "Dev Unsupervised Morphology Workflow"
output:
  html_document:
    toc: true
    toc_float: true
params:
  output_dir: "data/dev_unsupervised_analysis"
  umap_neighbors: 20
  umap_min_dist: 0.15
  max_pcs: 20
  cluster_k: 8
  representative_per_cluster: 8
  crop_display_size: 64
  write_cluster_examples: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Overview

This notebook assumes the expensive preprocessing has already been run once with:

```r
Rscript("R/preprocess_dev_unsupervised.R")
```

That preprocessing step:

- excludes all `event == "seeding"` images
- builds the merged per-object table
- derives classical morphology features from masks
- computes and saves a 20-component PCA of the object embeddings

This notebook then uses those cached objects for exploratory visualization and clustering.

## Packages

```{r packages}
required_packages <- c(
  "dplyr",
  "ggplot2",
  "purrr",
  "readr",
  "stringr",
  "tibble",
  "uwot",
  "tiff",
  "scales"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    sprintf(
      "Install required packages before rendering: %s",
      paste(missing_packages, collapse = ", ")
    ),
    call. = FALSE
  )
}

library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(scales)
```

## Helpers

```{r helpers}
resolve_project_root <- function() {
  candidates <- c(
    normalizePath(".", winslash = "/", mustWork = FALSE),
    normalizePath("..", winslash = "/", mustWork = FALSE),
    normalizePath("../..", winslash = "/", mustWork = FALSE)
  )

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "README.md")) &&
        dir.exists(file.path(candidate, "manifests")) &&
        dir.exists(file.path(candidate, "dev_data"))) {
      return(candidate)
    }
  }

  stop("Could not locate project root from current working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

project_path <- function(path) {
  is_absolute <- grepl("^[A-Za-z]:[/\\\\]|^/", path)
  out <- path
  out[!is_absolute] <- file.path(project_root, path[!is_absolute])
  out
}

output_dir <- project_path(params$output_dir)

required_cache_files <- c(
  "seg_manifest_filtered.rds",
  "dev_object_table.rds",
  "embedding_pca_table.rds"
)
missing_cache_files <- required_cache_files[!file.exists(file.path(output_dir, required_cache_files))]
if (length(missing_cache_files) > 0) {
  stop(
    sprintf(
      paste(
        "Missing preprocessing outputs:",
        paste(missing_cache_files, collapse = ", "),
        "Run `Rscript R/preprocess_dev_unsupervised.R` from the project root first."
      )
    ),
    call. = FALSE
  )
}

read_mask_tiff <- function(path) {
  tiff::readTIFF(path, as.is = TRUE)
}

read_grayscale_image <- function(path) {
  img <- tiff::readTIFF(path, native = FALSE)
  dims <- dim(img)
  if (is.null(dims) || length(dims) == 2) {
    return(img)
  }
  if (length(dims) == 3) {
    channels <- seq_len(min(3, dims[[3]]))
    return(apply(img[, , channels, drop = FALSE], c(1, 2), mean))
  }
  stop(sprintf("Unsupported image dimensionality for %s", path), call. = FALSE)
}

rescale_crop <- function(x) {
  x <- as.matrix(x)
  x[!is.finite(x)] <- 0
  hi <- stats::quantile(x, probs = 0.995, na.rm = TRUE, names = FALSE)
  lo <- min(x, na.rm = TRUE)
  if (!is.finite(hi) || hi <= lo) {
    hi <- max(x, na.rm = TRUE)
  }
  if (!is.finite(hi) || hi <= lo) {
    return(matrix(0, nrow = nrow(x), ncol = ncol(x)))
  }
  out <- (x - lo) / (hi - lo)
  out[out < 0] <- 0
  out[out > 1] <- 1
  out
}

resize_nn <- function(mat, out_height = 64, out_width = 64) {
  y_idx <- round(seq(1, nrow(mat), length.out = out_height))
  x_idx <- round(seq(1, ncol(mat), length.out = out_width))
  mat[y_idx, x_idx, drop = FALSE]
}

write_crop_png <- function(mat, out_path) {
  display_mat <- resize_nn(rescale_crop(mat), params$crop_display_size, params$crop_display_size)
  dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
  grDevices::png(out_path, width = ncol(display_mat), height = nrow(display_mat))
  op <- graphics::par(mar = c(0, 0, 0, 0))
  on.exit({
    graphics::par(op)
    grDevices::dev.off()
  }, add = TRUE)
  graphics::image(
    x = seq_len(ncol(display_mat)),
    y = seq_len(nrow(display_mat)),
    z = t(display_mat[nrow(display_mat):1, , drop = FALSE]),
    col = grDevices::gray.colors(256, start = 0, end = 1),
    axes = FALSE,
    useRaster = TRUE
  )
}

write_cluster_examples <- function(object_tbl, cluster_col, view_label, n_per_cluster) {
  selected <- object_tbl %>%
    group_by(.data[[cluster_col]]) %>%
    arrange(cluster_distance, .by_group = TRUE) %>%
    slice_head(n = n_per_cluster) %>%
    ungroup()

  example_dir <- file.path(output_dir, "cluster_examples", tolower(view_label))
  dir.create(example_dir, recursive = TRUE, showWarnings = FALSE)

  selected %>%
    group_split(filename) %>%
    purrr::map(function(file_group) {
      image <- read_grayscale_image(file_group$raw_relpath[[1]])
      mask <- read_mask_tiff(file_group$mask_relpath[[1]])

      purrr::pmap(
        list(
          file_group$label,
          file_group$bbox_min_row,
          file_group$bbox_min_col,
          file_group$bbox_max_row,
          file_group$bbox_max_col,
          file_group[[cluster_col]],
          file_group$filename
        ),
        function(label, min_row, min_col, max_row, max_col, cluster_value, filename_value) {
          row_idx <- seq.int(min_row + 1L, max_row)
          col_idx <- seq.int(min_col + 1L, max_col)
          crop <- image[row_idx, col_idx, drop = FALSE]
          crop_mask <- mask[row_idx, col_idx, drop = FALSE] == label
          crop[!crop_mask] <- 0
          out_path <- file.path(
            example_dir,
            paste0(
              "cluster_",
              as.character(cluster_value),
              "__",
              tools::file_path_sans_ext(basename(filename_value)),
              "__label_",
              label,
              ".png"
            )
          )
          write_crop_png(crop, out_path)
          tibble(
            view = view_label,
            cluster = as.character(cluster_value),
            filename = filename_value,
            label = label,
            example_path = out_path
          )
        }
      ) %>%
        bind_rows()
    }) %>%
    bind_rows()
}

run_unsupervised_view <- function(object_tbl, feature_cols, view_name) {
  feature_mat <- object_tbl %>%
    select(all_of(feature_cols)) %>%
    as.matrix()

  keep_rows <- complete.cases(feature_mat)
  filtered_tbl <- object_tbl[keep_rows, , drop = FALSE]
  feature_mat <- feature_mat[keep_rows, , drop = FALSE]

  scaled_mat <- scale(feature_mat)
  pca <- prcomp(scaled_mat, center = FALSE, scale. = FALSE)
  n_pcs <- min(params$max_pcs, ncol(pca$x))
  pc_scores <- pca$x[, seq_len(n_pcs), drop = FALSE]
  n_neighbors <- min(params$umap_neighbors, max(2, nrow(pc_scores) - 1))
  n_clusters <- min(params$cluster_k, nrow(pc_scores))

  set.seed(1)
  umap_coords <- uwot::umap(
    pc_scores,
    n_neighbors = n_neighbors,
    min_dist = params$umap_min_dist,
    metric = "euclidean",
    verbose = FALSE,
    ret_model = FALSE
  )

  set.seed(1)
  km <- stats::kmeans(pc_scores, centers = n_clusters, nstart = 20)
  centers <- km$centers[km$cluster, , drop = FALSE]
  cluster_distance <- sqrt(rowSums((pc_scores - centers) ^ 2))

  filtered_tbl %>%
    mutate(
      view = view_name,
      cluster = factor(km$cluster),
      cluster_distance = cluster_distance,
      PC1 = pca$x[, 1],
      PC2 = if (ncol(pca$x) >= 2) pca$x[, 2] else 0,
      UMAP1 = umap_coords[, 1],
      UMAP2 = umap_coords[, 2]
    )
}
```

## Load Cached Inputs

```{r load-data}
seg_manifest <- readRDS(file.path(output_dir, "seg_manifest_filtered.rds"))
object_tbl <- readRDS(file.path(output_dir, "dev_object_table.rds"))
embedding_pca_tbl <- readRDS(file.path(output_dir, "embedding_pca_table.rds"))

object_tbl %>%
  summarise(
    images = n_distinct(filename),
    objects = n(),
    mean_objects_per_image = mean(table(filename))
  )
```

## Feature Views

```{r feature-views}
morphology_features <- c(
  "area_px",
  "log1p_area",
  "bbox_height",
  "bbox_width",
  "bbox_aspect",
  "bbox_fill_fraction",
  "centroid_row_scaled",
  "centroid_col_scaled",
  "intensity_mean",
  "intensity_sd",
  "intensity_q90"
)

embedding_features <- names(embedding_pca_tbl)[stringr::str_detect(names(embedding_pca_tbl), "^embedding_PC")]

morphology_view <- run_unsupervised_view(object_tbl, morphology_features, "Morphology")
embedding_view <- embedding_pca_tbl %>%
  left_join(
    object_tbl %>%
      select(
        filename, label, id, field, cellLine, raw_relpath, mask_relpath,
        bbox_min_row, bbox_min_col, bbox_max_row, bbox_max_col,
        edge_touching, log1p_area, bbox_fill_fraction, intensity_mean
      ),
    by = c("filename", "label", "id", "field", "cellLine")
  ) %>%
  run_unsupervised_view(embedding_features, "Embedding")

all_views <- bind_rows(morphology_view, embedding_view)
```

## UMAP: Morphology

```{r umap-morphology}
ggplot(filter(all_views, view == "Morphology"), aes(UMAP1, UMAP2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.2) +
  coord_equal() +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Morphology UMAP", color = "Cluster") +
  theme_minimal()
```

## UMAP: Embeddings

```{r umap-embedding}
ggplot(filter(all_views, view == "Embedding"), aes(UMAP1, UMAP2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.2) +
  coord_equal() +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Embedding PCA UMAP", color = "Cluster") +
  theme_minimal()
```

## Known Labels Overlaid On Embedding Space

```{r known-labels}
plot_embedding_label <- function(label_col, title) {
  ggplot(filter(all_views, view == "Embedding"), aes(UMAP1, UMAP2, color = .data[[label_col]])) +
    geom_point(alpha = 0.7, size = 1.1) +
    coord_equal() +
    labs(title = title, color = label_col) +
    theme_minimal()
}

plot_embedding_label("cellLine", "Embedding UMAP Colored By Cell Line")
plot_embedding_label("filename", "Embedding UMAP Colored By Image Filename")
```

## Cluster Summaries

```{r cluster-summary}
cluster_summary <- all_views %>%
  group_by(view, cluster) %>%
  summarise(
    objects = n(),
    images = n_distinct(filename),
    edge_touching_fraction = mean(edge_touching),
    mean_log1p_area = mean(log1p_area, na.rm = TRUE),
    mean_fill_fraction = mean(bbox_fill_fraction, na.rm = TRUE),
    mean_intensity = mean(intensity_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(view, desc(objects))

readr::write_csv(cluster_summary, file.path(output_dir, "cluster_summary.csv"))
knitr::kable(cluster_summary)
```

## Cluster Composition By Cell Line

```{r cluster-composition-cellline}
composition_by_cellline <- all_views %>%
  count(view, cellLine, cluster) %>%
  group_by(view, cellLine) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup()

ggplot(composition_by_cellline, aes(cellLine, frac, fill = cluster)) +
  geom_col(position = "stack") +
  facet_wrap(~ view, ncol = 1) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cluster Composition By Cell Line",
    x = "Cell Line",
    y = "Fraction Of Objects"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Cluster Composition By Image

```{r cluster-composition-image}
composition_by_image <- all_views %>%
  count(view, filename, cluster) %>%
  group_by(view, filename) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup()

ggplot(composition_by_image, aes(filename, frac, fill = cluster)) +
  geom_col(position = "stack") +
  facet_wrap(~ view, ncol = 1) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cluster Composition By Image",
    x = "Image Filename",
    y = "Fraction Of Objects"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Representative Crops

```{r representative-crops}
if (isTRUE(params$write_cluster_examples)) {
  embedding_examples <- write_cluster_examples(
    object_tbl = filter(all_views, view == "Embedding"),
    cluster_col = "cluster",
    view_label = "Embedding",
    n_per_cluster = params$representative_per_cluster
  )

  morphology_examples <- write_cluster_examples(
    object_tbl = filter(all_views, view == "Morphology"),
    cluster_col = "cluster",
    view_label = "Morphology",
    n_per_cluster = params$representative_per_cluster
  )

  example_index <- bind_rows(embedding_examples, morphology_examples) %>%
    arrange(view, cluster, filename, label)

  readr::write_csv(example_index, file.path(output_dir, "cluster_examples_index.csv"))
  knitr::kable(example_index)
}
```

## Candidate Artifact Clusters

```{r artifact-candidates}
artifact_candidates <- all_views %>%
  group_by(view, cluster) %>%
  summarise(
    objects = n(),
    edge_touching_fraction = mean(edge_touching),
    mean_fill_fraction = mean(bbox_fill_fraction, na.rm = TRUE),
    mean_log1p_area = mean(log1p_area, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(view, desc(edge_touching_fraction), mean_fill_fraction)

knitr::kable(artifact_candidates)
```

## Session Outputs

```{r outputs}
readr::write_csv(all_views, file.path(output_dir, "dev_unsupervised_object_table.csv"))

tibble(
  output = c(
    file.path(output_dir, "dev_object_table.csv"),
    file.path(output_dir, "embedding_pca_table.csv"),
    file.path(output_dir, "dev_unsupervised_object_table.csv"),
    file.path(output_dir, "cluster_summary.csv"),
    file.path(output_dir, "cluster_examples_index.csv")
  )
)
```
